---
title: "grant_prep"
author: "Gabrielle"
date: "2026-01-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(tidyverse)
library(janitor)
library(knitr)
library(kableExtra)
library(viridis)  # colorblind-friendly colors
```

## Pare back the data

Variables of interest, focusing on core demographic, socioeconomic, health, anthropometric, chronic disease, and cognitive measures

 Monica: For the dataset attached, the following variables are the ones that are important:
(1) demencia_aj_alt = dementia only in younger adults (who took the ACE)
(2) demencia_am = dementia in older adult (DC = deterioro cognitivo, DCL = deterioro cognitivo level)
(3) ace_cal = ACE total score

Eder [Translated with DeepL.com]: 
1. In young adults, we apply a cutoff point of 70 to distinguish CD from non-CD.

2. In older adults, we apply cutoff points of 19 and 23 to distinguish non-CD, MCI, and dementia. 

3. In both cases, we do not use the functional assessment variable. We do apply it in the “dementia” variable, which identifies “severe DC without functional impairment” when PFAQ was <=5 and “severe DC with functional impairment” when PFAQ was >=6 points. In the case of older adults, the DC category included only dementia.

PFAQ = Pfeffer Activities of Daily Living questionnaire

```{r foundation}
vars_keep <- c(
  "var008","var009","var019","var042","var043","var044","var049","var050",
  "var052","var053","var054","var055","var056","varn031","var069_n","var070_n",
  "varn041","var079_a","var062","var063","var064","varn065","varn073","varn151",
  "observacio_detalle","var999","varn012_n","varn013_n","varn014_n","varn015_n",
  "varn017_n","varn018_n","varn019_n","varn020_n","varn021_n","varn022_n",
  "varn100_n","etnia","lengua_mat","nse_cat_inv","nse_n_cat_inv","varn111_n",
  "var051_cat","region","var053_cat","var054_cat_cor","var059_cat","var061_cat",
  "hta","imc","estado_nutricional","sobobe","dm","velocidad_marcha",
  "var078_cat","var079_a_01","var079_a_02","var079_a_03","var079_a_04",
  "var079_a_05","var079_a_06","var079_a_07","var079_a_08",
  "rudas_cal","pfaq","demencia_aj", "demencia_aj_alt","dco_dem","demencia", "dementia_am", "ace_cal","depresion_dic"
)

base_monica_reduced <- base_monica %>%
  select(any_of(vars_keep))

```


## Rename cleanly 
```{r}
base_monica_renamed <- base_monica_reduced %>%
  rename(
    ## Geography / IDs
    district                = any_of("var008"),
    sector                  = any_of("var009"),
    national_id             = any_of("var019"),
    region                  = any_of("region"),

    ## Household fuel
    fuel_type               = any_of("var042"),
    fuel_use_frequency      = any_of("var043"),
    fuel_use_years          = any_of("var044"),

    ## Demographics
    age                     = any_of("var049"),
    years_residence         = any_of("var050"),
    region_of_origin        = any_of("var052"),
    maternal_language       = any_of("var053"),
    education_level         = any_of("var054"),
    years_education         = any_of("var055"),
    marital_status          = any_of("var056"),
    sex                     = any_of("var051_cat"),
    ethnicity               = any_of("etnia"),
    language                = any_of("lengua_mat"),

    ## Socioeconomic status
    employment_paid         = any_of("varn031"),
    ses_category            = any_of("nse_cat_inv"),
    ses_quintile            = any_of("nse_n_cat_inv"),
    insufficient_food_income= any_of("varn100_n"),

    ## Housing & services
    electricity_service     = any_of("varn012_n"),
    piped_water             = any_of("varn013_n"),
    sewer_service           = any_of("varn014_n"),
    internet_service        = any_of("varn015_n"),
    computer_access         = any_of("varn017_n"),
    washing_machine         = any_of("varn018_n"),
    oven                    = any_of("varn019_n"),
    refrigerator            = any_of("varn020_n"),
    vehicle_ownership       = any_of("varn021_n"),
    domestic_service        = any_of("varn022_n"),

    ## Health access
    no_healthcare_access    = any_of("varn111_n"),
    health_insurance        = any_of("var061_cat"),

    ## Anthropometrics
    weight_kg               = any_of("var069_n"),
    height_m                = any_of("var070_n"),
    bmi                     = any_of("imc"),
    nutritional_status      = any_of("estado_nutricional"),
    overweight_obese        = any_of("sobobe"),

    ## Clinical measures
    blood_glucose           = any_of("varn041"),
    hypertension            = any_of("hta"),
    diabetes                = any_of("dm"),
    gait_speed_ms           = any_of("velocidad_marcha"),

    ## Health behaviors
    smoking                 = any_of("var064"),
    alcohol_use             = any_of("varn065"),

    ## Chronic conditions
    chronic_disease         = any_of("var078_cat"),
    reported_conditions     = any_of("var079_a"),

    ## Specific conditions
    diabetes_mellitus       = any_of("var079_a_01"),
    arterial_hypertension   = any_of("var079_a_02"),
    obesity_hyperalimentation = any_of("var079_a_03"),
    gastritis_duodenitis    = any_of("var079_a_04"),
    low_back_pain           = any_of("var079_a_05"),
    osteoarthritis          = any_of("var079_a_06"),
    prostate_hyperplasia    = any_of("var079_a_07"),
    cataract                = any_of("var079_a_08"),

    ## Cognitive & mental health
    confusion_memory_loss   = any_of("varn151"),
    rudas_score             = any_of("rudas_cal"),
    pfaq_score              = any_of("pfaq"),
    dementia_aj             = any_of("demencia_aj"),
    dementia_young          = any_of("demencia_aj_alt"),
    severe_objective_ci     = any_of("dco_dem"),
    dementia                = any_of("demencia"),
    dementia_older          = any_of("demencia_am"),
    ace_total_score         = any_of("ace_cal"),
    depression              = any_of("depresion_dic"),
    cognitive_eval_incomplete = any_of("var999"),

    ## Location / notes
    location_description    = any_of("varn073"),
    observation_detail      = any_of("observacio_detalle")
  )

write.csv(base_monica_renamed, "cleaned_data.csv", row.names = FALSE)

```


## Read in 11 Purple Air Sensor Data, one by one

# API: 7E14E44F-ECCB-11F0-B596-4201AC1DC123

```{r}
pa1 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/LV_129323 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()


pa2 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/MP_PP_West_159005 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa3 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/V_West_159435 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa4 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/L_159465 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa5 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/C_WestSouth_159599 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa6 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/Carab_PP_East_180021 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa7 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/SMDP_PP_South_180043 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa8 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/JM_180073 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa9 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/PP_Central_180119 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa10 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/B_180251 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa11 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/Comas_PP_East_180321 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()



glimpse(pa1$time_stamp)

```



# Combine pa1–pa11 into one tidy object
```{r}
# explicitly add sensor id
pa1 <- pa1 %>% mutate(sensor_id = "pa1")
pa2 <- pa2 %>% mutate(sensor_id = "pa2")
pa3 <- pa3 %>% mutate(sensor_id = "pa3")
pa4 <- pa4 %>% mutate(sensor_id = "pa4")
pa5 <- pa5 %>% mutate(sensor_id = "pa5")
pa6 <- pa6 %>% mutate(sensor_id = "pa6")
pa7 <- pa7 %>% mutate(sensor_id = "pa7")
pa8 <- pa8 %>% mutate(sensor_id = "pa8")
pa9 <- pa9 %>% mutate(sensor_id = "pa9")
pa10 <- pa10 %>% mutate(sensor_id = "pa10")
pa11 <- pa11 %>% mutate(sensor_id = "pa11")


pa_all <- bind_rows(
  pa1, pa2, pa3, pa4, pa5, pa6, pa7, pa8, pa9, pa10, pa11
)

write.csv(pa_all, "/Users/gabriellebenoit/Documents/GitHub/peru_air/pa_all.csv", row.names = FALSE)


```


# Summary by sensor

ab_summary table:
- internal consistency
- sensor reliability
- whether any sensor is systematically noisy

```{r}
pa_all <- pa_all %>%
  mutate(
    ab_diff = pm2_5_cf_1_a - pm2_5_cf_1_b,
    ab_abs_diff = abs(ab_diff)
  )

ab_summary <- pa_all %>%
  group_by(sensor_id) %>%
  summarise(
    mean_abs_diff = mean(ab_abs_diff, na.rm = TRUE),
    median_abs_diff = median(ab_abs_diff, na.rm = TRUE),
    pct_large_diff = mean(ab_abs_diff > 5, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )

```

# Investigate 

Stratify A–B agreement by month (to detect firmware periods)
```{r}
# Create monthly agreement summary
ab_monthly <- pa_all %>%
  mutate(
    time_stamp = as.Date(time_stamp),
    year_month = floor_date(time_stamp, "month")
  ) %>%
  group_by(sensor_id, year_month) %>%
  summarise(
    n_obs = n(),
    mean_diff = mean(ab_diff, na.rm = TRUE),
    sd_diff = sd(ab_diff, na.rm = TRUE),
    mean_abs_diff = mean(ab_abs_diff, na.rm = TRUE),
    median_abs_diff = median(ab_abs_diff, na.rm = TRUE),
    pct_large_diff = mean(ab_abs_diff > 5, na.rm = TRUE) * 100,
    # Correlation between channels A and B
    cor_ab = cor(pm2_5_cf_1_a, pm2_5_cf_1_b, use = "complete.obs"),
    # IQR of absolute differences (robustness check)
    iqr_abs_diff = IQR(ab_abs_diff, na.rm = TRUE),
    .groups = "drop"
  )

# View the monthly summary
print(ab_monthly)

# Identify potential firmware change points
# Look for months with notable shifts in agreement metrics
ab_monthly <- ab_monthly %>%
  group_by(sensor_id) %>%
  arrange(year_month) %>%
  mutate(
    # Change from previous month
    delta_mean_abs = mean_abs_diff - lag(mean_abs_diff),
    delta_cor = cor_ab - lag(cor_ab),
    # Flag large changes (adjust thresholds as needed)
    flag_agreement_shift = abs(delta_mean_abs) > 1.0 | abs(delta_cor) > 0.1
  ) %>%
  ungroup()

# Summary: overall monthly pattern (across all sensors)
ab_monthly_overall <- ab_monthly %>%
  group_by(year_month) %>%
  summarise(
    n_sensors = n(),
    avg_mean_abs_diff = mean(mean_abs_diff, na.rm = TRUE),
    avg_cor_ab = mean(cor_ab, na.rm = TRUE),
    n_flagged = sum(flag_agreement_shift, na.rm = TRUE),
    .groups = "drop"
  )

print(ab_monthly_overall)

# Visualization: time series of agreement by month
ggplot(ab_monthly, aes(x = year_month, y = mean_abs_diff, color = sensor_id)) +
  geom_line() +
  geom_point() +
  labs(
    title = "A-B Channel Agreement Over Time",
    subtitle = "Mean absolute difference by sensor and month",
    x = "Month",
    y = "Mean |A - B|",
    color = "Sensor"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# Correlation over time
ggplot(ab_monthly, aes(x = year_month, y = cor_ab, color = sensor_id)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "red") +
  labs(
    title = "A-B Channel Correlation Over Time",
    subtitle = "Pearson correlation by sensor and month",
    x = "Month",
    y = "Correlation (A, B)",
    color = "Sensor"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# Heatmap: mean absolute difference by sensor × month
ggplot(ab_monthly, aes(x = year_month, y = sensor_id, fill = mean_abs_diff)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    title = "A-B Agreement Heatmap",
    x = "Month",
    y = "Sensor",
    fill = "Mean |A - B|"
  ) +
  theme_minimal()
```





## Plots

Embed descriptive plots

```{r demographics, echo=FALSE}
pa_all %>%
  mutate(ab_abs_diff = abs(pm2_5_cf_1_a - pm2_5_cf_1_b)) %>%
  ggplot(aes(x = ab_abs_diff)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(~ sensor_id, scales = "free_y") +
  labs(
    title = "Absolute Difference Between PM2.5 A and B Channels",
    x = "|A − B| (µg/m³)",
    y = "Count"
  )
```

# Quality control of PM2.5 measurements.
To assess internal consistency of PurpleAir PM2.5 measurements, we evaluated agreement between the dual A and B sensing channels. For each observation, we calculated the absolute difference between channels (|A − B|). Observations with |A − B| > 10 µg/m³ were flagged as exhibiting poor channel agreement, consistent with established PurpleAir quality assurance practices. These flagged observations were excluded from primary analyses, and sensitivity analyses were conducted to evaluate robustness to alternative inclusion criteria.

```{r}
pa_all <- pa_all %>%
  mutate(
    ab_abs_diff = abs(pm2_5_cf_1_a - pm2_5_cf_1_b),
    ab_qc_flag = ab_abs_diff > 10
  )


pa_all %>%
  group_by(sensor_id) %>%
  summarize(
    n_obs = n(),
    n_flagged = sum(ab_qc_flag, na.rm = TRUE),
    pct_flagged = mean(ab_qc_flag, na.rm = TRUE) * 100,
    max_ab_diff = max(ab_abs_diff, na.rm = TRUE)
  ) %>%
  arrange(desc(pct_flagged))


```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


# Quick Summary of pm2.5 across all sensors
```{r}
pm_summary <- pa_all %>%
  group_by(sensor_id) %>%
  summarise(
    min = min(pm2_5_cf_1, na.rm = TRUE),
    q25 = quantile(pm2_5_cf_1, 0.25, na.rm = TRUE),
    median = median(pm2_5_cf_1, na.rm = TRUE),
    mean = mean(pm2_5_cf_1, na.rm = TRUE),
    q75 = quantile(pm2_5_cf_1, 0.75, na.rm = TRUE),
    max = max(pm2_5_cf_1, na.rm = TRUE),
    n = n()
  )

print(pm_summary)

kable(pm_summary, digits = 1, 
      caption = "PM2.5 Summary Statistics by Sensor") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, color = "black", background = "white") %>%
  row_spec(1:nrow(pm_summary), color = "black", background = "white")

# Quick visual comparison
ggplot(pa_all, aes(x = sensor_id, y = pm2_5_cf_1)) +
  geom_boxplot() +
  labs(title = "PM2.5 Distribution by Sensor",
       x = "Sensor", y = "PM2.5 (μg/m³)") +
  theme_minimal()
```

## Simple visualization 
```{r}
# Simple PM2.5 Boxplots for Lima, Peru Purple Air Sensors
# Two approaches: (1) raw data, (2) outliers capped at 150 μg/m³

# Rename sensors with Purple Air codes and district suffix
pa_all <- pa_all %>%
  mutate(sensor_id = case_when(
    sensor_id == "pa1" ~ "129323_JM",
    sensor_id == "pa2" ~ "159005_PP",
    sensor_id == "pa3" ~ "159435_PP",
    sensor_id == "pa4" ~ "159465_JM",
    sensor_id == "pa5" ~ "159599_PP",
    sensor_id == "pa6" ~ "180021_PP",
    sensor_id == "pa7" ~ "180043_PP",
    sensor_id == "pa8" ~ "180073_JM",
    sensor_id == "pa9" ~ "180119_PP",
    sensor_id == "pa10" ~ "180251_JM",
    sensor_id == "pa11" ~ "180321_PP",
    TRUE ~ sensor_id
  ))


# PLOT 1: Raw data (as-is)
p1 <- ggplot(pa_all, aes(x = sensor_id, y = pm2_5_cf_1)) +
  geom_boxplot(fill = "#440154FF", alpha = 0.7) +
  labs(title = "PM2.5 Distribution by Sensor - Raw Data",
       subtitle = "Lima, Peru (Puente Piedra and Jesus Maria regions)",
       x = "Sensor ID",
       y = "PM2.5 (μg/m³)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

print(p1)


# PLOT 2: Outliers capped at 150 μg/m³
# Note: Sensors pa5, pa6, and pa8 have extreme values (>1000 μg/m³) capped for visualization
pa_capped <- pa_all %>%
  mutate(pm2_5_capped = ifelse(pm2_5_cf_1 > 150, 150, pm2_5_cf_1))

p2 <- ggplot(pa_capped, aes(x = sensor_id, y = pm2_5_capped)) +
  geom_boxplot(fill = "#35B779FF", alpha = 0.7) +
  labs(title = "PM2.5 Distribution by Sensor - Outliers Capped at 150 μg/m³",
       subtitle = "Lima, Peru (Puente Piedra and Jesus Maria regions)\nNote: Extreme values from 159599_PP, 180021_PP, 180073_JM capped for better visualization",
       x = "Sensor ID",
       y = "PM2.5 (μg/m³)",
       caption = "Sensors 159599_PP (max=2444.7), 180021_PP (max=1050.9), 180073_JM (max=1917.0) have values capped at 150") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

print(p2)


# Save both plots
ggsave("pm25_raw.png", p1, width = 10, height = 6, dpi = 300, bg = "white")
ggsave("pm25_capped.png", p2, width = 10, height = 6, dpi = 300, bg = "white")

cat("Done! Two plots created:\n")
cat("1. pm25_raw.png - Shows all data as-is\n")
cat("2. pm25_capped.png - Outliers >150 capped for better visualization\n")
```


## Adding location of sensor
Unable to get this via the data download tool, need to use the API directly
```{r}
library(tidyverse)
library(httr2)

# Your PurpleAir API read key
api_key <- "7E14E44F-ECCB-11F0-B596-4201AC1DC123"

# Your list of sensor IDs
sensor_ids <- c(129323,159005,159435,159465,159599,180021,180043,180073,180119,180251,180321)

# Function to fetch one sensor's metadata
get_sensor_coords <- function(sensor_id, api_key) {
  
  response <- request("https://api.purpleair.com/v1/sensors") |>
    req_url_path_append(sensor_id) |>
    req_headers(`X-API-Key` = api_key) |>
    req_url_query(fields = "name,latitude,longitude") |>
    req_perform()
  
  data <- response |> resp_body_json()
  
  tibble(
    sensor_id   = sensor_id,
    sensor_name = data$sensor$name,
    latitude    = data$sensor$latitude,
    longitude   = data$sensor$longitude
  )
}

# Loop through all sensors, pause briefly between calls to be polite to the API
sensor_locations <- map(sensor_ids, \(id) {
  Sys.sleep(0.5)
  get_sensor_coords(id, api_key)
}) |> list_rbind()

print(sensor_locations)

# Join to your PurpleAir data
unique(pa_all$sensor_id)
unique(sensor_locations$sensor_id)

pa_all <- pa_all %>%
  mutate(sensor_id = case_when(
    sensor_id == "129323" ~ "pa1",
    sensor_id == "159005" ~ "pa2",
    sensor_id == "159435" ~ "pa3",
    sensor_id == "159465" ~ "pa4",
    sensor_id == "159599" ~ "pa5",
    sensor_id == "180021" ~ "pa6",
    sensor_id == "180043" ~ "pa7",
    sensor_id == "180073" ~ "pa8",
    sensor_id == "180119" ~ "pa9",
    sensor_id == "180251" ~ "pa10",
    sensor_id == "180321" ~ "pa11",
    TRUE ~ sensor_id
  ))

purpleair_with_coords <- pa_all |>
  mutate(sensor_id = as.numeric(sensor_id)) |>
  left_join(sensor_locations, by = "sensor_id")


sensor_locations <- sensor_locations |>
  mutate(sensor_id_name = c("pa1","pa2","pa3","pa4","pa5","pa6",
                             "pa7","pa8","pa9","pa10","pa11"))

purpleair_with_coords <- pa_all |>
  left_join(sensor_locations, by = c("sensor_id" = "sensor_id_name"))


write_csv(purpleair_with_coords, "/Users/gabriellebenoit/Documents/GitHub/peru_air/purpleair_with_coords.csv")

```

