---
title: "march_mapping"
author: "Gabrielle"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(tidyverse)
library(janitor)
library(knitr)
library(kableExtra)
library(viridis)  # colorblind-friendly colors
library(httr2)
library(sf)
library(lme4)
library(broom.mixed)
library(osmdata)



```

New map, numbered by section. [Final version of the map](https://www.google.com/maps/d/edit?mid=1ZxIT6hY4dc9rtQl3oUCLNVQoDkfRK5w&usp=sharing) includes sector 5.

New updated working dataset: "Dataset2025_PP.xlsx"

## Primary variables of interest

Geography:

Col 8: Distrito
Col 9: Sector
Col 10: Conglomerado/pin
Col 11: pin

Cognitive scores (the clean calculated ones at the bottom):

Col 247: ACE total (0-100)
Col 264: RUDAS total (0-30)
Col 276: PFAQ total (0-33)
Col 296: PHQ-9 total (0-27)
Col 297: rendimiento cognitivo
Col 298: deterioro cognitivo
Col 369: ACE (possibly a recoded/cleaned version)
Col 370: RUDAS (same)

```{r}
# Load in the socio-cog data, accounting for the fact that the first row of data are actually column names
updated_base_data <- read_excel(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/Dataset2025_PP.xlsx",
  skip = 1,       # skip rows before the header row
  col_names = TRUE
)

# Select relevant variables and pare back the data

vars_keep_new <- c(
  "Distrito",
  colnames(updated_base_data)[9],   # Sector (name is very long)
  colnames(updated_base_data)[10],  # Conglomerado
  "pin",
  "rendimiento cognitivo",
  "deterioro cognitivo",
  "ACE",
  "RUDAS",
  colnames(updated_base_data)[247], # ACE total auto-score
  colnames(updated_base_data)[264], # RUDAS total auto-score
  colnames(updated_base_data)[276], # PFAQ total
  colnames(updated_base_data)[296]  # PHQ-9 total
)

updated_trimmed <- updated_base_data[, vars_keep_new]

# Clean the data
updated_trimmed_renamed <- updated_trimmed %>%
  rename(
    district              = any_of(colnames(updated_base_data)[8]),
    sector                = any_of(colnames(updated_base_data)[9]),
    conglomerado          = any_of(colnames(updated_base_data)[10]),
    pin                   = "pin",
    cognitive_performance = "rendimiento cognitivo",
    cognitive_impairment  = "deterioro cognitivo",
    ace_score             = "ACE",
    rudas_score           = "RUDAS",
    ace_total_score       = any_of(colnames(updated_base_data)[247]),
    rudas_total_score     = any_of(colnames(updated_base_data)[264]),
    pfaq_score            = any_of(colnames(updated_base_data)[276]),
    phq9_score            = any_of(colnames(updated_base_data)[296])
  )

# Save the data
saveRDS(updated_trimmed_renamed, 
        "/Users/gabriellebenoit/Documents/GitHub/peru_air/updated_trimmed_renamed.rds")

write.csv(updated_trimmed_renamed, 
          "/Users/gabriellebenoit/Documents/GitHub/peru_air/updated_trimmed_renamed.csv", 
          row.names = FALSE)

```


## Purple Air

Read in 11 Purple Air Sensor Data, one by one

# API: 7E14E44F-ECCB-11F0-B596-4201AC1DC123

```{r}
pa1 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/LV_129323 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()


pa2 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/MP_PP_West_159005 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa3 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/V_West_159435 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa4 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/L_159465 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa5 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/C_WestSouth_159599 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa6 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/Carab_PP_East_180021 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa7 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/SMDP_PP_South_180043 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa8 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/JM_180073 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa9 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/PP_Central_180119 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa10 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/B_180251 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa11 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/Comas_PP_East_180321 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()



glimpse(pa1$time_stamp)
```

## Combine pa1–pa11 into one tidy object

```{r}
# explicitly add sensor id
pa1 <- pa1 %>% mutate(sensor_id = "pa1")
pa2 <- pa2 %>% mutate(sensor_id = "pa2")
pa3 <- pa3 %>% mutate(sensor_id = "pa3")
pa4 <- pa4 %>% mutate(sensor_id = "pa4")
pa5 <- pa5 %>% mutate(sensor_id = "pa5")
pa6 <- pa6 %>% mutate(sensor_id = "pa6")
pa7 <- pa7 %>% mutate(sensor_id = "pa7")
pa8 <- pa8 %>% mutate(sensor_id = "pa8")
pa9 <- pa9 %>% mutate(sensor_id = "pa9")
pa10 <- pa10 %>% mutate(sensor_id = "pa10")
pa11 <- pa11 %>% mutate(sensor_id = "pa11")


pa_all <- bind_rows(
  pa1, pa2, pa3, pa4, pa5, pa6, pa7, pa8, pa9, pa10, pa11
)

write.csv(pa_all, "/Users/gabriellebenoit/Documents/GitHub/peru_air/pa_all.csv", row.names = FALSE)


```


## Summary by sensor

ab_summary table:
- internal consistency
- sensor reliability
- whether any sensor is systematically noisy

```{r}
pa_all <- pa_all %>%
  mutate(
    ab_diff = pm2_5_cf_1_a - pm2_5_cf_1_b,
    ab_abs_diff = abs(ab_diff)
  )

ab_summary <- pa_all %>%
  group_by(sensor_id) %>%
  summarise(
    mean_abs_diff = mean(ab_abs_diff, na.rm = TRUE),
    median_abs_diff = median(ab_abs_diff, na.rm = TRUE),
    pct_large_diff = mean(ab_abs_diff > 5, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )

```

## Investigate 

Stratify A–B agreement by month (to detect firmware periods)
```{r}
# Create monthly agreement summary
ab_monthly <- pa_all %>%
  mutate(
    time_stamp = as.Date(time_stamp),
    year_month = floor_date(time_stamp, "month")
  ) %>%
  group_by(sensor_id, year_month) %>%
  summarise(
    n_obs = n(),
    mean_diff = mean(ab_diff, na.rm = TRUE),
    sd_diff = sd(ab_diff, na.rm = TRUE),
    mean_abs_diff = mean(ab_abs_diff, na.rm = TRUE),
    median_abs_diff = median(ab_abs_diff, na.rm = TRUE),
    pct_large_diff = mean(ab_abs_diff > 5, na.rm = TRUE) * 100,
    # Correlation between channels A and B
    cor_ab = cor(pm2_5_cf_1_a, pm2_5_cf_1_b, use = "complete.obs"),
    # IQR of absolute differences (robustness check)
    iqr_abs_diff = IQR(ab_abs_diff, na.rm = TRUE),
    .groups = "drop"
  )

# View the monthly summary
print(ab_monthly)

# Identify potential firmware change points
# Look for months with notable shifts in agreement metrics
ab_monthly <- ab_monthly %>%
  group_by(sensor_id) %>%
  arrange(year_month) %>%
  mutate(
    # Change from previous month
    delta_mean_abs = mean_abs_diff - lag(mean_abs_diff),
    delta_cor = cor_ab - lag(cor_ab),
    # Flag large changes (adjust thresholds as needed)
    flag_agreement_shift = abs(delta_mean_abs) > 1.0 | abs(delta_cor) > 0.1
  ) %>%
  ungroup()

# Summary: overall monthly pattern (across all sensors)
ab_monthly_overall <- ab_monthly %>%
  group_by(year_month) %>%
  summarise(
    n_sensors = n(),
    avg_mean_abs_diff = mean(mean_abs_diff, na.rm = TRUE),
    avg_cor_ab = mean(cor_ab, na.rm = TRUE),
    n_flagged = sum(flag_agreement_shift, na.rm = TRUE),
    .groups = "drop"
  )

print(ab_monthly_overall)

# Visualization: time series of agreement by month
ggplot(ab_monthly, aes(x = year_month, y = mean_abs_diff, color = sensor_id)) +
  geom_line() +
  geom_point() +
  labs(
    title = "A-B Channel Agreement Over Time",
    subtitle = "Mean absolute difference by sensor and month",
    x = "Month",
    y = "Mean |A - B|",
    color = "Sensor"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# Correlation over time
ggplot(ab_monthly, aes(x = year_month, y = cor_ab, color = sensor_id)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "red") +
  labs(
    title = "A-B Channel Correlation Over Time",
    subtitle = "Pearson correlation by sensor and month",
    x = "Month",
    y = "Correlation (A, B)",
    color = "Sensor"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# Heatmap: mean absolute difference by sensor × month
ggplot(ab_monthly, aes(x = year_month, y = sensor_id, fill = mean_abs_diff)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    title = "A-B Agreement Heatmap",
    x = "Month",
    y = "Sensor",
    fill = "Mean |A - B|"
  ) +
  theme_minimal()
```


## Plots

Embed descriptive plots

```{r, echo=FALSE}
pa_all %>%
  mutate(ab_abs_diff = abs(pm2_5_cf_1_a - pm2_5_cf_1_b)) %>%
  ggplot(aes(x = ab_abs_diff)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(~ sensor_id, scales = "free_y") +
  labs(
    title = "Absolute Difference Between PM2.5 A and B Channels",
    x = "|A − B| (µg/m³)",
    y = "Count"
  )
```

## Quality control of PM2.5 measurements.

To assess internal consistency of PurpleAir PM2.5 measurements, we evaluated agreement between the dual A and B sensing channels. For each observation, we calculated the absolute difference between channels (|A − B|). Observations with |A − B| > 10 µg/m³ were flagged as exhibiting poor channel agreement, consistent with established PurpleAir quality assurance practices. These flagged observations were excluded from primary analyses, and sensitivity analyses were conducted to evaluate robustness to alternative inclusion criteria.

```{r}
pa_all <- pa_all %>%
  mutate(
    ab_abs_diff = abs(pm2_5_cf_1_a - pm2_5_cf_1_b),
    ab_qc_flag = ab_abs_diff > 10
  )


pa_all %>%
  group_by(sensor_id) %>%
  summarize(
    n_obs = n(),
    n_flagged = sum(ab_qc_flag, na.rm = TRUE),
    pct_flagged = mean(ab_qc_flag, na.rm = TRUE) * 100,
    max_ab_diff = max(ab_abs_diff, na.rm = TRUE)
  ) %>%
  arrange(desc(pct_flagged))


```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


## Quick Summary of pm2.5 across all sensors
```{r}
pm_summary <- pa_all %>%
  group_by(sensor_id) %>%
  summarise(
    min = min(pm2_5_cf_1, na.rm = TRUE),
    q25 = quantile(pm2_5_cf_1, 0.25, na.rm = TRUE),
    median = median(pm2_5_cf_1, na.rm = TRUE),
    mean = mean(pm2_5_cf_1, na.rm = TRUE),
    q75 = quantile(pm2_5_cf_1, 0.75, na.rm = TRUE),
    max = max(pm2_5_cf_1, na.rm = TRUE),
    n = n()
  )

print(pm_summary)

kable(pm_summary, digits = 1, 
      caption = "PM2.5 Summary Statistics by Sensor") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, color = "black", background = "white") %>%
  row_spec(1:nrow(pm_summary), color = "black", background = "white")

# Quick visual comparison
ggplot(pa_all, aes(x = sensor_id, y = pm2_5_cf_1)) +
  geom_boxplot() +
  labs(title = "PM2.5 Distribution by Sensor",
       x = "Sensor", y = "PM2.5 (μg/m³)") +
  theme_minimal()
```

## Simple visualization 

```{r}
# Simple PM2.5 Boxplots for Lima, Peru Purple Air Sensors
# Two approaches: (1) raw data, (2) outliers capped at 150 μg/m³

# Rename sensors with Purple Air codes and district suffix
pa_all <- pa_all %>%
  mutate(sensor_id = case_when(
    sensor_id == "pa1" ~ "129323_JM",
    sensor_id == "pa2" ~ "159005_PP",
    sensor_id == "pa3" ~ "159435_PP",
    sensor_id == "pa4" ~ "159465_JM",
    sensor_id == "pa5" ~ "159599_PP",
    sensor_id == "pa6" ~ "180021_PP",
    sensor_id == "pa7" ~ "180043_PP",
    sensor_id == "pa8" ~ "180073_JM",
    sensor_id == "pa9" ~ "180119_PP",
    sensor_id == "pa10" ~ "180251_JM",
    sensor_id == "pa11" ~ "180321_PP",
    TRUE ~ sensor_id
  ))


# PLOT 1: Raw data (as-is)
p1 <- ggplot(pa_all, aes(x = sensor_id, y = pm2_5_cf_1)) +
  geom_boxplot(fill = "#440154FF", alpha = 0.7) +
  labs(title = "PM2.5 Distribution by Sensor - Raw Data",
       subtitle = "Lima, Peru (Puente Piedra and Jesus Maria regions)",
       x = "Sensor ID",
       y = "PM2.5 (μg/m³)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

print(p1)


# PLOT 2: Outliers capped at 150 μg/m³
# Note: Sensors pa5, pa6, and pa8 have extreme values (>1000 μg/m³) capped for visualization
pa_capped <- pa_all %>%
  mutate(pm2_5_capped = ifelse(pm2_5_cf_1 > 150, 150, pm2_5_cf_1))

p2 <- ggplot(pa_capped, aes(x = sensor_id, y = pm2_5_capped)) +
  geom_boxplot(fill = "#35B779FF", alpha = 0.7) +
  labs(title = "PM2.5 Distribution by Sensor - Outliers Capped at 150 μg/m³",
       subtitle = "Lima, Peru (Puente Piedra and Jesus Maria regions)\nNote: Extreme values from 159599_PP, 180021_PP, 180073_JM capped for better visualization",
       x = "Sensor ID",
       y = "PM2.5 (μg/m³)",
       caption = "Sensors 159599_PP (max=2444.7), 180021_PP (max=1050.9), 180073_JM (max=1917.0) have values capped at 150") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

print(p2)


# Save both plots
ggsave("pm25_raw.png", p1, width = 10, height = 6, dpi = 300, bg = "white")
ggsave("pm25_capped.png", p2, width = 10, height = 6, dpi = 300, bg = "white")

cat("Done! Two plots created:\n")
cat("1. pm25_raw.png - Shows all data as-is\n")
cat("2. pm25_capped.png - Outliers >150 capped for better visualization\n")
```


## Adding location of sensor
Unable to get this via the data download tool, need to use the API directly
```{r}
#  PurpleAir API read key
api_key <- "7E14E44F-ECCB-11F0-B596-4201AC1DC123"

# Your list of sensor IDs
sensor_ids <- c(129323,159005,159435,159465,159599,180021,180043,180073,180119,180251,180321)

# Function to fetch one sensor's metadata
get_sensor_coords <- function(sensor_id, api_key) {
  response <- request("https://api.purpleair.com/v1/sensors") |>
    req_url_path_append(sensor_id) |>
    req_headers(`X-API-Key` = api_key) |>
    req_url_query(fields = "name,latitude,longitude") |>
    req_perform()
  
  data <- response |> resp_body_json()
  
  tibble(
    sensor_id   = sensor_id,
    sensor_name = data$sensor$name,
    latitude    = data$sensor$latitude,
    longitude   = data$sensor$longitude
  )
}

# Loop through all sensors
sensor_locations <- map(sensor_ids, \(id) {
  Sys.sleep(0.5)
  get_sensor_coords(id, api_key)
}) |> list_rbind()

# Strip suffix from pa_all sensor_id and join
purpleair_with_coords <- pa_all %>%
  mutate(sensor_id_clean = as.numeric(gsub("_.*", "", sensor_id))) %>%
  left_join(sensor_locations, by = c("sensor_id_clean" = "sensor_id"))

# Check
table(is.na(purpleair_with_coords$latitude))
print(purpleair_with_coords)

write_csv(purpleair_with_coords, "/Users/gabriellebenoit/Documents/GitHub/peru_air/purpleair_with_coords.csv")

# Save as a spatial file
purpleair_spatial <- purpleair_with_coords %>%
  distinct(sensor_id, sensor_name, latitude, longitude) %>%  # one row per sensor
  filter(!is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32718)  # reproject to match your other layers

st_write(purpleair_spatial,
         "/Users/gabriellebenoit/Documents/GitHub/peru_air/purpleair_sensors.gpkg",
         delete_dsn = TRUE)

```


## Using the google map data 

Only 9 unmatched out of 1,646 rows.. pretty negligible. Will proceed with the join. Those 9 participants simply won't have coordinates and will be dropped from the spatial analysis.

Many-to-many warning -- makes sense, as it is expected that multiple survey participants will live in the same conglomerate (same pin/location point), so one KML point matches many survey rows. That's expected for cluster-based sampling. 2,334 rows indicates one row per participant with their conglomerate's coordinates attached).

```{r}
# Load and combine all KML layers
kml_layers <- st_layers("/Users/gabriellebenoit/Documents/GitHub/peru_air/Neuroepidemiología_Puente_Piedra_Julio_2022.kml")
print(kml_layers)

layers <- kml_layers$name

kml_data <- do.call(rbind, lapply(layers, function(layer) {
  st_read("/Users/gabriellebenoit/Documents/GitHub/peru_air/Neuroepidemiología_Puente_Piedra_Julio_2022.kml",
          layer = layer, quiet = TRUE)
}))

# Inspect
nrow(kml_data)
colnames(kml_data)
head(kml_data)

updated_trimmed_renamed <- updated_trimmed_renamed %>%
  mutate(conglomerado_chr = as.character(conglomerado))

# Check overlap:
table(updated_trimmed_renamed$conglomerado_chr %in% kml_data$Name)

unmatched <- updated_trimmed_renamed %>%
  filter(!conglomerado_chr %in% kml_data$Name)
table(unmatched$sector)

# Only 9 unmatched out of 1,646 rows.. pretty negligible. Will proceed with the join. Those 9 participants simply won't have coordinates and will be dropped from the spatial analysis.

# Join survey data to KML geometry
spatial_data <- kml_data %>%
  rename(conglomerado_chr = Name) %>%
  left_join(updated_trimmed_renamed, by = "conglomerado_chr")

nrow(spatial_data)  # check

st_write(spatial_data,
         "/Users/gabriellebenoit/Documents/GitHub/peru_air/spatial_survey_data.gpkg",
         delete_dsn = TRUE)

# Save outputs
saveRDS(updated_trimmed_renamed,
        "/Users/gabriellebenoit/Documents/GitHub/peru_air/updated_trimmed_renamed.rds")

write.csv(updated_trimmed_renamed,
          "/Users/gabriellebenoit/Documents/GitHub/peru_air/updated_trimmed_renamed.csv",
          row.names = FALSE)

```


# Working with the data for QGIS prep and buffer analysis

None of the sensors have July 2022 data. The earliest data starts in 2023. This means the sensors were installed after the survey was conducted (July 2022). This is a substantial methodological issue. An option for now: Use the overall mean (pm25_mean) as a proxy for chronic exposure — this assumes PM2.5 levels are relatively stable over time at each location, which may be reasonable for this chronic exposure/cognitive health study.

```{r}

# Average PM2.5 per sensor across the full period
pm25_avg <- purpleair_with_coords %>%
  group_by(sensor_id_clean, sensor_name, latitude, longitude) %>%
  summarise(pm25_mean = mean(pm2_5_cf_1, na.rm = TRUE),
            pm25_july2022 = mean(pm2_5_cf_1[format(time_stamp, "%Y-%m") == "2022-07"], 
                                 na.rm = TRUE)) %>%
  ungroup()

print(pm25_avg)

# Make pm25_avg spatial
pm25_spatial <- pm25_avg %>%
  filter(!is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32718)

# Make sure spatial_data is in the same CRS
spatial_data_proj <- spatial_data %>%
  st_transform(32718)

# For each participant point, find the nearest sensor and join its PM2.5 value
nearest_sensor <- st_nearest_feature(spatial_data_proj, pm25_spatial)

spatial_data_proj$pm25_mean     <- pm25_spatial$pm25_mean[nearest_sensor]
spatial_data_proj$pm25_july2022 <- pm25_spatial$pm25_july2022[nearest_sensor]
spatial_data_proj$nearest_sensor_name <- pm25_spatial$sensor_name[nearest_sensor]

# Check
head(spatial_data_proj[, c("conglomerado_chr", "nearest_sensor_name", "pm25_mean", "pm25_july2022")])

# Investigating the issue:
  ## pm25_july2022 is NaN — meaning there's no July 2022 data for this sensor. check what months are actually available
  
purpleair_with_coords %>%
  mutate(year_month = format(time_stamp, "%Y-%m")) %>%
  count(sensor_name, year_month) %>%
  print(n = 50)

# Check if any sensors have earlier data -- see what the earliest date is per sensor
purpleair_with_coords %>%
  group_by(sensor_name) %>%
  summarise(earliest = min(time_stamp),
            latest = max(time_stamp),
            n_months = n()) %>%
  arrange(earliest)

# Save as geopackage for QGIS
  # analysis-ready dataset with participant locations, cognitive scores, and PM2.5 exposure all in one file
st_write(spatial_data_proj,
         "/Users/gabriellebenoit/Documents/GitHub/peru_air/spatial_data_final.gpkg",
         delete_dsn = TRUE)

# Save as RDS for continued R analysis
saveRDS(spatial_data_proj,
        "/Users/gabriellebenoit/Documents/GitHub/peru_air/spatial_data_final.rds")

```
## Begin modeling

General plan/aim

Core confounders (standard in cognitive epidemiology):

age
sex
years_education
education_level

Health-related:

bmi or nutritional_status
diabetes
hypertension
gait_speed_ms (marker of physical function/frailty)
depression (PHQ-9)

Socioeconomic:

ses_category
health_insurance

Exposure-related:

fuel_type (indoor air pollution — important confounder given your PM2.5 focus)
fuel_use_years

Geographic:

sector or district (to account for clustering)

```{r}


# Pull covariates from updated_base_data
covariates <- updated_base_data %>%
  select(
    conglomerado = colnames(updated_base_data)[10],
    fuel_type    = colnames(updated_base_data)[42],
    sex          = colnames(updated_base_data)[51],
    age          = colnames(updated_base_data)[49],
    language     = colnames(updated_base_data)[53],
    education    = colnames(updated_base_data)[54],
    employment   = colnames(updated_base_data)[57],
    obesity      = colnames(updated_base_data)[72],
    gait_speed   = colnames(updated_base_data)[76],
    smoking      = colnames(updated_base_data)[87],
    alcohol      = colnames(updated_base_data)[88]
  ) %>%
  mutate(conglomerado = as.character(conglomerado))

spatial_data_proj <- spatial_data_proj %>%
  left_join(covariates, by = c("conglomerado_chr" = "conglomerado")) %>%
  mutate(
    gait_speed = as.numeric(gait_speed),
    age        = as.numeric(age),
    sex        = as.factor(sex),
    education  = as.factor(education),
    employment = as.factor(employment),
    obesity    = as.factor(obesity),
    smoking    = as.factor(smoking),
    alcohol    = as.factor(alcohol),
    fuel_type  = as.factor(fuel_type)
  )

colnames(spatial_data_proj)

# Check
colnames(spatial_data_proj)
lapply(spatial_data_proj[, c("sex", "education", "employment",
                              "obesity", "smoking", "alcohol", "fuel_type")],
       levels)


# Run the model
model1 <- lm(ace_score ~ pm25_mean + 
               age + 
               sex + 
               education +
               employment +
               obesity +
               gait_speed +
               smoking +
               alcohol +
               fuel_type,
             data = spatial_data_proj)

summary(model1)

# Clustered version of the main model
model2 <- lmer(ace_score ~ pm25_mean +
                 age +
                 sex +
                 education +
                 employment +
                 obesity +
                 gait_speed +
                 smoking +
                 alcohol +
                 fuel_type +
                 (1 | conglomerado_chr),
               data = spatial_data_proj)

summary(model2)
```
## Results, model 1

Main exposure:

pm25_mean is not significant (p=0.22), with a very small negative coefficient (-0.016). There's no detectable association between PM2.5 exposure and ACE score in this model.

Significant covariates:

Age is the strongest predictor — each additional year is associated with about a 0.5 point decrease in ACE score (p<0.001)
Education matters a lot — all levels above the reference category show higher ACE scores, with Postgrado showing the largest effect (+47 points)
Obesity is surprisingly positive (+2.9, p=0.006) — is that a real finding or confounding?
Retirement (employment D) is associated with lower ACE scores (-5.3, p=0.017)

Not significant: sex, gait speed, smoking, alcohol, fuel type

Model fit: R² = 0.065; the model explains only about 6.5% of variance in ACE score — quite low, important predictors are likely missing, or there's a lot of individual variability

NOTES:
* 867 deleted observations due to missingness.
* The non-significant PM2.5 result could reflect the temporal mismatch we discussed (sensors installed after the survey)


## Results, model 2
Key differences from model 1--

The random effect for conglomerado has substantial variance (202.3) — confirming clustering is real and model2 is the more appropriate model than model 1
Age effect is attenuated (-0.26 vs -0.51) — some of the age effect in model1 was actually due to conglomerado-level clustering
Education effects are smaller but similar pattern — higher education still associated with better ACE scores
Obesity is no longer significant (p~0.27) — that finding in model1 was likely a clustering artifact
* In model 1, it captured obesity as significant because people within the same conglomerate tend to be similar in many ways — similar diet, lifestyle, socioeconomic conditions — and obesity clusters geographically. Once model 2 accounts for that conglomerate-level clustering with the random intercept, the obesity effect shrinks and loses significance. It wasn't really an individual-level effect; it was reflecting neighborhood-level patterns.

Consistent findings:

pm25_mean remains non-significant (-0.006, t=-0.38) — same conclusion as model 1
Age remains the strongest predictor
Sex, gait speed, smoking, alcohol, fuel type all non-significant

Takeaway: PM2.5 as measured here is not associated with ACE cognitive scores after accounting for age, education, and clustering. The temporal mismatch (sensors installed after the 2022 survey) is the most likely explanation and is an important limitation to acknowledge.


## Sensitivity analyses & Interaction Model

```{r}
# Sensitivity 1: Puente Piedra only
spatial_pp <- spatial_data_proj %>% filter(district == "Puente Piedra")

model2_pp <- lmer(ace_score ~ pm25_mean +
                    age + sex + education + employment +
                    obesity + gait_speed + smoking + alcohol + fuel_type +
                    (1 | conglomerado_chr),
                  data = spatial_pp)
summary(model2_pp)

# Sensitivity 2: RUDAS as outcome
model2_rudas <- lmer(rudas_score ~ pm25_mean +
                       age + sex + education + employment +
                       obesity + gait_speed + smoking + alcohol + fuel_type +
                       (1 | conglomerado_chr),
                     data = spatial_data_proj)
summary(model2_rudas)

# Interaction: PM2.5 x education
model2_interaction <- lmer(ace_score ~ pm25_mean * education +
                             age + sex + employment +
                             obesity + gait_speed + smoking + alcohol + fuel_type +
                             (1 | conglomerado_chr),
                           data = spatial_data_proj)
summary(model2_interaction)
```
Tested for effect modification by education and found none.

The interaction terms are all non-significant (all t-values near zero) -- education does not modify the relationship between PM2.5 and ACE score. 

The effect of PM2.5 on cognition doesn't differ by education level.


## Rudas Score
```{r}
# RUDAS as outcome
model2_rudas <- lmer(rudas_score ~ pm25_mean +
                       age + sex + education + employment +
                       obesity + gait_speed + smoking + alcohol + fuel_type +
                       (1 | conglomerado_chr),
                     data = spatial_data_proj)
summary(model2_rudas)
```
RUDAS shows a reversed age pattern compared to ACE.

For RUDAS, higher education is associated with lower scores, and older age with higher scores — which is the opposite of ACE. RUDAS was designed to be less sensitive to education and cultural background than other cognitive tests. It may be capturing something different than ACE in this population. 


## PFAQ Score

```{r}
# PFAQ as outcome
model2_pfaq <- lmer(pfaq_score ~ pm25_mean +
                      age + sex + education + employment +
                      obesity + gait_speed + smoking + alcohol + fuel_type +
                      (1 | conglomerado_chr),
                    data = spatial_data_proj)
summary(model2_pfaq)

hist(spatial_data_proj$pfaq_score, main = "PFAQ distribution")
summary(spatial_data_proj$pfaq_score)
```

"floor effect" happening. The vast majority of participants score 0-1 on PFAQ, meaning almost everyone in this sample has no functional impairment. Linear regression is not appropriate for this distribution.


## Create a table displaying these initial results
```{r}
# Extract and combine results from both models
ace_results <- tidy(model2, effects = "fixed") %>%
  mutate(outcome = "ACE")

rudas_results <- tidy(model2_rudas, effects = "fixed") %>%
  mutate(outcome = "RUDAS")

results_table <- bind_rows(ace_results, rudas_results) %>%
  select(outcome, term, estimate, std.error, statistic) %>%
  mutate(
    estimate  = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3)
  ) %>%
  rename(
    Outcome   = outcome,
    Term      = term,
    Estimate  = estimate,
    SE        = std.error,
    t_value   = statistic
  )

print(results_table, n = 50)

# Save as CSV
write.csv(results_table,
          "/Users/gabriellebenoit/Documents/GitHub/peru_air/model_results.csv",
          row.names = FALSE)
```
## Prettier table
```{r}
library(broom.mixed)

# Extract results with confidence intervals
ace_results <- tidy(model2, effects = "fixed", conf.int = TRUE) %>%
  mutate(outcome = "ACE")

rudas_results <- tidy(model2_rudas, effects = "fixed", conf.int = TRUE) %>%
  mutate(outcome = "RUDAS")

results_table <- bind_rows(ace_results, rudas_results) %>%
  select(outcome, term, estimate, conf.low, conf.high, std.error, statistic) %>%
  mutate(
    across(c(estimate, conf.low, conf.high, std.error, statistic), ~round(., 3)),
    ci = paste0("(", conf.low, ", ", conf.high, ")"),
    # Clean up term names
    term = recode(term,
      "(Intercept)"                                          = "Intercept",
      "pm25_mean"                                            = "PM2.5 (mean)",
      "age"                                                  = "Age",
      "sexMasculino"                                         = "Sex (Male)",
      "educationNinguno/ Sin educación"                      = "Education: None",
      "educationPostgrado"                                   = "Education: Postgrad",
      "educationPrimaria"                                    = "Education: Primary",
      "educationSecundaria"                                  = "Education: Secondary",
      "educationSuperior no universitaria"                   = "Education: Non-university higher",
      "educationSuperior universitaria"                      = "Education: University",
      "employmentB. Sí, sin pago"                            = "Employment: Unpaid",
      "employmentC. Sí, los dos, pagado y sin pago"          = "Employment: Both",
      "employmentD. No trabajo actualmente porque me jubilé" = "Employment: Retired",
      "employmentE. No trabajo actualmente por otro motivo"  = "Employment: Not working (other)",
      "obesitySí (IMC ≥ 30)"                                 = "Obesity (BMI ≥ 30)",
      "gait_speed"                                           = "Gait speed (s)",
      "smokingEx fumador"                                    = "Smoking: Ex-smoker",
      "smokingNunca ha fumado"                               = "Smoking: Never",
      "alcoholDiariamente"                                   = "Alcohol: Daily",
      "alcoholNunca o dejó de beber el año pasado"           = "Alcohol: Never/quit",
      "fuel_typeElectricidad"                                = "Fuel: Electricity",
      "fuel_typeGas"                                         = "Fuel: Gas",
      "fuel_typeLeña"                                        = "Fuel: Wood"
    )
  ) %>%
  select(Outcome = outcome,
         Term = term,
         Estimate = estimate,
         `95% CI` = ci,
         SE = std.error,
         `t-value` = statistic)

print(results_table, n = 50)

# Save
write.csv(results_table,
          "/Users/gabriellebenoit/Documents/GitHub/peru_air/model_results_formatted.csv",
          row.names = FALSE)
```


## Another idea: highways as proxy for poor air quality
```{r}
# Download major roads in the Puente Piedra area
highways <- opq(bbox = c(-77.15, -11.90, -77.05, -11.83)) %>%
  add_osm_feature(key = "highway", 
                  value = c("motorway", "trunk", "primary", "secondary")) %>%
  osmdata_sf()

highway_lines <- highways$osm_lines %>%
  st_transform(32718)  # reproject to match your other layers

# Save for QGIS
st_write(highway_lines,
         "/Users/gabriellebenoit/Documents/GitHub/peru_air/highways.gpkg",
         delete_dsn = TRUE)

# Distance from each participant point to nearest highway
spatial_data_proj$dist_highway_m <- as.numeric(
  st_distance(spatial_data_proj %>% st_transform(32718),
              highway_lines %>% st_union()) %>%
  apply(1, min)
)

summary(spatial_data_proj$dist_highway_m)

# Calculate distance from each participant to nearest highway
spatial_data_proj <- spatial_data_proj %>%
  st_transform(32718)

highway_lines <- highways$osm_lines %>%
  st_transform(32718)

# Distance to nearest highway for each participant
spatial_data_proj$dist_highway_m <- as.numeric(
  st_distance(spatial_data_proj, st_union(highway_lines))
)

summary(spatial_data_proj$dist_highway_m)
hist(spatial_data_proj$dist_highway_m, main = "Distance to nearest highway (m)")

# Save the highway data
st_write(highway_lines,
         "/Users/gabriellebenoit/Documents/GitHub/peru_air/highways.gpkg",
         delete_dsn = TRUE)
```

## Add highways data to both models

```{r}
spatial_data_proj <- spatial_data_proj %>%
  mutate(dist_highway_km = dist_highway_m / 1000)

# Re-run both models with km instead
model2_ace_highway <- lmer(ace_score ~ pm25_mean + dist_highway_km +
                             age + sex + education + employment +
                             obesity + gait_speed + smoking + alcohol + fuel_type +
                             (1 | conglomerado_chr),
                           data = spatial_data_proj)

model2_rudas_highway <- lmer(rudas_score ~ pm25_mean + dist_highway_km +
                               age + sex + education + employment +
                               obesity + gait_speed + smoking + alcohol + fuel_type +
                               (1 | conglomerado_chr),
                             data = spatial_data_proj)

summary(model2_ace_highway)
summary(model2_rudas_highway)
```

distance to hwy is non-significant in both models (t=0.19 for ACE, t=-0.67 for RUDAS). No association between highway proximity and cognitive scores.

Consistent null results — dist_highway_km is non-significant in both models (t=0.19 for ACE, t=-0.67 for RUDAS). No scaling warning this time either.

Overall summary of all  analyses:
Neither PM2.5 nor highway proximity is associated with cognitive scores (ACE or RUDAS) in this sample. The only consistent predictors are:

Age (stronger predictor across both outcomes)
Education (strongly predicts ACE but not RUDAS, as expected given RUDAS design)

The most likely explanations for the null air pollution findings:

Temporal mismatch — sensors installed after the 2022 survey
Limited spatial variability in PM2.5 — only 11 sensors across a relatively small area, so everyone gets a similar exposure estimate
The population may not have sufficient exposure contrast to detect an effect
