---
title: "march25update"
author: "Gabrielle"
date: "2026-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(tidyverse)
library(janitor)
library(knitr)
library(kableExtra)
library(viridis)  # colorblind-friendly colors
library(httr2)
library(sf)
library(lme4)
library(broom)
library(broom.mixed)
library(osmdata)
library(nnet)
library(forcats)

```

New map, numbered by section. [Final version of the map](https://www.google.com/maps/d/edit?mid=1ZxIT6hY4dc9rtQl3oUCLNVQoDkfRK5w&usp=sharing) includes sector 5.

Confirmed (as of March 25th, 2026) working dataset: "Base de datos (Monica)_09.16.2024_MN edits 09.19.2024 copy 3.xlsx"

## Primary variables of interest

Geography:

- Distrito
- Sector
- Conglomerado/pin

Cognitive scores:

- ACE
- RUADS
- PFAQ
- Dementia
- Dementia AM
- Dementia AJ
- Dementia AJ ALT
- deterioro cognitivo


```{r}
# Load in the socio-cog data, accounting for the fact that the first row of data are actually column names
updated_base_data <- read_excel(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/Base de datos (Monica)_09.16.2024_MN edits 09.19.2024 copy 3.xlsx")

updated_base_data <- updated_base_data %>%
  slice(-1)

# Select relevant variables and pare back the data

vars_keep_new <- c(
  "var008",   # District
  "var009",   # Sector
  "var010",  # Conglomerado
  "demencia", # Todos (para definir demencia segun RUDAS o ACE-III + PFAQ)
  "demencia_am", # Adulto mayor (no DC, DCL, demencia, segun solo RUDAS)
  "demencia_aj_alt", # Adulto joven (DC o no DC, segun solo ACE-III)
  "KL",   # Deterioro cognitivo
  "ace_cal",
  "rudas",
  "rudas_cal",
  "pfaq"
)

updated_trimmed <- updated_base_data[, vars_keep_new]

# Clean the data
updated_trimmed_renamed <- updated_trimmed %>%
  rename(
    district              = "var008",
    sector                = "var009",
    conglomerado          = "var010",
    any_relevant_dem      = "demencia",
    demencia_older        = "demencia_am",
    demencia_younger      = "demencia_aj_alt",
    cognitive_decline     = "KL",
    ace_score             = "ace_cal",
    rudas                 = "rudas",
    rudas_cal             = "rudas_cal",
    pfaq                  = "pfaq"
  )

# Save the data
saveRDS(updated_trimmed_renamed, 
        "/Users/gabriellebenoit/Documents/GitHub/peru_air/updated_trimmed_renamed.rds")

write.csv(updated_trimmed_renamed, 
          "/Users/gabriellebenoit/Documents/GitHub/peru_air/updated_trimmed_renamed.csv")

```


## Purple Air

Read in 11 Purple Air Sensor Data, one by one

# API: 7E14E44F-ECCB-11F0-B596-4201AC1DC123

```{r}
pa1 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/LV_129323 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()


pa2 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/MP_PP_West_159005 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa3 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/V_West_159435 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa4 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/L_159465 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa5 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/C_WestSouth_159599 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa6 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/Carab_PP_East_180021 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa7 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/SMDP_PP_South_180043 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa8 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/JM_180073 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa9 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/PP_Central_180119 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa10 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/B_180251 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()

pa11 <- read_csv(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 1-10-2026/download_puente_maria_2022_2026/Comas_PP_East_180321 2022-01-01 2026-01-10 43200-Minute Average.csv",
  show_col_types = FALSE
) %>%
  clean_names()



glimpse(pa1$time_stamp)
```

## Combine pa1–pa11 into one tidy object

```{r}
# explicitly add sensor id
pa1 <- pa1 %>% mutate(sensor_id = "pa1")
pa2 <- pa2 %>% mutate(sensor_id = "pa2")
pa3 <- pa3 %>% mutate(sensor_id = "pa3")
pa4 <- pa4 %>% mutate(sensor_id = "pa4")
pa5 <- pa5 %>% mutate(sensor_id = "pa5")
pa6 <- pa6 %>% mutate(sensor_id = "pa6")
pa7 <- pa7 %>% mutate(sensor_id = "pa7")
pa8 <- pa8 %>% mutate(sensor_id = "pa8")
pa9 <- pa9 %>% mutate(sensor_id = "pa9")
pa10 <- pa10 %>% mutate(sensor_id = "pa10")
pa11 <- pa11 %>% mutate(sensor_id = "pa11")


pa_all <- bind_rows(
  pa1, pa2, pa3, pa4, pa5, pa6, pa7, pa8, pa9, pa10, pa11
)

write.csv(pa_all, "/Users/gabriellebenoit/Documents/GitHub/peru_air/pa_all.csv", row.names = FALSE)


```


## Summary by sensor

ab_summary table:
- internal consistency
- sensor reliability
- whether any sensor is systematically noisy

```{r}
pa_all <- pa_all %>%
  mutate(
    ab_diff = pm2_5_cf_1_a - pm2_5_cf_1_b,
    ab_abs_diff = abs(ab_diff)
  )

ab_summary <- pa_all %>%
  group_by(sensor_id) %>%
  summarise(
    mean_abs_diff = mean(ab_abs_diff, na.rm = TRUE),
    median_abs_diff = median(ab_abs_diff, na.rm = TRUE),
    pct_large_diff = mean(ab_abs_diff > 5, na.rm = TRUE) * 100,
    n = n(),
    .groups = "drop"
  )

```

## Investigate 

Stratify A–B agreement by month (to detect firmware periods)
```{r}
# Create monthly agreement summary
ab_monthly <- pa_all %>%
  mutate(
    time_stamp = as.Date(time_stamp),
    year_month = floor_date(time_stamp, "month")
  ) %>%
  group_by(sensor_id, year_month) %>%
  summarise(
    n_obs = n(),
    mean_diff = mean(ab_diff, na.rm = TRUE),
    sd_diff = sd(ab_diff, na.rm = TRUE),
    mean_abs_diff = mean(ab_abs_diff, na.rm = TRUE),
    median_abs_diff = median(ab_abs_diff, na.rm = TRUE),
    pct_large_diff = mean(ab_abs_diff > 5, na.rm = TRUE) * 100,
    # Correlation between channels A and B
    cor_ab = cor(pm2_5_cf_1_a, pm2_5_cf_1_b, use = "complete.obs"),
    # IQR of absolute differences (robustness check)
    iqr_abs_diff = IQR(ab_abs_diff, na.rm = TRUE),
    .groups = "drop"
  )

# View the monthly summary
print(ab_monthly)

# Identify potential firmware change points
# Look for months with notable shifts in agreement metrics
ab_monthly <- ab_monthly %>%
  group_by(sensor_id) %>%
  arrange(year_month) %>%
  mutate(
    # Change from previous month
    delta_mean_abs = mean_abs_diff - lag(mean_abs_diff),
    delta_cor = cor_ab - lag(cor_ab),
    # Flag large changes (adjust thresholds as needed)
    flag_agreement_shift = abs(delta_mean_abs) > 1.0 | abs(delta_cor) > 0.1
  ) %>%
  ungroup()

# Summary: overall monthly pattern (across all sensors)
ab_monthly_overall <- ab_monthly %>%
  group_by(year_month) %>%
  summarise(
    n_sensors = n(),
    avg_mean_abs_diff = mean(mean_abs_diff, na.rm = TRUE),
    avg_cor_ab = mean(cor_ab, na.rm = TRUE),
    n_flagged = sum(flag_agreement_shift, na.rm = TRUE),
    .groups = "drop"
  )

print(ab_monthly_overall)

# Visualization: time series of agreement by month
ggplot(ab_monthly, aes(x = year_month, y = mean_abs_diff, color = sensor_id)) +
  geom_line() +
  geom_point() +
  labs(
    title = "A-B Channel Agreement Over Time",
    subtitle = "Mean absolute difference by sensor and month",
    x = "Month",
    y = "Mean |A - B|",
    color = "Sensor"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# Correlation over time
ggplot(ab_monthly, aes(x = year_month, y = cor_ab, color = sensor_id)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "red") +
  labs(
    title = "A-B Channel Correlation Over Time",
    subtitle = "Pearson correlation by sensor and month",
    x = "Month",
    y = "Correlation (A, B)",
    color = "Sensor"
  ) +
  theme_minimal() +
  theme(legend.position = "right")

# Heatmap: mean absolute difference by sensor × month
ggplot(ab_monthly, aes(x = year_month, y = sensor_id, fill = mean_abs_diff)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    title = "A-B Agreement Heatmap",
    x = "Month",
    y = "Sensor",
    fill = "Mean |A - B|"
  ) +
  theme_minimal()
```


## Plots

Embed descriptive plots

```{r, echo=FALSE}
pa_all %>%
  mutate(ab_abs_diff = abs(pm2_5_cf_1_a - pm2_5_cf_1_b)) %>%
  ggplot(aes(x = ab_abs_diff)) +
  geom_histogram(binwidth = 0.5) +
  facet_wrap(~ sensor_id, scales = "free_y") +
  labs(
    title = "Absolute Difference Between PM2.5 A and B Channels",
    x = "|A − B| (µg/m³)",
    y = "Count"
  )
```

## Quality control of PM2.5 measurements.

To assess internal consistency of PurpleAir PM2.5 measurements, we evaluated agreement between the dual A and B sensing channels. For each observation, we calculated the absolute difference between channels (|A − B|). Observations with |A − B| > 10 µg/m³ were flagged as exhibiting poor channel agreement, consistent with established PurpleAir quality assurance practices. These flagged observations were excluded from primary analyses, and sensitivity analyses were conducted to evaluate robustness to alternative inclusion criteria.

```{r}
pa_all <- pa_all %>%
  mutate(
    ab_abs_diff = abs(pm2_5_cf_1_a - pm2_5_cf_1_b),
    ab_qc_flag = ab_abs_diff > 10
  )


pa_all %>%
  group_by(sensor_id) %>%
  summarize(
    n_obs = n(),
    n_flagged = sum(ab_qc_flag, na.rm = TRUE),
    pct_flagged = mean(ab_qc_flag, na.rm = TRUE) * 100,
    max_ab_diff = max(ab_abs_diff, na.rm = TRUE)
  ) %>%
  arrange(desc(pct_flagged))


```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


## Quick Summary of pm2.5 across all sensors
```{r}
pm_summary <- pa_all %>%
  group_by(sensor_id) %>%
  summarise(
    min = min(pm2_5_cf_1, na.rm = TRUE),
    q25 = quantile(pm2_5_cf_1, 0.25, na.rm = TRUE),
    median = median(pm2_5_cf_1, na.rm = TRUE),
    mean = mean(pm2_5_cf_1, na.rm = TRUE),
    q75 = quantile(pm2_5_cf_1, 0.75, na.rm = TRUE),
    max = max(pm2_5_cf_1, na.rm = TRUE),
    n = n()
  )

print(pm_summary)

kable(pm_summary, digits = 1, 
      caption = "PM2.5 Summary Statistics by Sensor") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, color = "black", background = "white") %>%
  row_spec(1:nrow(pm_summary), color = "black", background = "white")

# Quick visual comparison
ggplot(pa_all, aes(x = sensor_id, y = pm2_5_cf_1)) +
  geom_boxplot() +
  labs(title = "PM2.5 Distribution by Sensor",
       x = "Sensor", y = "PM2.5 (μg/m³)") +
  theme_minimal()
```

## Simple visualization 

```{r}
# Simple PM2.5 Boxplots for Lima, Peru Purple Air Sensors
# Two approaches: (1) raw data, (2) outliers capped at 150 μg/m³

# Rename sensors with Purple Air codes and district suffix
pa_all <- pa_all %>%
  mutate(sensor_id = case_when(
    sensor_id == "pa1" ~ "129323_JM",
    sensor_id == "pa2" ~ "159005_PP",
    sensor_id == "pa3" ~ "159435_PP",
    sensor_id == "pa4" ~ "159465_JM",
    sensor_id == "pa5" ~ "159599_PP",
    sensor_id == "pa6" ~ "180021_PP",
    sensor_id == "pa7" ~ "180043_PP",
    sensor_id == "pa8" ~ "180073_JM",
    sensor_id == "pa9" ~ "180119_PP",
    sensor_id == "pa10" ~ "180251_JM",
    sensor_id == "pa11" ~ "180321_PP",
    TRUE ~ sensor_id
  ))


# PLOT 1: Raw data (as-is)
p1 <- ggplot(pa_all, aes(x = sensor_id, y = pm2_5_cf_1)) +
  geom_boxplot(fill = "#440154FF", alpha = 0.7) +
  labs(title = "PM2.5 Distribution by Sensor - Raw Data",
       subtitle = "Lima, Peru (Puente Piedra and Jesus Maria regions)",
       x = "Sensor ID",
       y = "PM2.5 (μg/m³)") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

print(p1)


# PLOT 2: Outliers capped at 150 μg/m³
# Note: Sensors pa5, pa6, and pa8 have extreme values (>1000 μg/m³) capped for visualization
pa_capped <- pa_all %>%
  mutate(pm2_5_capped = ifelse(pm2_5_cf_1 > 150, 150, pm2_5_cf_1))

p2 <- ggplot(pa_capped, aes(x = sensor_id, y = pm2_5_capped)) +
  geom_boxplot(fill = "#35B779FF", alpha = 0.7) +
  labs(title = "PM2.5 Distribution by Sensor - Outliers Capped at 150 μg/m³",
       subtitle = "Lima, Peru (Puente Piedra and Jesus Maria regions)\nNote: Extreme values from 159599_PP, 180021_PP, 180073_JM capped for better visualization",
       x = "Sensor ID",
       y = "PM2.5 (μg/m³)",
       caption = "Sensors 159599_PP (max=2444.7), 180021_PP (max=1050.9), 180073_JM (max=1917.0) have values capped at 150") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

print(p2)


# Save both plots
ggsave("pm25_raw.png", p1, width = 10, height = 6, dpi = 300, bg = "white")
ggsave("pm25_capped.png", p2, width = 10, height = 6, dpi = 300, bg = "white")

cat("Done! Two plots created:\n")
cat("1. pm25_raw.png - Shows all data as-is\n")
cat("2. pm25_capped.png - Outliers >150 capped for better visualization\n")
```


## Adding location of sensor
Unable to get this via the data download tool, need to use the API directly
```{r}
#  PurpleAir API read key
api_key <- "7E14E44F-ECCB-11F0-B596-4201AC1DC123"

# Your list of sensor IDs
sensor_ids <- c(129323,159005,159435,159465,159599,180021,180043,180073,180119,180251,180321)

# Function to fetch one sensor's metadata
get_sensor_coords <- function(sensor_id, api_key) {
  response <- request("https://api.purpleair.com/v1/sensors") |>
    req_url_path_append(sensor_id) |>
    req_headers(`X-API-Key` = api_key) |>
    req_url_query(fields = "name,latitude,longitude") |>
    req_perform()
  
  data <- response |> resp_body_json()
  
  tibble(
    sensor_id   = sensor_id,
    sensor_name = data$sensor$name,
    latitude    = data$sensor$latitude,
    longitude   = data$sensor$longitude
  )
}

# Loop through all sensors
sensor_locations <- map(sensor_ids, \(id) {
  Sys.sleep(0.5)
  get_sensor_coords(id, api_key)
}) |> list_rbind()

# Strip suffix from pa_all sensor_id and join
purpleair_with_coords <- pa_all %>%
  mutate(sensor_id_clean = as.numeric(gsub("_.*", "", sensor_id))) %>%
  left_join(sensor_locations, by = c("sensor_id_clean" = "sensor_id"))

# Check
table(is.na(purpleair_with_coords$latitude))
print(purpleair_with_coords)

write_csv(purpleair_with_coords, "/Users/gabriellebenoit/Documents/GitHub/peru_air/purpleair_with_coords.csv")

# Save as a spatial file
purpleair_spatial <- purpleair_with_coords %>%
  distinct(sensor_id, sensor_name, latitude, longitude) %>%  # one row per sensor
  filter(!is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32718)  # reproject to match your other layers

st_write(purpleair_spatial,
         "/Users/gabriellebenoit/Documents/GitHub/peru_air/purpleair_sensors.gpkg",
         delete_dsn = TRUE)

```


## Using the google map data 

Only 4 unmatched out of 1,099 rows.. negligible. Will proceed with the join. Those 4 participants simply won't have coordinates and will be dropped from the spatial analysis.

Many-to-many warning -- makes sense, as it is expected that multiple survey participants will live in the same conglomerate (same pin/location point), so one KML point matches many survey rows. That's expected for cluster-based sampling.

```{r}
# Load and combine all KML layers
kml_layers <- st_layers("/Users/gabriellebenoit/Documents/GitHub/peru_air/Neuroepidemiología_Puente_Piedra_Julio_2022.kml")
print(kml_layers)

layers <- kml_layers$name

kml_data <- do.call(rbind, lapply(layers, function(layer) {
  st_read("/Users/gabriellebenoit/Documents/GitHub/peru_air/Neuroepidemiología_Puente_Piedra_Julio_2022.kml",
          layer = layer, quiet = TRUE)
}))

# Inspect
nrow(kml_data)
colnames(kml_data)
head(kml_data)

updated_trimmed_renamed <- updated_trimmed_renamed %>%
  mutate(conglomerado_chr = as.character(conglomerado))

# Check overlap:
table(updated_trimmed_renamed$conglomerado_chr %in% kml_data$Name)

unmatched <- updated_trimmed_renamed %>%
  filter(!conglomerado_chr %in% kml_data$Name)
table(unmatched$sector)

# Only 4 unmatched out of 1,099 rows..  negligible. Will proceed with the join. Those 9 participants simply won't have coordinates and will be dropped from the spatial analysis.

# Join survey data to KML geometry
spatial_data <- kml_data %>%
  rename(conglomerado_chr = Name) %>%
  left_join(updated_trimmed_renamed, by = "conglomerado_chr")

nrow(spatial_data)  # check

st_write(spatial_data,
         "/Users/gabriellebenoit/Documents/GitHub/peru_air/spatial_survey_data.gpkg",
         delete_dsn = TRUE)

# Save outputs
saveRDS(updated_trimmed_renamed,
        "/Users/gabriellebenoit/Documents/GitHub/peru_air/updated_trimmed_renamed.rds")

write.csv(updated_trimmed_renamed,
          "/Users/gabriellebenoit/Documents/GitHub/peru_air/updated_trimmed_renamed.csv",
          row.names = FALSE)

```


# Working with the data for QGIS prep and buffer analysis

None of the sensors have July 2022 data. The earliest data starts in 2023. This means the sensors were installed after the survey was conducted (July 2022). This is a substantial methodological issue. An option for now: Use the overall mean (pm25_mean) as a proxy for chronic exposure — this assumes PM2.5 levels are relatively stable over time at each location, which may be reasonable for this chronic exposure/cognitive health study.

```{r}

# Average PM2.5 per sensor across the full period
pm25_avg <- purpleair_with_coords %>%
  group_by(sensor_id_clean, sensor_name, latitude, longitude) %>%
  summarise(pm25_mean = mean(pm2_5_cf_1, na.rm = TRUE),
            pm25_july2022 = mean(pm2_5_cf_1[format(time_stamp, "%Y-%m") == "2022-07"], 
                                 na.rm = TRUE)) %>%
  ungroup()

print(pm25_avg)

# Make pm25_avg spatial
pm25_spatial <- pm25_avg %>%
  filter(!is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32718)

# Make sure spatial_data is in the same CRS
spatial_data_proj <- spatial_data %>%
  st_transform(32718)

# For each participant point, find the nearest sensor and join its PM2.5 value
nearest_sensor <- st_nearest_feature(spatial_data_proj, pm25_spatial)

spatial_data_proj$pm25_mean     <- pm25_spatial$pm25_mean[nearest_sensor]
spatial_data_proj$pm25_july2022 <- pm25_spatial$pm25_july2022[nearest_sensor]
spatial_data_proj$nearest_sensor_name <- pm25_spatial$sensor_name[nearest_sensor]

# Check
head(spatial_data_proj[, c("conglomerado_chr", "nearest_sensor_name", "pm25_mean", "pm25_july2022")])

# Investigating the issue:
  ## pm25_july2022 is NaN — meaning there's no July 2022 data for this sensor. check what months are actually available
  
purpleair_with_coords %>%
  mutate(year_month = format(time_stamp, "%Y-%m")) %>%
  count(sensor_name, year_month) %>%
  print(n = 50)

# Check if any sensors have earlier data -- see what the earliest date is per sensor
purpleair_with_coords %>%
  group_by(sensor_name) %>%
  summarise(earliest = min(time_stamp),
            latest = max(time_stamp),
            n_months = n()) %>%
  arrange(earliest)

# Save as geopackage for QGIS
  # analysis-ready dataset with participant locations, cognitive scores, and PM2.5 exposure all in one file
st_write(spatial_data_proj,
         "/Users/gabriellebenoit/Documents/GitHub/peru_air/spatial_data_final.gpkg",
         delete_dsn = TRUE)

# Save as RDS for continued R analysis
saveRDS(spatial_data_proj,
        "/Users/gabriellebenoit/Documents/GitHub/peru_air/spatial_data_final.rds")

```
## Begin modeling

General plan/aim

Core confounders (standard in cognitive epidemiology):

age
sex
education
occupation

Health-related:

bmi
glucose_lev
hypertension

Socioeconomic:

overcrowding
health_insurance
civil_stat

Exposure-related:

pm25_mean

Geographic:

sector or district (to account for clustering)

```{r}


# Pull covariates from updated_base_data
covariates <- updated_base_data %>%
  select(
    conglomerado = "var010",
    sex          = "var051",
    age          = "var049",
    education    = "var055",
    civil_stat   = "var056",
    bmi          = "imc",
    glucose_lev  = "varn041",
    occupation   = "var059",
    health_ins   = "var061_cat...444",
    h_chron_dis  = "var078",
    htn          = "hta",
    nat_lang     = "lengua_mat",
    overcrowding = "hacinamiento"
  ) %>%
  mutate(conglomerado = as.character(conglomerado))

spatial_data_proj <- spatial_data_proj %>%
  left_join(covariates, by = c("conglomerado_chr" = "conglomerado")) %>%
  mutate(
    age        = as.numeric(age),
    sex        = as.factor(sex),
    education  = as.factor(education),
    civil_stat = as.factor(civil_stat),
    bmi    = as.factor(bmi),
    glucose_lev    = as.factor(glucose_lev),
    occupation    = as.factor(occupation),
    health_ins  = as.factor(health_ins),
    h_chron_dis  = as.factor(h_chron_dis),
    htn  = as.factor(htn),
    nat_lang  = as.factor(nat_lang),
    overcrowding  = as.factor(overcrowding)
  )

colnames(spatial_data_proj)

# Check
colnames(spatial_data_proj)
lapply(spatial_data_proj[, c("sex", "education", "occupation",
                              "bmi", "htn", "overcrowding", "civil_stat")],
       levels)

# Check the outcome variable:
str(spatial_data_proj$any_relevant_dem)
summary(spatial_data_proj$any_relevant_dem)
table(spatial_data_proj$any_relevant_dem, useNA = "ifany")

# Check missingness and types in predictors
str(spatial_data_proj %>% select(pm25_mean, age, sex, education,
                                 occupation, bmi, glucose_lev, h_chron_dis,
                                 htn, overcrowding))
sapply(spatial_data_proj %>% select(pm25_mean, age, sex, education,
                                    occupation, bmi, glucose_lev, h_chron_dis,
                                    htn, overcrowding), function(x) sum(is.na(x)))

# Approach 1: Clean, small dataset
spatial_data_simple <- spatial_data_proj %>%
  select(any_relevant_dem, pm25_mean, age, sex, education) %>%
  filter(!is.na(any_relevant_dem)) %>%       # Drop NA in outcome
  na.omit()                                  # Drop rows with NA in predictors

# Check size
dim(spatial_data_simple)
table(spatial_data_simple$any_relevant_dem)

# 1. Prepare data
spatial_data_simple <- spatial_data_proj %>%
  select(any_relevant_dem, pm25_mean, age, sex, education) %>%
  filter(!is.na(any_relevant_dem)) %>%    # Drop NA in outcome
  na.omit() %>%                           # Drop rows with NA in predictors
  mutate(
    any_relevant_dem = relevel(any_relevant_dem, ref = "Sin DC severo"),
    education = fct_lump(education, n = 5) # keep top 5 levels, lump rest as "Other"
  )

# 2. Fit multinomial logistic regression
model_simple <- multinom(any_relevant_dem ~ pm25_mean + age + sex + education,
                         data = spatial_data_simple)

# 3. Extract coefficients and standard errors
coefs <- summary(model_simple)$coefficients
ses   <- summary(model_simple)$standard.errors

# 4. Compute odds ratios and 95% confidence intervals
odds_ratios <- exp(coefs)
lower_ci <- exp(coefs - 1.96 * ses)
upper_ci <- exp(coefs + 1.96 * ses)

# 5. Tidy table: predictors as rows, outcomes as columns
odds_table <- as.data.frame(coefs) %>%
  rownames_to_column("Outcome_vs_Reference") %>%
  pivot_longer(-Outcome_vs_Reference, names_to = "Predictor", values_to = "Coef") %>%
  mutate(
    Odds_Ratio = exp(Coef),
    Lower_CI   = exp(Coef - 1.96 * ses[as.matrix(expand.grid(1:nrow(coefs), 1:ncol(coefs)))]),
    Upper_CI   = exp(Coef + 1.96 * ses[as.matrix(expand.grid(1:nrow(coefs), 1:ncol(coefs)))])
  ) %>%
  select(Outcome_vs_Reference, Predictor, Odds_Ratio, Lower_CI, Upper_CI)

# 6. View results
odds_table
  # each predictor is a row and the odds ratios (with 95% CI) are clearly displayed for each outcome vs "Sin DC severo"


# Prettier table
tidy_or <- as.data.frame(coefs) %>%
  rownames_to_column("Outcome_vs_Reference") %>%
  pivot_longer(-Outcome_vs_Reference, names_to = "Predictor", values_to = "Coef") %>%
  mutate(
    Odds_Ratio = exp(Coef),
    Lower_CI   = exp(Coef - 1.96 * ses[as.matrix(expand.grid(1:nrow(coefs), 1:ncol(coefs)))]),
    Upper_CI   = exp(Coef + 1.96 * ses[as.matrix(expand.grid(1:nrow(coefs), 1:ncol(coefs)))])
  ) %>%
  # Round numeric values
  mutate(
    Odds_Ratio = round(Odds_Ratio, 2),
    Lower_CI = round(Lower_CI, 2),
    Upper_CI = round(Upper_CI, 2)
  ) %>%
  # Combine OR and CI into one string
  unite("OR_with_CI", Odds_Ratio, Lower_CI, Upper_CI, sep = " (", remove = FALSE) %>%
  mutate(OR_with_CI = str_c(OR_with_CI, ")")) %>%
  select(-Odds_Ratio, -Lower_CI, -Upper_CI) %>%
  # Pivot wider so each outcome is a column
  pivot_wider(names_from = Outcome_vs_Reference, values_from = OR_with_CI)

# View publication-ready table
tidy_or


## Add clustered version, using conglomerado

```
## Results, model simple

Main exposure:


Significant covariates:


Not significant: 

Model fit: 

NOTES:
* ? deleted observations due to missingness.
* PM2.5 results could reflect the temporal mismatch (sensors installed after the survey)


## Results, model 2 (clustered at conglomerado)
Key differences from model 1--


Consistent findings:

Takeaway: 

## Sensitivity analyses & Interaction Model

```{r}
## REVISIT THIS (it is old data)


# Sensitivity 1: Puente Piedra only
spatial_pp <- spatial_data_proj %>% filter(district == "Puente Piedra")

model2_pp <- lmer(ace_score ~ pm25_mean +
                    age + sex + education + employment +
                    obesity + gait_speed + smoking + alcohol + fuel_type +
                    (1 | conglomerado_chr),
                  data = spatial_pp)
summary(model2_pp)

# Sensitivity 2: RUDAS as outcome
model2_rudas <- lmer(rudas_score ~ pm25_mean +
                       age + sex + education + employment +
                       obesity + gait_speed + smoking + alcohol + fuel_type +
                       (1 | conglomerado_chr),
                     data = spatial_data_proj)
summary(model2_rudas)

# Interaction: PM2.5 x education
model2_interaction <- lmer(ace_score ~ pm25_mean * education +
                             age + sex + employment +
                             obesity + gait_speed + smoking + alcohol + fuel_type +
                             (1 | conglomerado_chr),
                           data = spatial_data_proj)
summary(model2_interaction)
```
Does the effect of PM2.5 on cognition differ by education level?



## [REVISIT THIS -- OLD DATA] Another idea: highways as proxy for poor air quality
```{r}
# Download major roads in the Puente Piedra area
highways <- opq(bbox = c(-77.15, -11.90, -77.05, -11.83)) %>%
  add_osm_feature(key = "highway", 
                  value = c("motorway", "trunk", "primary", "secondary")) %>%
  osmdata_sf()

highway_lines <- highways$osm_lines %>%
  st_transform(32718)  # reproject to match your other layers

# Save for QGIS
st_write(highway_lines,
         "/Users/gabriellebenoit/Documents/GitHub/peru_air/highways.gpkg",
         delete_dsn = TRUE)

# Distance from each participant point to nearest highway
spatial_data_proj$dist_highway_m <- as.numeric(
  st_distance(spatial_data_proj %>% st_transform(32718),
              highway_lines %>% st_union()) %>%
  apply(1, min)
)

summary(spatial_data_proj$dist_highway_m)

# Calculate distance from each participant to nearest highway
spatial_data_proj <- spatial_data_proj %>%
  st_transform(32718)

highway_lines <- highways$osm_lines %>%
  st_transform(32718)

# Distance to nearest highway for each participant
spatial_data_proj$dist_highway_m <- as.numeric(
  st_distance(spatial_data_proj, st_union(highway_lines))
)

summary(spatial_data_proj$dist_highway_m)
hist(spatial_data_proj$dist_highway_m, main = "Distance to nearest highway (m)")

# Save the highway data
st_write(highway_lines,
         "/Users/gabriellebenoit/Documents/GitHub/peru_air/highways.gpkg",
         delete_dsn = TRUE)
```

## Add highways data to both models

```{r}
spatial_data_proj <- spatial_data_proj %>%
  mutate(dist_highway_km = dist_highway_m / 1000)

# Re-run both models with km instead
model2_ace_highway <- lmer(ace_score ~ pm25_mean + dist_highway_km +
                             age + sex + education + employment +
                             obesity + gait_speed + smoking + alcohol + fuel_type +
                             (1 | conglomerado_chr),
                           data = spatial_data_proj)

model2_rudas_highway <- lmer(rudas_score ~ pm25_mean + dist_highway_km +
                               age + sex + education + employment +
                               obesity + gait_speed + smoking + alcohol + fuel_type +
                               (1 | conglomerado_chr),
                             data = spatial_data_proj)

summary(model2_ace_highway)
summary(model2_rudas_highway)
```

distance to hwy is non-significant in both models (t=0.19 for ACE, t=-0.67 for RUDAS). No association between highway proximity and cognitive scores.

Consistent null results — dist_highway_km is non-significant in both models (t=0.19 for ACE, t=-0.67 for RUDAS). No scaling warning this time either.

Overall summary of all  analyses:
Neither PM2.5 nor highway proximity is associated with cognitive scores (ACE or RUDAS) in this sample. The only consistent predictors are:

Age (stronger predictor across both outcomes)
Education (strongly predicts ACE but not RUDAS, as expected given RUDAS design)

The most likely explanations for the null air pollution findings:

Temporal mismatch — sensors installed after the 2022 survey
Limited spatial variability in PM2.5 — only 11 sensors across a relatively small area, so everyone gets a similar exposure estimate
The population may not have sufficient exposure contrast to detect an effect
