---
title: "Chapter 3: Environmental and Social Determinants of Cognitive Health in Puente Piedra, Peru"
author: "Gabrielle Husted"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    latex_engine: xelatex
fontsize: 12pt
geometry: margin=1in
linestretch: 2
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,      # hides code from output
  warning = FALSE,      # hides warnings
  message = FALSE,      # hides messages
  fig.width  = 7,
  fig.height = 4.5,
  fig.align  = "center"
)


library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(lubridate)
library(geosphere)  # for distHaversine
library(naniar) # visualization of missingness
library(lme4)
library(broom.mixed)  # for tidy() on mixed models
library(sandwich)
library(lmtest)
library(logistf)
library(readxl)
library(stringr)
library(tidyverse)
library(janitor)
library(knitr)
library(kableExtra)
library(viridis)  # colorblind-friendly colors
library(httr2)
library(sf)
library(lme4)
library(broom)
library(osmdata)
library(nnet)
library(forcats)
library(gtsummary)
library(car)
library(writexl)
library(ggeffects)
library(marginaleffects)
library(ggh4x)





recode <- dplyr::recode  # fix masking conflict



```

## Download Historical Purple Air Data via Data Download Tool

Sensor history:
<https://api.purpleair.com/#api-sensors-get-sensor-history>

Fields:
pm10.0_atm_a,pm10.0_atm_b,pm10.0_atm,voc_a,voc_b,voc,humidity,humidity_a,humidity_b,
temperature,temperature_a,temperature_b,pm2.5_atm_a,pm2.5_atm_b,pm2.5_atm

Date range: 22-01-01 to 26-01-10

1-month average

```{r purpleair_clean_pipeline, message=FALSE, warning=FALSE}

# 1. READ + LABEL SENSORS \ raw data import
pa_159005 <- read_csv("/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 4-7-2026(1)/159005 2022-01-01 2026-01-10 43200-Minute Average.csv", show_col_types = FALSE) %>% clean_names() %>% mutate(sensor_id = "159005")
pa_159435 <- read_csv("/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 4-7-2026(1)/159435 2022-01-01 2026-01-10 43200-Minute Average.csv", show_col_types = FALSE) %>% clean_names() %>% mutate(sensor_id = "159435")
pa_159599 <- read_csv("/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 4-7-2026(1)/159599 2022-01-01 2026-01-10 43200-Minute Average.csv", show_col_types = FALSE) %>% clean_names() %>% mutate(sensor_id = "159599")
pa_180021 <- read_csv("/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 4-7-2026(1)/180021 2022-01-01 2026-01-10 43200-Minute Average.csv", show_col_types = FALSE) %>% clean_names() %>% mutate(sensor_id = "180021")
pa_180043 <- read_csv("/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 4-7-2026(1)/180043 2022-01-01 2026-01-10 43200-Minute Average.csv", show_col_types = FALSE) %>% clean_names() %>% mutate(sensor_id = "180043")
pa_180119 <- read_csv("/Users/gabriellebenoit/Documents/GitHub/peru_air/PurpleAir Download 4-7-2026(1)/180119 2022-01-01 2026-01-10 43200-Minute Average.csv", show_col_types = FALSE) %>% clean_names() %>% mutate(sensor_id = "180119")

purpleair_all <- bind_rows(
  pa_159005, pa_159435, pa_159599,
  pa_180021, pa_180043, pa_180119
)

# 2. CLEAN QC + FEATURE CREATION (MY OWN A/B + averages)

clean_purpleair_ab <- function(df) {
  
  df |>
    mutate(
      datetime = as.POSIXct(time_stamp, origin = "1970-01-01", tz = "UTC"),
      month    = floor_date(datetime, "month"),
      sensor_id = as.factor(sensor_id),
      
      # --- QC (A/B agreement) ---
      diff_pm25 = abs(pm2_5_atm_a - pm2_5_atm_b),
      diff_pm10 = abs(pm10_0_atm_a - pm10_0_atm_b),
      
      reldiff_pm25 = diff_pm25 / ((pm2_5_atm_a + pm2_5_atm_b) / 2),
      reldiff_pm10 = diff_pm10 / ((pm10_0_atm_a + pm10_0_atm_b) / 2),
      
      flag_pm25 = diff_pm25 > 5 | reldiff_pm25 > 0.70,
      flag_pm10 = diff_pm10 > 5 | reldiff_pm10 > 0.70,
      flag = flag_pm25 | flag_pm10,
      
      # --- FINAL EXPOSURE DEFINITION ---
      pm25_avg = (pm2_5_atm_a + pm2_5_atm_b) / 2,
      pm10_avg = (pm10_0_atm_a + pm10_0_atm_b) / 2,
      
      # meteorology (corrected)
      humidity_avg = humidity,
      temperature_avg = temperature
    ) |>
    filter(!flag | is.na(flag))
}

purpleair_all <- clean_purpleair_ab(purpleair_all)

# 3. A/B DIAGNOSTICS 
ab_summary <- purpleair_all |>
  mutate(
    diff_pm25 = abs(pm2_5_atm_a - pm2_5_atm_b),
    diff_pm10 = abs(pm10_0_atm_a - pm10_0_atm_b)
  )

# 4. FIGURE: A/B AGREEMENT HEATMAP
monthly_ab <- purpleair_all |>
  mutate(
    diff_pm25 = abs(pm2_5_atm_a - pm2_5_atm_b),
    diff_pm10 = abs(pm10_0_atm_a - pm10_0_atm_b),
    month = floor_date(datetime, "month")
  ) |>
  group_by(sensor_id, month) |>
  summarise(
    mean_diff_pm25 = mean(diff_pm25, na.rm = TRUE),
    mean_diff_pm10 = mean(diff_pm10, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

  # PM 2.5
ggplot(monthly_ab, aes(month, sensor_id, fill = mean_diff_pm25)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(title = "PM2.5 A–B Agreement")

  # PM 10
ggplot(monthly_ab, aes(month, sensor_id, fill = mean_diff_pm10)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(title = "PM10 A–B Agreement")

# 5. RAW TIME SERIES FIGURE
plot_data <- purpleair_all |>
  select(datetime, sensor_id,
         pm25_avg, pm10_avg,
         humidity_avg, temperature_avg) |>
  pivot_longer(-c(datetime, sensor_id),
               names_to = "variable",
               values_to = "value")

ggplot(plot_data, aes(datetime, value, color = sensor_id)) +
  geom_line(alpha = 0.4) +
  facet_wrap(~variable, scales = "free_y") +
  labs(title = "Processed PurpleAir Time Series")

# 6. SENSITIVITY ANSLYSIS
compute_exposure <- function(data, abs_thresh, rel_thresh) {
  
  data |>
    mutate(
      diff_pm25 = abs(pm2_5_atm_a - pm2_5_atm_b),
      diff_pm10 = abs(pm10_0_atm_a - pm10_0_atm_b),
      
      reldiff_pm25 = diff_pm25 / ((pm2_5_atm_a + pm2_5_atm_b) / 2),
      reldiff_pm10 = diff_pm10 / ((pm10_0_atm_a + pm10_0_atm_b) / 2),
      
      flag = (diff_pm25 > abs_thresh | reldiff_pm25 > rel_thresh) |
             (diff_pm10 > abs_thresh | reldiff_pm10 > rel_thresh)
    ) |>
    filter(!flag | is.na(flag)) |>
    group_by(sensor_id) |>
    summarise(
      mean_pm25 = mean(pm25_avg, na.rm = TRUE),
      mean_pm10 = mean(pm10_avg, na.rm = TRUE),
      mean_humidity = mean(humidity_avg, na.rm = TRUE),
      mean_temperature = mean(temperature_avg, na.rm = TRUE),
      .groups = "drop"
    ) |>
    
    mutate(
      abs_thresh = abs_thresh,
      rel_thresh = rel_thresh
    )
}

# 7. FINAL EXPOSURE OBJECTS (TO USE IN MODELS)
sensor_exposure_final <- compute_exposure(purpleair_all, 5, 0.70)

# 8. SENSITIVITY DELTA
abs_thresholds <- c(2, 5, 10, 20, Inf)
rel_thresholds <- c(0.25, 0.50, 0.70, 1.0)

threshold_grid <- expand.grid(
  abs_thresh = abs_thresholds,
  rel_thresh = rel_thresholds
)

sensitivity_results <- pmap_dfr(threshold_grid, function(abs_thresh, rel_thresh) {
  compute_exposure(purpleair_all, abs_thresh, rel_thresh)
})

baseline <- sensitivity_results |>
  filter(is.infinite(abs_thresh), rel_thresh == 1.0) |>
  select(sensor_id, base_pm25 = mean_pm25, base_pm10 = mean_pm10)

sensitivity_delta <- sensitivity_results |>
  left_join(baseline, by = "sensor_id") |>
  mutate(
    delta_pm25 = mean_pm25 - base_pm25,
    delta_pm10 = mean_pm10 - base_pm10
  )

# 9. DROP RAW CHANNELS

# save a raw copy
purpleair_raw <- purpleair_all  # save raw copy before cleaning


purpleair_all <- purpleair_all |>
  select(
    -pm2_5_atm_a, -pm2_5_atm_b,
    -pm10_0_atm_a, -pm10_0_atm_b,
    -voc, -voc_a, -voc_b
  )



## Cleaning up the agreement plots

# Shared sensor labels (reuse from before)
sensor_labels <- c(
  "180119" = "180119 – PuenteP Estadio",
  "159005" = "159005 – MiPeru",
  "180043" = "180043 – SanMa",
  "159599" = "159599 – Comas",
  "159435" = "159435 – Ventanilla",
  "180021" = "180021 – Carab"
)

# PM2.5
ggplot(monthly_ab, aes(month, sensor_id, fill = mean_diff_pm25)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Mean A–B\nDifference\n(µg/m³)") +
  scale_y_discrete(labels = sensor_labels) +
  labs(
    title = "PM2.5 A–B Channel Agreement by Sensor and Month",
    x     = "Month",
    y     = "Sensor"
  ) +
  theme_minimal()

# PM10
ggplot(monthly_ab, aes(month, sensor_id, fill = mean_diff_pm10)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Mean A–B\nDifference\n(µg/m³)") +
  scale_y_discrete(labels = sensor_labels) +
  labs(
    title = "PM10 A–B Channel Agreement by Sensor and Month",
    x     = "Month",
    y     = "Sensor"
  ) +
  theme_minimal()



# SENSITIVITY: restrict to 2023 – June 2024 (pre-degradation period)
purpleair_sensitivity_raw <- purpleair_raw |>
  mutate(datetime = as.POSIXct(time_stamp, origin = "1970-01-01", tz = "UTC")) |>
  filter(datetime < as.Date("2024-07-01")) |>
  clean_purpleair_ab()

sensor_exposure_sensitivity <- compute_exposure(purpleair_sensitivity_raw, 5, 0.70)




sensor_exposure_final
sensor_exposure_sensitivity

```

## Visualizations: Processed Exposure Time Series

```{r}
vars_of_interest <- c("humidity_avg", "temperature_avg", "pm25_avg", "pm10_avg")

plot_data <- purpleair_all |>
  select(datetime, sensor_id, all_of(vars_of_interest)) |>
  pivot_longer(all_of(vars_of_interest),
               names_to = "variable",
               values_to = "value")

ggplot(plot_data, aes(x = datetime, y = value, color = sensor_id)) +
  geom_line(alpha = 0.4, linewidth = 0.3) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "PurpleAir Time Series (Post-QC Exposure Variables)",
    x = NULL, y = NULL, color = "Sensor"
  ) +
  theme_minimal()


## make a cleaner plot of above
# Define sensor ID to name mapping
sensor_labels <- c(
  "180119" = "180119 – PuenteP Estadio",
  "159005" = "159005 – MiPeru",
  "180043" = "180043 – SanMa",
  "159599" = "159599 – Comas",
  "159435" = "159435 – Ventanilla",
  "180021" = "180021 – Carab"
)

# Define cleaner facet labels
var_labels <- c(
  humidity_avg    = "Humidity (%)",
  pm10_avg        = "PM10 (µg/m³)",
  pm25_avg        = "PM2.5 (µg/m³)",
  temperature_avg = "Temperature (°F)"
)

ggplot(plot_data, aes(x = datetime, y = value, color = sensor_id)) +
  geom_line(alpha = 0.4, linewidth = 0.3) +
  facet_wrap(~ variable, scales = "free_y", labeller = labeller(variable = var_labels)) +
  scale_color_discrete(labels = sensor_labels) +
  labs(
    title = "PurpleAir Time Series (Post-QC Exposure Variables)",
    x     = "Date",
    y     = "Measurement Value",
    color = "Sensor ID"
  ) +
  theme_minimal()

##



plot_data_capped <- plot_data |>
  group_by(variable) |>
  mutate(
    lo = quantile(value, 0.01, na.rm = TRUE),
    hi = quantile(value, 0.99, na.rm = TRUE),
    value = pmin(pmax(value, lo), hi)
  ) |>
  ungroup()

ggplot(plot_data_capped, aes(x = datetime, y = value, color = sensor_id)) +
  geom_line(alpha = 0.4, linewidth = 0.3) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title = "Capped Time Series (1st–99th percentile) — Exposure Variables",
    x = NULL, y = NULL, color = "Sensor"
  ) +
  theme_minimal()

sensor_exposure_final |>
  select(mean_pm25, mean_pm10,
         mean_humidity, mean_temperature) |>
  
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "value") |>
  
  group_by(variable) |>
  summarise(
    n      = sum(!is.na(value)),
    mean   = mean(value, na.rm = TRUE),
    sd     = sd(value, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    p25    = quantile(value, 0.25, na.rm = TRUE),
    p75    = quantile(value, 0.75, na.rm = TRUE),
    min    = min(value, na.rm = TRUE),
    max    = max(value, na.rm = TRUE),
    .groups = "drop"
  ) |>
    mutate(
  variable = case_when(
    variable == "mean_pm25"        ~ "PM2.5 (µg/m³)",
    variable == "mean_pm10"        ~ "PM10 (µg/m³)",
    variable == "mean_humidity"    ~ "Humidity (%)",
    variable == "mean_temperature" ~ "Temperature (°F)",
    TRUE                           ~ variable
  )
  )
```

## Adding location of sensor

Unable to get this via the data download tool, need to use the API
directly

```{r}
#  PurpleAir API read key
api_key <- "7E14E44F-ECCB-11F0-B596-4201AC1DC123"

# Your list of sensor IDs
sensor_ids <- c(159005,159435,159599,180021,180043,180119)

# Function to fetch one sensor's metadata
get_sensor_coords <- function(sensor_id, api_key) {
  response <- request("https://api.purpleair.com/v1/sensors") |>
    req_url_path_append(sensor_id) |>
    req_headers(`X-API-Key` = api_key) |>
    req_url_query(fields = "name,latitude,longitude") |>
    req_perform()
  
  data <- response |> resp_body_json()
  
  tibble(
    sensor_id   = sensor_id,
    sensor_name = data$sensor$name,
    latitude    = data$sensor$latitude,
    longitude   = data$sensor$longitude
  )
}

# Loop through all sensors
sensor_locations <- map(sensor_ids, \(id) {
  Sys.sleep(0.5)
  get_sensor_coords(id, api_key)
}) |> list_rbind()

# Strip suffix from purpleair_all_clean sensor_id and join
purpleair_all_with_coords <- sensor_exposure_final %>%
  mutate(sensor_id_clean = as.numeric(gsub("_.*", "", sensor_id))) %>%
  left_join(sensor_locations, by = c("sensor_id_clean" = "sensor_id"))

# Check
table(is.na(purpleair_all_with_coords$latitude))
print(purpleair_all_with_coords)

write_csv(purpleair_all_with_coords, "/Users/gabriellebenoit/Documents/GitHub/peru_air/purpleair_all_with_coords.csv")

# Save as a spatial file
purpleair_spatial <- purpleair_all_with_coords %>%
  distinct(sensor_id, sensor_name, latitude, longitude) %>%  # one row per sensor
  filter(!is.na(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(32718)  # reproject to match your other layers

st_write(purpleair_spatial,
         "/Users/gabriellebenoit/Documents/GitHub/peru_air/purpleair__all_sensors.gpkg",
         delete_dsn = TRUE)


```

## Socio-demographic Data

New map, numbered by section. [Final version of the
map](https://www.google.com/maps/d/edit?mid=1ZxIT6hY4dc9rtQl3oUCLNVQoDkfRK5w&usp=sharing)
includes sector 5.

Confirmed (as of March 25th, 2026) working dataset: "Base de datos
(Monica)\_09.16.2024_MN edits 09.19.2024 copy 3.xlsx"

## Primary variables of interest

Cognitive scores: - ACE - RUADS - PFAQ - Dementia - Dementia AM -
Dementia AJ - Dementia AJ ALT - deterioro cognitivo

Exposure-related: - pollution (pm2.5, pm10) - temperature - humidity

Geography: (account for clustering) - Distrito - Sector -
Conglomerado/pin

Core standard confounders: - age - sex - education - occupation

Health-related: - bmi - glucose_cat - hypertension

Socioeconomic: - overcrowding - health_insurance - civil_stat - SES
level (A - E)

```{r}
# Load in the socio-cog data, accounting for the fact that the first row of data are actually column names
updated_base_data <- read_excel(
  "/Users/gabriellebenoit/Documents/GitHub/peru_air/Base de datos (Monica)_09.16.2024_MN edits 09.19.2024 copy 3.xlsx")

updated_base_data <- updated_base_data %>%
  slice(-1)

# Select relevant variables and pare back the data

vars_keep_new <- c(
  "var008",   # District
  "var009",   # Sector
  "var010",  # Conglomerado
  "demencia", # Todos (para definir demencia segun RUDAS o ACE-III + PFAQ)
  "demencia_am", # Adulto mayor (no DC, DCL, demencia, segun solo RUDAS)
  "demencia_aj_alt", # Adulto joven (DC o no DC, segun solo ACE-III)
  "KL",   # Deterioro cognitivo
  "ace_cal",
  "rudas",
  "rudas_cal",
  "pfaq", 
  "var051", # sex
  "var049", # age
  "var055", # education
  "var056", # civil_stat
  "imc", # bmi
  "varn041", # glucose_lev
  "var059", # occupation
  "var061_cat...444", # health insurance
  "var078", # chronic disease
  "hta", # hypertension
  "lengua_mat", # native language
  "hacinamiento", # overcrowding
  "nse_cat", # nivel socioeconomico
  "nse_n_cat",
  "nse_n_cat_inv"
)

updated_trimmed <- updated_base_data[, vars_keep_new]

# Clean the data
updated_trimmed_renamed <- updated_trimmed %>%
  rename(
    district              = "var008",
    sector                = "var009",
    conglomerado          = "var010",
    any_relevant_dem      = "demencia",
    demencia_older        = "demencia_am",
    demencia_younger      = "demencia_aj_alt",
    cognitive_decline     = "KL",
    ace_score             = "ace_cal",
    rudas                 = "rudas",
    rudas_cal             = "rudas_cal",
    pfaq                  = "pfaq",
    sex                   = "var051",
    age                   = "var049",
    education             = "var055",
    civil_stat            = "var056",
    bmi                   = "imc",
    glucose_lev           = "varn041",
    occupation            = "var059",
    health_ins            = "var061_cat...444",
    chronic_dis           = "var078",
    htn                   = "hta",
    native_lang           = "lengua_mat",
    overcrowding          = "hacinamiento",
    ses_level             = "nse_cat", # nivel socioeconomico
    ses_quint             = "nse_n_cat",
    ses_quint_inv         = "nse_n_cat_inv"
  ) %>% 
  # ── Fix ses_level: remove rogue header row, set factor levels ──────────────
  mutate(
    ses_level = as.character(ses_level),
    ses_level = na_if(ses_level, "ses_level"),
    ses_level = factor(ses_level,
                       levels = c("NSE E", "NSE D", "NSE C", "NSE B", "NSE A"))
)

# Confirm before saving
table(updated_trimmed_renamed$ses_level, useNA = "always")
# Expect: NSE E=66, NSE D=95, NSE C=352, NSE B=497, NSE A=80, NA=10


# Save the data
saveRDS(updated_trimmed_renamed, 
        "/Users/gabriellebenoit/Documents/GitHub/peru_air/updated_trimmed_renamed.rds")

write.csv(updated_trimmed_renamed, 
          "/Users/gabriellebenoit/Documents/GitHub/peru_air/updated_trimmed_renamed.csv")
```

## Using the google map data

```{r}
# Load, deduplicate, and join — all one block

kml_layers <- st_layers("/Users/gabriellebenoit/Documents/GitHub/peru_air/Neuroepidemiología_Puente_Piedra_Julio_2022.kml")
layers <- kml_layers$name

kml_data <- do.call(rbind, lapply(layers, function(layer) {
  st_read("/Users/gabriellebenoit/Documents/GitHub/peru_air/Neuroepidemiología_Puente_Piedra_Julio_2022.kml",
          layer = layer, quiet = TRUE)
})) |>
  (\(x) { cat("Raw KML rows:", nrow(x), "| Dupes:", sum(duplicated(x$Name)), "\n"); x })() |>
  filter(!duplicated(Name)) |>        # dedup inline — cannot be skipped
  (\(x) { cat("Clean KML rows:", nrow(x), "\n"); x })()

# Prep survey IDs
updated_trimmed_renamed <- updated_trimmed_renamed |>
  mutate(conglomerado_chr = as.character(conglomerado))

# Join
spatial_data <- kml_data |>
  rename(conglomerado_chr = Name) |>
  right_join(updated_trimmed_renamed, by = "conglomerado_chr")
  #  ^^^ right_join: keeps all rows from updated_trimmed_renamed (1099)
  #      the 4 unmatched participants get NA geometry — expected

cat("spatial_data rows:", nrow(spatial_data), "\n")  # expect 1099



# Save files and outputs
st_write(spatial_data,
         "/Users/gabriellebenoit/Documents/GitHub/peru_air/spatial_survey_data.gpkg",
         delete_dsn = TRUE)

write.csv(spatial_data,
          "/Users/gabriellebenoit/Documents/GitHub/peru_air/spatial_survey_data.csv",
          row.names = FALSE)

saveRDS(updated_trimmed_renamed,
        "/Users/gabriellebenoit/Documents/GitHub/peru_air/updated_trimmed_renamed.rds")

write.csv(updated_trimmed_renamed,
          "/Users/gabriellebenoit/Documents/GitHub/peru_air/updated_trimmed_renamed.csv",
          row.names = FALSE)

```

## Spatial Join for All Variables

```{r}
# Make sensors spatial
# purpleair_all_with_coords already has sensor means + coordinates from code above
sensors_spatial <- purpleair_all_with_coords |>
  filter(!is.na(latitude), !is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(32718)

# Project participant data
spatial_data_proj <- spatial_data |>
  st_transform(32718)

# Nearest sensor join
nearest_idx <- st_nearest_feature(spatial_data_proj, sensors_spatial)

spatial_data_proj <- spatial_data_proj |>
  mutate(
    nearest_sensor_id   = purpleair_all_with_coords$sensor_id_clean[nearest_idx],
    nearest_sensor_name = purpleair_all_with_coords$sensor_name[nearest_idx],
    exposure_humidity   = purpleair_all_with_coords$mean_humidity[nearest_idx],
    exposure_temp       = purpleair_all_with_coords$mean_temperature[nearest_idx],
    exposure_pm25       = purpleair_all_with_coords$mean_pm25[nearest_idx],
    exposure_pm10       = purpleair_all_with_coords$mean_pm10[nearest_idx],
    n_obs_sensor        = purpleair_all_with_coords$n_obs[nearest_idx]
  )

# Check
head(spatial_data_proj[, c("conglomerado_chr", "nearest_sensor_name",
                            "exposure_pm25", "exposure_pm10",
                            "exposure_temp", "exposure_humidity")])

# Check how many observations are missing for any_relevant_dem and dc_binary
  # Then, drop observations that have NA
      # 1. Check number of NAs in each variable
spatial_data_proj |>
  summarise(
    n_na_any_relevant_dem = sum(is.na(any_relevant_dem)),
    n_na_sensor_id        = sum(is.na(nearest_sensor_id))
  )

      # 2. Drop rows with any NA in these variables
spatial_data_proj <- spatial_data_proj |>
  filter(
    !is.na(any_relevant_dem)
  )


# four observations are missing geometry, and consequently nearest_sensor_id values; see what the unmatched look like:
spatial_data_proj |>
  filter(is.na(nearest_sensor_id)) |>
  select(any_relevant_dem, nearest_sensor_id, conglomerado_chr, conglomerado, geometry)

# drop these four values
spatial_data_proj <- spatial_data_proj |>
  filter(!is.na(nearest_sensor_id))

# make all zeros for rudas NA instead
spatial_data_proj$rudas[spatial_data_proj$rudas == 0] <- NA


# Save
st_write(spatial_data_proj,
         "/Users/gabriellebenoit/Documents/GitHub/peru_air/spatial_data_final.gpkg",
         delete_dsn = TRUE)

saveRDS(spatial_data_proj,
        "/Users/gabriellebenoit/Documents/GitHub/peru_air/spatial_data_final.rds")

write.csv(st_drop_geometry(spatial_data_proj),
          "/Users/gabriellebenoit/Documents/GitHub/peru_air/spatial_data_final.csv",
          row.names = FALSE)
```

## Explore which sensors were used, how many participants were assigned to it

A few things to note: - 180119 and 159005 cover the bulk of participants
(371 and 517) — results will be heavily influenced by these two
sensors - 159435 and 180021 have very few participants assigned (4) —
these are likely on the geographic periphery of the study area

Limitations: exposure assignment was constrained to 6 sensors, with the
majority of participants assigned to just 2 sensors — which reduces
spatial variability in exposure estimates.

# Load Data

```{r}
spatial_data_proj <- read_rds("/Users/gabriellebenoit/Documents/GitHub/peru_air/spatial_data_final.rds")
```

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  count(nearest_sensor_id, nearest_sensor_name) |>
  arrange(desc(n))

# as a graphic
spatial_data_proj |>
  st_drop_geometry() |>
  count(nearest_sensor_id, nearest_sensor_name) |>
  ggplot(aes(x = reorder(nearest_sensor_name, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Participants Assigned per Nearest Sensor",
       x = NULL, y = "Number of participants") +
  theme_minimal() + scale_x_discrete(labels = c(
  "GEOHealth_MiPeru_Chavinillo" = "MiPeru",
  "GEOHealth_PuenteP_Estadio_1" = "PuenteP_Estadio",
  "GEOHealth_SanMa_IE005" = "SanMa",
  "GeoHealth_Comas_2055PAbril" = "Comas",
  "GEOHealth_Ventanilla_NSBelen" = "Ventanilla",
  "GEOHealth_Carab_VictorRaulHT" = "Carab"
))
```

## Examine contextual variables -- clean

```{r}
# Re-level outcome
spatial_data_proj <- spatial_data_proj %>%
  mutate(any_relevant_dem = factor(any_relevant_dem)) %>%
  mutate(any_relevant_dem = relevel(any_relevant_dem,
                                    ref = "Sin DC severo"))

# Binary outcome
spatial_data_proj <- spatial_data_proj %>%
  mutate(dc_binary = ifelse(any_relevant_dem == "Sin DC severo", 0, 1))


# inspect education
edu_counts <- spatial_data_proj |>
  st_drop_geometry() |>
  pull(education) |>
  table(useNA = "always")
print(as.data.frame(edu_counts))

# COLLAPSE
spatial_data_proj <- spatial_data_proj |>
  mutate(
    education_cat = case_when(
      education == 0                                             ~ "none",
      as.numeric(education) >= 1 & as.numeric(education) <= 6  ~ "primary",
      as.numeric(education) >= 7                                ~ "secondary_or_higher",
      TRUE                                                       ~ NA_character_
    ),
    education_cat = factor(education_cat,
                           levels = c("none", "primary", "secondary_or_higher"))
  )

table(spatial_data_proj$education_cat, spatial_data_proj$dc_binary)


# inspect civil_stat
spatial_data_proj |>
  st_drop_geometry() |>
  count(civil_stat) |>
  arrange(desc(n))
# inspect occupation
spatial_data_proj |>
  st_drop_geometry() |>
  count(occupation) |>
  arrange(desc(n))

# COLLAPSE
spatial_data_proj <- spatial_data_proj |>
  mutate(
    civil_stat_cat = case_when(
      civil_stat %in% c("Conviviente", "Casado(a)")    ~ "partnered",
      civil_stat == "Soltero(a)"                        ~ "single",
      civil_stat %in% c("Separado(a)", "Divorciado(a)",
                        "Viudo(a)")                     ~ "separated_widowed",
      TRUE                                              ~ NA_character_
    ),
    civil_stat_cat = factor(civil_stat_cat,
                            levels = c("partnered", "single", "separated_widowed")),

    occupation_cat = case_when(
      occupation == "Amo(a) de casa"                    ~ "homemaker",
      occupation %in% c("Auto-empleado(a)", "Freelance",
                        "Empleado(a)")                  ~ "employed",
      occupation %in% c("Sin empleo actual",
                        "Jubilado(a)", "Estudiante")    ~ "not_employed",
      TRUE                                              ~ NA_character_
    ),
    occupation_cat = factor(occupation_cat,
                            levels = c("employed", "homemaker", "not_employed"))
  )

# Verify
table(spatial_data_proj$civil_stat_cat, useNA = "always")
table(spatial_data_proj$occupation_cat, useNA = "always")

# trouble shoot a confounder that is perfectly predicting the outcome
model_data_check <- spatial_data_proj |>
  st_drop_geometry() |>
  filter(!is.na(dc_binary)) |>
  select(dc_binary, age, sex, education_cat, civil_stat_cat,
         bmi, htn, native_lang, health_ins, occupation_cat, chronic_dis) |>
  na.omit()

for (var in c("education_cat", "civil_stat_cat", "occupation_cat",
              "htn", "native_lang", "health_ins", "chronic_dis", "sex")) {
  cat("\n---", var, "---\n")
  print(table(model_data_check[[var]], model_data_check$dc_binary))
}
  # health_ins has the problem.  "Fuerzas armadas o policiales" has 46 participants and zero cognitive decline cases, causing complete separation. It also has too many levels generally. Collapse it:

spatial_data_proj <- spatial_data_proj |>
  mutate(
    health_ins_cat = case_when(
      health_ins == "Seguro integral de salud (SIS)"      ~ "SIS",
      health_ins == "Essalud"                             ~ "Essalud",
      health_ins == "No está afiliado"                    ~ "uninsured",
      health_ins %in% c("Fuerzas armadas o policiales",
                        "Entidad prestadora de salud",
                        "Seguro privado de salud",
                        "No sabe")                        ~ "other",
      TRUE                                                ~ NA_character_
    ),
    health_ins_cat = factor(health_ins_cat,
                            levels = c("SIS", "Essalud", "uninsured", "other"))
  )

table(spatial_data_proj$health_ins_cat, spatial_data_proj$dc_binary)

# htn
model_data_check2 <- spatial_data_proj |>
  st_drop_geometry() |>
  filter(!is.na(dc_binary)) |>
  select(dc_binary, age, sex, education_cat, civil_stat_cat,
         bmi, htn, native_lang, health_ins_cat, occupation_cat, chronic_dis) |>
  na.omit()

table(model_data_check2$htn, model_data_check2$dc_binary)

  # COLLAPSE
spatial_data_proj <- spatial_data_proj |>
  mutate(
    htn_cat = case_when(
      htn == "Sin hipertensión"                          ~ "none",
      htn %in% c("HTA sistólica", "HTA diastólica",
                 "HTA sistodiastólica")                  ~ "hypertension",
      TRUE                                               ~ NA_character_
    ),
    htn_cat = factor(htn_cat, levels = c("none", "hypertension"))
  )

table(spatial_data_proj$htn_cat, spatial_data_proj$dc_binary)


# BMI
spatial_data_proj <- spatial_data_proj |>
  mutate(bmi = as.numeric(as.character(bmi)))

# Verify
class(spatial_data_proj$bmi)
summary(spatial_data_proj$bmi)

# SES level
class(spatial_data_proj$ses_quint)
summary(spatial_data_proj$ses_quint)

table(spatial_data_proj$ses_level, useNA = "always")

counts <- spatial_data_proj |>
  count(any_relevant_dem)

ggplot(spatial_data_proj, aes(x = any_relevant_dem, fill = ses_level)) +
  geom_bar(position = "fill") +
  geom_text(
    data = counts,
    aes(x = any_relevant_dem, y = 1.05, label = paste0("n = ", n)),
    inherit.aes = FALSE
  ) +
  scale_x_discrete(labels = \(x) str_wrap(x, width = 10)) +
  labs(
    title = "Distribution of Dementia Status by SES Level",
    x = "Dementia Status",
    y = "Proportion",
    fill = "Socioeconomic Status"
  ) +
  coord_cartesian(ylim = c(0, 1.1)) +
  theme_minimal() +
  scale_x_discrete(
  labels = c(
    "Sin DC severo" = "Cognitively Unimpaired", 
    "DC severo con compromiso funcional" = "Cognitive impairment with functional impairment",
    "DC severo sin compromiso funcional" = "Cognitive impairment without functional impairment"
  )
)

# glucose
spatial_data_proj <- spatial_data_proj |>
  mutate(glucose_cat = case_when(
    as.numeric(glucose_lev) < 100  ~ "Normal",
    as.numeric(glucose_lev) < 126  ~ "Prediabetes",
    as.numeric(glucose_lev) >= 126 ~ "Diabetes",
    TRUE ~ NA_character_
  ),
  glucose_cat = factor(glucose_cat, levels = c("Normal", "Prediabetes", "Diabetes")))


```

## Check missingness

```{r}
gg_miss_var(spatial_data_proj)

# 1. How many events (dc_binary = 1)?
table(spatial_data_proj$dc_binary)
  ## 279 events

# 2. How many complete cases does fit_robust is actually working with?
spatial_data_proj |>
  st_drop_geometry() |>
  filter(!is.na(dc_binary)) |>
  select(dc_binary, exposure_pm25, age, sex, education_cat, civil_stat_cat,
         bmi, htn, native_lang, health_ins, occupation_cat, chronic_dis, glucose_cat) |>
  na.omit() |>
  nrow()

# exposure variable variance/values?
spatial_data_proj |>
  st_drop_geometry() |>
  count(exposure_pm25)
  ## essentially a 6-level factor masquerading as continuous

```

## Create standardized versions of PM 2.5, PM 10, temperature, and humidity

\# and quartile for PM 2.5

\# Interpretation: SD models: OR = dementia odds per 1 SD increase in
exposure Quartile models: OR = dementia odds relative to lowest exposure
group

```{r}
spatial_data_proj <- spatial_data_proj |>
  mutate(
    pm25_sd = as.numeric(scale(exposure_pm25)),
    pm10_sd = as.numeric(scale(exposure_pm10)),
    temp_sd = as.numeric(scale(exposure_temp)),
    humidity_sd = as.numeric(scale(exposure_humidity)),

    pm25_q = ntile(exposure_pm25, 4),
    pm25_q = factor(pm25_q,
                    levels = 1:4,
                    labels = c("Q1 lowest","Q2","Q3","Q4 highest"))
  )

spatial_data_proj <- spatial_data_proj |>
  mutate(
    age = as.numeric(as.character(age))
  )
class(spatial_data_proj$age)
```

## Descriptive Exploration

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  group_by(dc_binary) |>
  summarise(
    n = n(),
    pm25_mean = mean(exposure_pm25, na.rm = TRUE),
    pm25_sd   = sd(exposure_pm25, na.rm = TRUE),
    pm10_mean = mean(exposure_pm10, na.rm = TRUE),
    pm10_sd   = sd(exposure_pm10, na.rm = TRUE)
  )
```

\# boxplots \# pm 2.5

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  ggplot(aes(x = factor(dc_binary), y = exposure_pm25)) +
  geom_boxplot() +
  labs(
    x = "Dementia status",
    y = "PM2.5 exposure",
    title = "PM2.5 exposure by dementia status"
  ) +
  theme_minimal()
```

# pm 10

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  ggplot(aes(x = factor(dc_binary), y = exposure_pm10)) +
  geom_boxplot() +
  labs(
    x = "Dementia status",
    y = "PM10 exposure",
    title = "PM10 exposure by dementia status"
  ) +
  theme_minimal()
```

## Analysis (Thursday, April 9)

1.  dc_binary or any_relevant_dem as primary outcome
2.  stratified analyses for older and younger adults with their
    continuous scores as secondary outcome

\# determine how many NAs there are for any_relevant dementia,
dc_binary, rudas, ace, pfaq \# how do they or do they not overlap?

```{r}
# 1. Define the variables of interest
outcome_vars <- c("any_relevant_dem", "dc_binary", "rudas", "ace_score", "pfaq")

# 2. Count NAs per row for just these columns
na_per_row <- rowSums(is.na(spatial_data_proj[, outcome_vars]))

# 3. See the distribution (how many are missing 0, 1, 2, ..., 5)
table(na_per_row)
  # 0 = observations iwth complete data for all 5 variables
  # 1 = observations missing exactly one variable
  # 5 = observations missing all 5 variables

# alternative visual
colSums(is.na(spatial_data_proj[, outcome_vars]))

# Shows exactly which combinations of variables are most commonly missing together
gg_miss_upset(spatial_data_proj[, outcome_vars])


```

\## Delete observations that are missing all five variables because they
need at least one of those measures to be able to be in our analyses
meaningfully.

```{r}
# Define the variables to check
outcome_vars <- c("any_relevant_dem", "dc_binary", "rudas", "ace_score", "pfaq")

# Filter to KEEP rows where NOT all 5 are NA
spatial_data_proj <- spatial_data_proj %>%
  filter(rowSums(is.na(select(., any_of(outcome_vars)))) < 5)

# verification
nrow(spatial_data_proj)

```

```{r}
# Who got which test?
spatial_data_proj |>
  st_drop_geometry() |>
  summarise(
    n_rudas    = sum(!is.na(as.numeric(rudas))),
    n_ace      = sum(!is.na(as.numeric(ace_score))),
    n_pfaq     = sum(!is.na(as.numeric(pfaq))),
    n_dc       = sum(!is.na(dc_binary))
  )
# rudas, n = 1066
# ace, n = 753 
# pfaq, n = 896
# dc, n = 1066

# 753 took rudas and ACE, 313 did not
# 313 took rudas but not ACE

# Age distribution by test
spatial_data_proj |>
  st_drop_geometry() |>
  mutate(
    took_rudas = !is.na(as.numeric(rudas)),
    took_ace   = !is.na(as.numeric(ace_score))
  ) |>
  group_by(took_rudas, took_ace) |>
  summarise(
    mean_age = mean(as.numeric(age), na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# Confirm age cutoff used to assign tests
spatial_data_proj |>
  st_drop_geometry() |>
  mutate(
    took_ace  = !is.na(as.numeric(ace_score)),
    age_n     = as.numeric(age)
  ) |>
  group_by(took_ace) |>
  summarise(
    min_age  = min(age_n, na.rm = TRUE),
    max_age  = max(age_n, na.rm = TRUE),
    mean_age = mean(age_n, na.rm = TRUE),
    n        = n()
  )

```

## Primary analysis: full sample

```{r}
dc_binary ~ exposure + age + sex
```

## Secondary analysis: older adults (60+)

```{r}
as.numeric(rudas) ~ exposure + age + sex
```

## Secondary analysis: younger adults (30-59):

```{r}
as.numeric(ace_score) ~ exposure + age + sex
# n = 1,365, continuous ACE score
```

## Create the age groups:

```{r}
spatial_data_proj <- spatial_data_proj |>
  mutate(
    age_n     = as.numeric(age),
    age_group = case_when(
      age_n >= 60 ~ "older",
      age_n <  60 ~ "younger",
      TRUE        ~ NA_character_
    ),
    rudas_n   = as.numeric(rudas),
    ace_n     = as.numeric(ace_score)
  )
```

## Prep

```{r}
spatial_data_proj <- spatial_data_proj |>
  mutate(
    age_n     = as.numeric(age),
    age_group = case_when(
      age_n >= 60 ~ "older",
      age_n <  60 ~ "younger",
      TRUE        ~ NA_character_
    ),
    rudas_n   = as.numeric(rudas),
    ace_n     = as.numeric(ace_score),
    across(c(exposure_pm25, exposure_pm10,
             exposure_temp, exposure_humidity),
           ~ as.numeric(scale(.)),
           .names = "{.col}_z")
  )
```

## How many unique conglomerados per sector?

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  group_by(sector) |>
  summarise(
    n_conglomerados = n_distinct(conglomerado_chr),
    .groups = "drop"
  ) |>
  arrange(desc(n_conglomerados))
```

\# Confirm cluster counts

```{r}
length(unique(spatial_data_proj$conglomerado))  # expect 684
length(unique(spatial_data_proj$sector))         # expect 5
```

Our data is such that administrative codes are not equal to spatial
units. Many conglomerados values appear in multiple sectors. Some appear
in 2 sectors. Some appear in 3 or 4 (e.g. 128,409, 480)... conglomerado
is NOT globally unique.

We have this hierarchy: -District: Puente Piedra -Sector (5 levels)
-Conglomerado (reused within district, overlaps across sectors).. a
local code system, not a globally unique cluster ID

Requires TWO-WAY clustering or a composite ID derives from sector +
conglomerado

## Create Unique spatial cluster ID

```{r}
spatial_data_proj <- spatial_data_proj |>
  mutate(
    cluster_id = paste(sector, conglomerado_chr, sep = "_")
  )
```

## Standardize exposures

```{r}
spatial_data_proj <- spatial_data_proj |>
  mutate(across(
    c(exposure_pm25, exposure_pm10, exposure_temp, exposure_humidity),
    ~ as.numeric(scale(.)),
    .names = "{.col}_z"
  ))
```

## Tidy: unified cluster robust model function

```{r}
run_cluster_model <- function(data, outcome, exposure, covariates,
                              family = NULL,
                              exponentiate = FALSE,
                              outcome_label = "outcome",
                              keep_all_terms = FALSE) {

  # ----------------------------
  # 1. build model data
  # ----------------------------
  model_data <- data |>
    st_drop_geometry() |>
    select(all_of(c(outcome, exposure, covariates, "cluster_id"))) |>
    na.omit()

  # ----------------------------
  # 2. formula
  # ----------------------------
  fml <- as.formula(
    paste(outcome, "~", paste(c(exposure, covariates), collapse = " + "))
  )

  # ----------------------------
  # 3. fit model
  # ----------------------------
  if (is.null(family)) {
    model <- lm(fml, data = model_data)
  } else {
    model <- glm(fml, data = model_data, family = family)
  }

  # ----------------------------
  # 4. cluster-robust SE
  # ----------------------------
  vcov_cl <- sandwich::vcovCL(model, cluster = model_data$cluster_id)

  ct <- lmtest::coeftest(model, vcov = vcov_cl)

  # ----------------------------
  # 5. SAFE tidy conversion
  # ----------------------------
  ct_df <- as.data.frame(unclass(ct))
  ct_df$term <- rownames(ct_df)
  rownames(ct_df) <- NULL

  colnames(ct_df)[1:4] <- c("estimate", "std.error", "statistic", "p.value")

  # ----------------------------
  # 6. keep exposure row only
  # ----------------------------
if (!keep_all_terms) {
    ct_df <- ct_df |>
      filter(term == exposure)
  } else {
    ct_df <- ct_df |>
      filter(term != "(Intercept)")  # keep everything except intercept
  }
  
  ct_df <- ct_df |>
    mutate(
      conf.low  = estimate - 1.96 * std.error,
      conf.high = estimate + 1.96 * std.error
    )

  # ----------------------------
  # 7. transform if needed
  # ----------------------------
  if (!is.null(family) && exponentiate) {
    ct_df <- ct_df |>
      mutate(
        estimate  = exp(estimate),
        conf.low  = exp(conf.low),
        conf.high = exp(conf.high)
      )
  }

  # ----------------------------
  # 8. add labels
  # ----------------------------
  ct_df |>
    mutate(
      exposure = exposure,
      outcome  = outcome_label
    )
}
```


```{r}
covariate_sets <- list(
  "Unadjusted"       = NULL,
  "Age & sex adjusted" = c("age_n", "sex"),
  "Fully adjusted"   = c("age_n", "sex", "ses_level", "bmi", 
                          "education_cat", "civil_stat_cat", 
                          "occupation_cat", "htn_cat", "glucose_cat", 
                          "chronic_dis")
)

all_exposures <- c("exposure_pm25_z", "exposure_pm10_z", 
                   "exposure_temp_z", "exposure_humidity_z")

# Binary outcome
bin_full <- bind_rows(lapply(names(covariate_sets), function(model_name) {
  bind_rows(lapply(all_exposures, function(x) {
    run_cluster_model(
      data          = spatial_data_proj,
      outcome       = "dc_binary",
      exposure      = x,
      covariates    = covariate_sets[[model_name]],
      family        = binomial,
      exponentiate  = TRUE,
      outcome_label = "DC binary",
      keep_all_terms = TRUE        # <-- the key change
    ) |>
      mutate(model = model_name, exposure_var = x)
  }))
}))

# RUDAS (older adults)
rudas_full <- bind_rows(lapply(names(covariate_sets), function(model_name) {
  bind_rows(lapply(all_exposures, function(x) {
    run_cluster_model(
      data          = spatial_data_proj |> filter(age_group == "older"),
      outcome       = "rudas_n",
      exposure      = x,
      covariates    = covariate_sets[[model_name]],
      family        = NULL,
      exponentiate  = FALSE,
      outcome_label = "RUDAS",
      keep_all_terms = TRUE
    ) |>
      mutate(model = model_name, exposure_var = x)
  }))
}))

# ACE (younger adults)
ace_full <- bind_rows(lapply(names(covariate_sets), function(model_name) {
  bind_rows(lapply(all_exposures, function(x) {
    run_cluster_model(
      data          = spatial_data_proj |> filter(age_group == "younger"),
      outcome       = "ace_n",
      exposure      = x,
      covariates    = covariate_sets[[model_name]],
      family        = NULL,
      exponentiate  = FALSE,
      outcome_label = "ACE",
      keep_all_terms = TRUE
    ) |>
      mutate(model = model_name, exposure_var = x)
  }))
}))

susie_table <- bind_rows(bin_full, rudas_full, ace_full)
```
```{r}
susie_table |>
  mutate(
    exposure_var = recode(exposure_var,
      exposure_pm25_z     = "PM2.5",
      exposure_pm10_z     = "PM10",
      exposure_temp_z     = "Temperature",
      exposure_humidity_z = "Humidity"
    ),
    term = recode(term,
      "age_n"        = "Age",
      "sexMasculino" = "Sex (Male)",
      "ses_levelNSE D" = "SES: NSE D",
      "ses_levelNSE C" = "SES: NSE C",
      "ses_levelNSE B" = "SES: NSE B",
      "ses_levelNSE A" = "SES: NSE A (Highest)",
      "exposure_pm25_z"     = "PM2.5 (per SD)",
      "exposure_pm10_z"     = "PM10 (per SD)",
      "exposure_temp_z"     = "Temperature (per SD)",
      "exposure_humidity_z" = "Humidity (per SD)"
    ),
    estimate  = round(estimate, 3),
    conf.low  = round(conf.low, 3),
    conf.high = round(conf.high, 3),
    p.value   = round(p.value, 3)
  ) |>
  select(outcome, exposure_var, model, term, estimate, conf.low, conf.high, p.value) |>
  arrange(outcome, exposure_var, model) |>
  write.csv("susie_table_full_results.csv", row.names = FALSE)
```


## Models

\# Exposures

```{r}
exposures <- c("exposure_pm25_z",
               "exposure_pm10_z",
               "exposure_temp_z",
               "exposure_humidity_z")
```

\# Binary

```{r}
bin_results <- bind_rows(lapply(exposures, function(x) {

  run_cluster_model(
    data = spatial_data_proj,
    outcome = "dc_binary",
    exposure = x,
    covariates = c("age_n", "sex"),
    family = binomial,
    exponentiate = TRUE,
    outcome_label = "DC binary (clustered)"
  ) |>
    mutate(exposure = x)
}))
```

\# Rudas

```{r}
rudas_results <- bind_rows(lapply(exposures, function(x) {

  run_cluster_model(
    data = spatial_data_proj |> filter(age_group == "older"),
    outcome = "rudas_n",
    exposure = x,
    covariates = c("age_n", "sex"),
    family = NULL,
    exponentiate = FALSE,
    outcome_label = "RUDAS (clustered)"
  ) |>
    mutate(exposure = x)
}))
```

\# ACE

```{r}
ace_results <- bind_rows(lapply(exposures, function(x) {

  run_cluster_model(
    data = spatial_data_proj |> filter(age_group == "younger"),
    outcome = "ace_n",
    exposure = x,
    covariates = c("age_n", "sex"),
    family = NULL,
    exponentiate = FALSE,
    outcome_label = "ACE (clustered)"
  ) |>
    mutate(exposure = x)
}))
```

## Combine everything

```{r}
all_results <- bind_rows(
  bin_results,
  rudas_results,
  ace_results
) |>
  mutate(
    exposure = recode(exposure,
                      exposure_pm25_z = "PM2.5",
                      exposure_pm10_z = "PM10",
                      exposure_temp_z = "Temperature",
                      exposure_humidity_z = "Humidity")
  )

all_results <- all_results |>
  mutate(
    # reorder outcomes (facet order)
    outcome = factor(outcome,
                     levels = c("DC binary (clustered)",
                                "ACE (clustered)",
                                "RUDAS (clustered)")),

    # reorder exposures (y-axis order)
    exposure = factor(exposure,
                      levels = c("Temperature",
                                 "Humidity",
                                 "PM2.5",
                                 "PM10"))
  )
```

## Susie Table (Full model results)
```{r}
library(kableExtra)

susie_table |>
  mutate(
    exposure_var = recode(exposure_var,
      exposure_pm25_z     = "PM2.5",
      exposure_pm10_z     = "PM10",
      exposure_temp_z     = "Temperature",
      exposure_humidity_z = "Humidity"
    ),
    term = recode(term,
      "age_n"                          = "Age",
      "sexMasculino"                   = "Sex (Male)",
      "ses_levelNSE D"                 = "SES: NSE D",
      "ses_levelNSE C"                 = "SES: NSE C",
      "ses_levelNSE B"                 = "SES: NSE B",
      "ses_levelNSE A"                 = "SES: NSE A (Highest)",
      "education_catprimary"           = "Education: Primary",
      "education_catsecondary_or_higher" = "Education: Secondary+",
      "chronic_disSí"                  = "Chronic Disease",
      "htn_cathypertension"            = "Hypertension",
      "exposure_pm25_z"                = "PM2.5 (per SD)",
      "exposure_pm10_z"                = "PM10 (per SD)",
      "exposure_temp_z"                = "Temperature (per SD)",
      "exposure_humidity_z"            = "Humidity (per SD)"
    ),
    estimate  = round(estimate, 2),
    conf.low  = round(conf.low, 2),
    conf.high = round(conf.high, 2),
    p.value   = ifelse(p.value < 0.001, "<0.001", round(p.value, 3)),
    ci        = paste0("(", conf.low, ", ", conf.high, ")")
  ) |>
  select(outcome, exposure_var, model, term, estimate, ci, p.value) |>
  arrange(outcome, exposure_var, model) |>
  kbl(
    col.names = c("Outcome", "Exposure", "Model", "Covariate", 
                  "β / OR", "95% CI", "p-value"),
    caption = "Supplementary Table. Full model results — all covariates"
  ) |>
  kable_styling(bootstrap_options = c("striped", "condensed"), 
                full_width = FALSE, font_size = 11) |>
  collapse_rows(columns = 1:3, valign = "top")
```


## Forest Plot

```{r}
ggplot(all_results,
       aes(x = estimate, y = exposure, color = exposure)) +

  geom_vline(data = subset(all_results, outcome == "DC binary (clustered)"),
             aes(xintercept = 1),
             linetype = "dashed", color = "grey50") +

  geom_vline(data = subset(all_results, outcome != "DC binary (clustered)"),
             aes(xintercept = 0),
             linetype = "dashed", color = "grey50") +

  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2) +

  geom_point(size = 3) +

  facet_wrap(~ outcome, scales = "free_x", ncol = 3) +

  theme_minimal() +
  theme(legend.position = "none") +

  labs(
  title = "Association between environmental exposures and cognitive outcomes",
  subtitle = paste0(
    "Adjusted for age and sex; exposures standardized (per 1 SD)\n",
    "Binary outcome: odds ratios | Continuous outcomes: β coefficients\n",
    "Cluster-robust SEs at sector × conglomerado level"
  ),
  x = "Effect estimate",
  y = NULL
)
```

## Sensor zone comparison table to show descriptively whether zones with higher PM2.5 tend to have worse cognitive scores

\# No dose-response pattern is observed between PM2.5 and cognitive
decline. Among the four sensors with adequate sample size (Ventanilla
n=19 and Carab n=4 are too small), the highest-PM2.5 zone (Comas, 40.1
µg/m³) has the lowest rate of cognitive decline (9.3%) and the highest
RUDAS scores (11.3), while the lowest-PM2.5 zones cluster around 16%
decline. This is an ecological confounding artifact: the bulk of
participants are assigned to just two sensors — MiPeru (n=501) and
PuenteP Estadio (n=361) — which cover most of the study area and show
nearly identical PM2.5 values (\~32–34 µg/m³). There is effectively no
meaningful exposure gradient across participants. The sensor zones are
not exchangeable populations, and any apparent association between
sensor-assigned PM2.5 and cognition reflects compositional differences
between zones rather than a pollution effect. This demonstrates why
naive ecological exposure assignment produces misleading results, and
motivates future work using personal or indoor exposure monitoring.

This table is a valuable finding — it demonstrates why naive ecological
exposure assignment produces misleading results, and motivates future
work with better exposure characterization; a methodological lesson

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  group_by(nearest_sensor_name) |>
  summarise(
    pm25          = mean(exposure_pm25, na.rm = TRUE),
    pm10          = mean(exposure_pm10, na.rm = TRUE),
    temp          = mean(exposure_temp, na.rm = TRUE),
    hum          = mean(exposure_humidity, na.rm = TRUE),
    pct_dc        = mean(dc_binary, na.rm = TRUE) * 100,
    mean_rudas    = mean(rudas_n, na.rm = TRUE),
    mean_ace      = mean(ace_n, na.rm = TRUE),
    n             = n(),
    .groups       = "drop"
  ) |>
  arrange(pm25)
```

## Add in SES level

```{r}
# SES-ADJUSTED MODELS (PARALLEL TO EXISTING PIPELINE)

# --- 1. PM2.5 (SES-adjusted) ---
pm25_ses <- glm(
  dc_binary ~ exposure_pm25_z + age_n + sex + ses_level,
  data = spatial_data_proj |> st_drop_geometry() |> filter(!is.na(dc_binary)),
  family = binomial
)

pm25_ses_results <- tidy(pm25_ses, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term == "exposure_pm25_z") |>
  mutate(
    exposure = "PM2.5",
    model = "SES-adjusted",
    outcome = "DC binary"
  )


# --- 2. PM10 (SES-adjusted) ---
pm10_ses <- glm(
  dc_binary ~ exposure_pm10_z + age_n + sex + ses_level,
  data = spatial_data_proj |> st_drop_geometry() |> filter(!is.na(dc_binary)),
  family = binomial
)

pm10_ses_results <- tidy(pm10_ses, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term == "exposure_pm10_z") |>
  mutate(
    exposure = "PM10",
    model = "SES-adjusted",
    outcome = "DC binary"
  )


# --- 3. Temperature (SES-adjusted) ---
temp_ses <- glm(
  dc_binary ~ exposure_temp_z + age_n + sex + ses_level,
  data = spatial_data_proj |> st_drop_geometry() |> filter(!is.na(dc_binary)),
  family = binomial
)

temp_ses_results <- tidy(temp_ses, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term == "exposure_temp_z") |>
  mutate(
    exposure = "Temperature",
    model = "SES-adjusted",
    outcome = "DC binary"
  )


# --- 4. Humidity (SES-adjusted) ---
hum_ses <- glm(
  dc_binary ~ exposure_humidity_z + age_n + sex + ses_level,
  data = spatial_data_proj |> st_drop_geometry() |> filter(!is.na(dc_binary)),
  family = binomial
)

hum_ses_results <- tidy(hum_ses, conf.int = TRUE, exponentiate = TRUE) |>
  filter(term == "exposure_humidity_z") |>
  mutate(
    exposure = "Humidity",
    model = "SES-adjusted",
    outcome = "DC binary"
  )


# COMBINE (SES VERSION ONLY — SAFE, SEPARATE OBJECT)

all_results_ses <- bind_rows(
  pm25_ses_results,
  pm10_ses_results,
  temp_ses_results,
  hum_ses_results
)
```

No statistically significant associations were observed between any sensor-assigned environmental exposure (PM2.5, PM10, temperature, humidity) and cognitive outcomes in unadjusted or adjusted models. Age and sex adjusted estimates for the binary cognitive impairment outcome were: PM2.5 OR 0.91 (95% CI 0.75–1.11, p=0.370), PM10 OR 0.90 (95% CI 0.74–1.11, p=0.333), temperature OR 1.08 (95% CI 0.90–1.30, p=0.406), humidity OR 1.07 (95% CI 0.90–1.27, p=0.469). ACE models were null across all four exposures at all adjustment stages. One exception was humidity, which showed a significant association with RUDAS-PE scores in unadjusted (β −0.57, 95% CI −1.13 to −0.02, p=0.046) and age/sex adjusted (β −0.58, 95% CI −1.14 to −0.02, p=0.045) models; however, this attenuated to null after full covariate adjustment (β −0.42, 95% CI −0.95 to 0.11, p=0.122), suggesting the association was driven by sociodemographic factors rather than a direct meteorological effect on cognition.

## Highways

The study area runs along what looks like a major highway corridor (the
blue trunk road running north-south). Our participant points are
clustered along and near this road. The red/pink dementia cases don't
obviously cluster in one spot, but proximity to that trunk road varies
meaningfully across participants — unlike the PM2.5 values which barely
vary at all.

```{r}
highways <- st_read("/Users/gabriellebenoit/Documents/GitHub/peru_air/highways_export.gpkg")
colnames(highways)
table(highways$type)

# Filter to major roads only
major_roads <- highways |>
  filter(highway %in% c("trunk", "primary", "secondary"))

# Check how many features you kept
nrow(major_roads)
table(major_roads$highway)

# Reproject to match spatial_data_proj
major_roads_proj <- major_roads |> st_transform(32718)

# Compute distance from each participant to nearest major road (meters)
dist_matrix <- st_distance(spatial_data_proj |> st_transform(32718),
                           major_roads_proj)

spatial_data_proj <- spatial_data_proj |>
  mutate(dist_to_road_m = apply(dist_matrix, 1, min) |> as.numeric())

# Check distribution
summary(spatial_data_proj$dist_to_road_m)

```

Meaningful continuous variation from 9 meters to 4.6 km. - Median 674m —
half of the participants live within \~674m of a major road - Some very
close (9m) — essentially on the road - Max 4.5km — good spread at the
far end - Right-skewed (mean 1001 \>\> median 674m) — confirms the need
to log-transform

```{r}
# Add log distance and standardized versions
spatial_data_proj <- spatial_data_proj |>
  mutate(
    log_dist_road  = log(dist_to_road_m + 1),
    dist_road_z    = as.numeric(scale(dist_to_road_m)),
    near_road_100m = dist_to_road_m <= 100,
    near_road_500m = dist_to_road_m <= 500
  )


# --- Binary outcome: full sample ---
road_binary_results <- run_cluster_model(
  data = spatial_data_proj,
  outcome = "dc_binary",
  exposure = "log_dist_road",
  covariates = c("age_n", "sex"),
  family = binomial,
  exponentiate = TRUE,
  outcome_label = "DC binary (clustered)"
)

# --- RUDAS: older adults ---
road_rudas_results <- run_cluster_model(
  data = spatial_data_proj |> filter(age_group == "older"),
  outcome = "rudas_n",
  exposure = "log_dist_road",
  covariates = c("age_n", "sex"),
  family = NULL,
  exponentiate = FALSE,
  outcome_label = "RUDAS (clustered)"
)

# --- ACE: younger adults ---
road_ace_results <- run_cluster_model(
  data = spatial_data_proj |> filter(age_group == "younger"),
  outcome = "ace_n",
  exposure = "log_dist_road",
  covariates = c("age_n", "sex"),
  family = NULL,
  exponentiate = FALSE,
  outcome_label = "ACE (clustered)"
)


# --- Quick descriptive by buffer zone ---
spatial_data_proj |>
  st_drop_geometry() |>
  group_by(near_road_500m) |>
  summarise(
    pct_dc     = mean(dc_binary, na.rm = TRUE) * 100,
    mean_rudas = mean(rudas_n,   na.rm = TRUE),
    mean_ace   = mean(ace_n,     na.rm = TRUE),
    n          = n()
  )
```

This is the most interesting result yet.

Participants living within 500m of a major road have lower rates of
cognitive decline (12.2% vs 16.4%) and higher cognitive scores (RUDAS
8.29 vs 6.82, ACE 80.24 vs 78.1) — the opposite of what would be
expected if road proximity were harmful. This mirrors the ecological
confounding pattern seen with PM2.5: proximity to urban infrastructure
in this peri-urban Lima context likely reflects better access to
services, healthcare, and economic opportunity rather than greater
pollution burden.

In adjusted models, road proximity was not significantly associated with
the binary cognitive decline outcome (OR 1.07, 95% CI 0.93-1.24, p=0.33)
or with RUDAS scores in older adults (β -0.45, 95% CI -1.35-0.45,
p=0.32). The only statistically significant finding was for ACE in
younger adults: each unit increase in log-distance from a major road was
associated with a 0.92-point decrease in ACE score (β -0.92, 95% CI
-1.65 – -0.28, p=0.005). The direction of this effect – farther from
roads associated with worse cognition – is counterintuitive from a
pollution-harm perspective and instead consistent with urban access
advantages accruing to those living closer to major road infrastructure.

Why only younger adults? ACE has a higher ceiling and greater
sensitivity than RUDAS, and younger adults (30–59) have more cognitive
reserve and score variance — making subtle environmental gradients
detectable where RUDAS in older adults cannot resolve them. Older
adults' scores may already reflect age-related decline that obscures any
road proximity signal.

The direction of the ACE effect — farther from roads associates with
worse cognition — points to urban access as the operative mechanism in
this setting, not traffic-related pollution.

## View results

```{r}
bind_rows(
  road_binary_results,
  road_rudas_results,
  road_ace_results
) |>
  select(outcome, estimate, conf.low, conf.high, p.value) |>
  mutate(across(where(is.numeric), ~ round(., 3)))
```

## Figure

```{r}
# Buffer comparison table
buffer_summary <- spatial_data_proj |>
  st_drop_geometry() |>
  mutate(
    buffer_zone = case_when(
      dist_to_road_m <= 100  ~ "Within 100m",
      dist_to_road_m <= 500  ~ "100–500m",
      dist_to_road_m <= 1000 ~ "500m–1km",
      TRUE                   ~ "Beyond 1km"
    ),
    buffer_zone = factor(buffer_zone, 
                         levels = c("Within 100m", "100–500m", 
                                    "500m–1km", "Beyond 1km"))
  ) |>
  group_by(buffer_zone) |>
  summarise(
    pct_dc     = mean(dc_binary,  na.rm = TRUE) * 100,
    mean_rudas = mean(rudas_n,    na.rm = TRUE),
    mean_ace   = mean(ace_n,      na.rm = TRUE),
    n          = n(),
    .groups    = "drop"
  )

print(buffer_summary)


buffer_summary |>
  pivot_longer(cols = c(pct_dc, mean_rudas, mean_ace),
               names_to = "outcome", values_to = "value") |>
  mutate(outcome = recode(outcome,
    pct_dc     = "% cognitive impairment",
    mean_rudas = "mean RUDAS (out of 30)",
    mean_ace   = "mean ACE (out of 100)"
  ),
  outcome = factor(outcome, levels = c(
    "% cognitive impairment", "mean RUDAS (out of 30)", "mean ACE (out of 100)"
  ))) |>
  ggplot(aes(x = buffer_zone, y = value, fill = buffer_zone)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = round(value, 1)),
            vjust = -0.4, size = 3,
            color = "grey30") +
  facet_wrap(~ outcome, scales = "free_y") +
  scale_fill_manual(values = c(
    "Within 100m" = "#B5D4F4",
    "100–500m"    = "#85B7EB",
    "500m–1km"    = "#378ADD",
    "Beyond 1km"  = "#185FA5"
  )) +
  labs(
    title   = "Cognitive outcomes by distance to nearest major road",
    caption = "Distance to nearest trunk, primary, or secondary road (OpenStreetMap). Unadjusted means.",
    x = NULL, y = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 8, angle = 20, hjust = 1),
        strip.text  = element_text(size = 10))
```

## Request from Eder
"can you use a y axis for the maximum value (ie., 100%, 30 points, and 100 points)? Thus, the bars are more proportional, and the visual comparison is better. Additionally, it would be interesting to know if a stratified by sensor analysis identified important differences"

```{r}
buffer_summary |>
  pivot_longer(cols = c(pct_dc, mean_rudas, mean_ace),
               names_to = "outcome", values_to = "value") |>
  mutate(
    outcome = recode(outcome,
      pct_dc    = "% cognitive impairment",
      mean_rudas = "mean RUDAS (out of 30)",
      mean_ace   = "mean ACE (out of 100)"
    ),
    outcome = factor(outcome, levels = c(
      "% cognitive impairment", "mean RUDAS (out of 30)", "mean ACE (out of 100)"
    )),
    y_max = case_when(
      outcome == "% cognitive impairment" ~ 100,
      outcome == "mean RUDAS (out of 30)" ~ 30,
      outcome == "mean ACE (out of 100)"  ~ 100
    )
  ) |>
  ggplot(aes(x = buffer_zone, y = value, fill = buffer_zone)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = round(value, 1)), vjust = -0.4, size = 3, color = "grey30") +
  facet_wrap(~ outcome, scales = "free_y") +
  scale_y_continuous(limits = function(x) c(0, max(x) * 1.1)) +
  ggh4x::facetted_pos_scales(y = list(
    outcome == "% cognitive impairment" ~ scale_y_continuous(limits = c(0, 100)),
    outcome == "mean RUDAS (out of 30)" ~ scale_y_continuous(limits = c(0, 30)),
    outcome == "mean ACE (out of 100)"  ~ scale_y_continuous(limits = c(0, 100))
  )) +
  scale_fill_manual(values = c(
    "Within 100m" = "#B5D4F4",
    "100–500m"    = "#85B7EB",
    "500m–1km"    = "#378ADD",
    "Beyond 1km"  = "#185FA5"
  )) +
  labs(
    title   = "Cognitive outcomes by distance to nearest major road",
    caption = "Distance to nearest trunk, primary, or secondary road (OpenStreetMap). Unadjusted means.",
    x = NULL, y = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(size = 8, angle = 20, hjust = 1),
        strip.text  = element_text(size = 10))
```
```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  mutate(buffer_zone = case_when(
    dist_to_road_m <= 100  ~ "Within 100m",
    dist_to_road_m <= 500  ~ "100–500m",
    dist_to_road_m <= 1000 ~ "500m–1km",
    TRUE                   ~ "Beyond 1km"
  ),
  buffer_zone = factor(buffer_zone,
    levels = c("Within 100m", "100–500m", "500m–1km", "Beyond 1km"))) |>
  group_by(nearest_sensor_name, buffer_zone) |>
  summarise(
    pct_dc     = mean(dc_binary, na.rm = TRUE) * 100,
    mean_rudas = mean(rudas_n, na.rm = TRUE),
    mean_ace   = mean(ace_n, na.rm = TRUE),
    n          = n(),
    .groups    = "drop"
  ) |>
  pivot_longer(cols = c(pct_dc, mean_rudas, mean_ace),
               names_to = "outcome", values_to = "value") |>
  mutate(outcome = recode(outcome,
    pct_dc     = "% cognitive impairment",
    mean_rudas = "mean RUDAS (out of 30)",
    mean_ace   = "mean ACE (out of 100)"
  )) |>
  ggplot(aes(x = buffer_zone, y = value, fill = buffer_zone)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(round(value, 1), "\n(n=", n, ")")),
            vjust = -0.3, size = 2.5, color = "grey30") +
  facet_grid(nearest_sensor_name ~ outcome, scales = "free_y") +
  scale_fill_manual(values = c(
    "Within 100m" = "#B5D4F4",
    "100–500m"    = "#85B7EB",
    "500m–1km"    = "#378ADD",
    "Beyond 1km"  = "#185FA5"
  )) +
  labs(
    title   = "Cognitive outcomes by road distance, stratified by sensor zone",
    caption = "Unadjusted means. n per cell shown on bars.",
    x = NULL, y = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x    = element_text(size = 7, angle = 25, hjust = 1),
        strip.text     = element_text(size = 9))
```









## Add in SES level

```{r}
table(spatial_data_proj$ses_level, useNA = "always")

# Does SES predict cognitive decline?
spatial_data_proj |>
  st_drop_geometry() |>
  group_by(ses_level) |>
  summarise(
    pct_dc     = mean(dc_binary, na.rm = TRUE) * 100,
    mean_rudas = mean(rudas_n,   na.rm = TRUE),
    mean_ace   = mean(ace_n,     na.rm = TRUE),
    n          = n()
  ) |>
  arrange(ses_level)

# Is SES associated with exposure variables?
spatial_data_proj |>
  st_drop_geometry() |>
  group_by(ses_level) |>
  summarise(
    mean_pm25      = mean(exposure_pm25,  na.rm = TRUE),
    mean_dist_road = mean(dist_to_road_m, na.rm = TRUE),
    n              = n()
  ) |>
  arrange(ses_level)

```

# Re-run models

```{r}
# --- Binary + SES ---
road_binary_ses <- run_cluster_model(
  data = spatial_data_proj,
  outcome = "dc_binary",
  exposure = "log_dist_road",
  covariates = c("age_n", "sex", "ses_level"),
  family = binomial,
  exponentiate = TRUE,
  outcome_label = "DC binary (SES, clustered)"
)

# --- RUDAS + SES ---
road_rudas_ses <- run_cluster_model(
  data = spatial_data_proj |> filter(age_group == "older"),
  outcome = "rudas_n",
  exposure = "log_dist_road",
  covariates = c("age_n", "sex", "ses_level"),
  outcome_label = "RUDAS (SES, clustered)"
)

# --- ACE + SES ---
road_ace_ses <- run_cluster_model(
  data = spatial_data_proj |> filter(age_group == "younger"),
  outcome = "ace_n",
  exposure = "log_dist_road",
  covariates = c("age_n", "sex", "ses_level"),
  outcome_label = "ACE (SES, clustered)"
)






# Comparison table: before vs after SES 
bind_rows(
  road_binary_results |> mutate(model = "without SES"),
  road_binary_ses     |> mutate(model = "with SES"),
  road_rudas_results  |> mutate(model = "without SES"),
  road_rudas_ses      |> mutate(model = "with SES"),
  road_ace_results    |> mutate(model = "without SES"),
  road_ace_ses        |> mutate(model = "with SES")
) |>
  select(outcome, model, estimate, conf.low, conf.high, p.value) |>
  mutate(across(where(is.numeric), ~ round(., 3)))
```

In adjusted models, road proximity is null for both the binary outcome
(OR 1.07 without SES, OR 1.04 with SES, both p\>0.33) and RUDAS in older
adults (β −0.45 without SES, β −0.30 with SES, both p\>0.30). SES does
not meaningfully change either null result.

For ACE in younger adults, the road proximity effect attenuates modestly
after SES adjustment (β −0.92 without SES vs β −0.78 with SES) but
remains statistically significant (p=0.013). SES accounts for roughly
15% of the original effect estimate — consistent with the descriptive
finding that lower-SES participants tend to live slightly farther from
major roads — but does not explain it away. The association remains
statistically significant after SES adjustment (β −0.785, 95% CI −1.40–
−0.17, p=0.013). The association between greater distance from major
roads and lower ACE scores in 30–59 year olds is robust to socioeconomic
adjustment.

This pattern suggests the road proximity effect in younger adults is not
simply a proxy for SES, though both urbanization and socioeconomic
resources likely operate through related pathways — access to
healthcare, education, and economic opportunity — in this peri-urban
Lima context.

## The *storyline* for the chapter:

1.  PM2.5 — null, insufficient spatial variability. Only 6 sensors cover
    the study area; most participants are assigned to just two, with a
    PM2.5 range of roughly 32–40 µg/m³ across all zones. There is no
    dose-response relationship. The apparent inverse pattern
    (highest-PM2.5 zone has best cognition) is an ecological confounding
    artifact reflecting population heterogeneity across sensor zones,
    not a protective effect of pollution.

2.  Distance to road — significant for ACE in younger adults, robust to
    SES adjustment. The β attenuates from −0.92 to −0.78 after
    controlling for SES level, but remains significant (p=0.013). SES is
    a strong independent predictor of cognition across all outcomes, but
    explains only \~15% of the road proximity effect, indicating the two
    operate through partially overlapping but distinct pathways.

3.  Consistent interpretation across both exposures — urbanization as
    confounder. Both findings tell the same story: in peri-urban Lima,
    living closer to roads and higher-pollution urban centers correlates
    with better, not worse, cognitive outcomes. Structural access — to
    healthcare, education, economic resources — likely dominates any
    pollution effect at the exposure levels and spatial resolution
    available in this study.

4.  Methodological implication. Future studies in similar peri-urban
    settings require personal or residential-level exposure monitors,
    fine-grained SES adjustment, and larger sensor networks to
    disentangle pollution effects from the urbanization gradient. The
    current findings are best framed as a demonstration of ecological
    confounding and the limits of sensor-assigned exposure, rather than
    evidence of null pollution effects per se.

*Implication — future studies need personal, granular exposure
monitors + SES adjustment to disentangle pollution from access; indoor
air monitoring*

```{r}
summary_table <- spatial_data_proj %>%
  group_by(nearest_sensor_name) %>%
  summarise(
    PM2.5      = mean(exposure_pm25, na.rm = TRUE),
    `%DC`      = mean(dc_binary, na.rm = TRUE) * 100,
    Mean_RUDAS = mean(rudas[!is.na(rudas) & rudas != 0]),
    SD_RUDAS   = sd(rudas[!is.na(rudas) & rudas != 0]),
    Mean_ACE   = mean(ace_score[!is.na(ace_score) & ace_score != 0]),
    SD_ACE     = sd(ace_score[!is.na(ace_score) & ace_score != 0]),
    n          = n(),
    .groups = "drop"
  )
```

## Summary table of exposure variables

```{r}
spatial_data_proj %>%
  summarise(
    Mean_ACE         = mean(ace_score[!is.na(ace_score) & ace_score != 0], na.rm = TRUE),
    SD_ACE           = sd(ace_score[!is.na(ace_score) & ace_score != 0], na.rm = TRUE),
    n_ACE            = sum(!is.na(ace_score) & ace_score != 0),

    Mean_RUDAS       = mean(rudas[!is.na(rudas) & rudas != 0], na.rm = TRUE),
    SD_RUDAS         = sd(rudas[!is.na(rudas) & rudas != 0], na.rm = TRUE),
    n_RUDAS          = sum(!is.na(rudas) & rudas != 0),

    Mean_Humidity    = mean(exposure_humidity, na.rm = TRUE),
    SD_Humidity      = sd(exposure_humidity, na.rm = TRUE),
    n_Humidity       = sum(!is.na(exposure_humidity)),

    Mean_Temp        = mean(exposure_temp, na.rm = TRUE),
    SD_Temp          = sd(exposure_temp, na.rm = TRUE),
    n_Temp           = sum(!is.na(exposure_temp)),

    Mean_PM25        = mean(exposure_pm25, na.rm = TRUE),
    SD_PM25          = sd(exposure_pm25, na.rm = TRUE),
    n_PM25           = sum(!is.na(exposure_pm25)),

    Mean_PM10        = mean(exposure_pm10, na.rm = TRUE),
    SD_PM10          = sd(exposure_pm10, na.rm = TRUE),
    n_PM10           = sum(!is.na(exposure_pm10))
  )
```

## Summary tables by sensor

```{r}
# define variable groups
continuous_vars <- c("age", "bmi")
categorical_vars <- c(
  "sex", "education_cat", "civil_stat_cat",
  "htn_cat", "native_lang", "health_ins_cat",
  "occupation_cat", "chronic_dis", "ses_level"
)

# ---- continuous summaries ----
cont_summary <- spatial_data_proj |>
  st_drop_geometry() |>
  group_by(nearest_sensor_name) |>
  summarise(
    across(all_of(continuous_vars),
           ~ mean(.x, na.rm = TRUE),
           .names = "mean_{.col}"),
    n = n(),
    .groups = "drop"
  )

# ---- categorical summaries (percent breakdowns) ----
cat_summary <- spatial_data_proj |>
  st_drop_geometry() |>
  group_by(nearest_sensor_name) |>
  select(nearest_sensor_name, all_of(categorical_vars)) |>
  pivot_longer(-nearest_sensor_name,
               names_to = "variable",
               values_to = "value") |>
  group_by(nearest_sensor_name, variable, value) |>
  summarise(n = n(), .groups = "drop_last") |>
  mutate(pct = 100 * n / sum(n)) |>
  ungroup()

# ---- optional: wide format for categorical ----
cat_summary_wide <- cat_summary |>
  unite(var_level, variable, value, sep = "_") |>
  pivot_wider(names_from = var_level, values_from = pct)

# ---- combine everything ----
final_summary <- cont_summary |>
  left_join(cat_summary_wide, by = "nearest_sensor_name") |>
  arrange(nearest_sensor_name)



# prettier
# Assign table to an object
my_table <- spatial_data_proj |>
  st_drop_geometry() |>
  select(
    nearest_sensor_name, age, sex, ace_score, rudas, education_cat, civil_stat_cat,
    bmi, htn_cat, native_lang, health_ins_cat, occupation_cat, chronic_dis, ses_level
  ) |>
  tbl_summary(
    by = nearest_sensor_name,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = all_continuous() ~ 1,
    missing = "ifany"
  ) |>
  add_n() |>
  add_p(
    # Fix for FEXACT errors: simulate p-values for Fisher's tests
    test.args = all_tests("fisher.test") ~ list(simulate.p.value = TRUE)
  ) |>
  bold_labels() |>
  bold_p()


# Export to Excel
gtsummary::as_hux_xlsx(my_table, file = "descriptives_by_sensor_table.xlsx")



```

## CURRENTLY NOT INCLUDING THIS IN THE MANUSCRIPT !!!!!

## Change exposure gradient

\# clean exposure variable

```{r}
spatial_data_proj <- spatial_data_proj |>
  mutate(
    log_dist_road = log(dist_to_road_m + 1),
    
    road_exposure = case_when(
      dist_to_road_m <= 100 ~ "≤100m",
      dist_to_road_m <= 500 ~ "100–500m",
      TRUE ~ ">500m"
    ),
    
    road_exposure = factor(
      road_exposure,
      levels = c("≤100m", "100–500m", ">500m")
    )
  )

spatial_data_proj <- spatial_data_proj |>
  mutate(
    road_exposure = relevel(factor(road_exposure), ref = ">500m")
  )

```

\# primary model (continuous exposure)

```{r}
model_continuous <- glm(
  dc_binary ~ log_dist_road + age_n + sex,
  data = spatial_data_proj,
  family = binomial
)

continuous_results <- tidy(model_continuous, exponentiate = TRUE, conf.int = TRUE)

continuous_results
```

\# sensitivity model (buffer categories)

```{r}


model_buffer <- glm(
  dc_binary ~ road_exposure + age_n + sex,
  data = spatial_data_proj,
  family = binomial
)

buffer_results <- tidy(model_buffer, exponentiate = TRUE, conf.int = TRUE)

buffer_results



```

\# cleaned combined output

```{r}
bind_rows(
  continuous_results |> mutate(model = "Continuous (log distance)"),
  buffer_results |> mutate(model = "Buffer zones")
) |>
  select(model, term, estimate, conf.low, conf.high, p.value)



bind_rows(
  continuous_results |> mutate(model = "Continuous"),
  buffer_results |> mutate(model = "Buffer zones")
) |>
  mutate(across(c(estimate, conf.low, conf.high), ~round(.x, 3))) |>
  select(model, term, estimate, conf.low, conf.high, p.value)
```

\# predicted probability plot

```{r}
newdat <- expand.grid(
  log_dist_road = seq(
    min(spatial_data_proj$log_dist_road, na.rm = TRUE),
    max(spatial_data_proj$log_dist_road, na.rm = TRUE),
    length.out = 100
  ),
  age_n = mean(spatial_data_proj$age_n, na.rm = TRUE),
  sex = unique(spatial_data_proj$sex)[1]
)

newdat$pred <- predict(
  model_continuous,
  newdata = newdat,
  type = "response"
)

ggplot(newdat, aes(x = log_dist_road, y = pred)) +
  geom_line() +
  labs(
    title = "Predicted Probability of Cognitive Impairment by Road Proximity",
    x = "Log Distance to Road (meters)",
    y = "Predicted Probability"
  ) +
  theme_minimal()
```

\# simple buffer comparison

```{r}
ggplot(spatial_data_proj, aes(x = road_exposure, y = dc_binary)) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun = mean, geom = "line", aes(group = 1)) +
  labs(
    title = "Cognitive Impairment by Road Proximity",
    x = "Distance to Road",
    y = "Mean Probability of Cognitive Impairment"
  ) +
  theme_minimal()
```

## Re-run models for each road type

```{r}
# --- Compute distance to each road type separately ---
for (road_type in c("trunk", "primary", "secondary")) {
  roads_subset <- major_roads |> filter(highway == road_type)
  dist_mat <- st_distance(spatial_data_proj |> st_transform(32718), roads_subset)
  spatial_data_proj <- spatial_data_proj |>
    mutate(!!paste0("dist_", road_type, "_m") := apply(dist_mat, 1, min) |> as.numeric())
}

# Log-transform each
spatial_data_proj <- spatial_data_proj |>
  mutate(
    log_dist_trunk     = log(dist_trunk_m     + 1),
    log_dist_primary   = log(dist_primary_m   + 1),
    log_dist_secondary = log(dist_secondary_m + 1)
  )

# --- Run models for each road type x outcome ---
road_sensitivity <- bind_rows(
  lapply(c("trunk", "primary", "secondary"), function(rt) {
    exp_var <- paste0("log_dist_", rt)
    bind_rows(
      run_cluster_model(
        data = spatial_data_proj,
        outcome = "dc_binary", exposure = exp_var,
        covariates = c("age_n", "sex", "ses_level"),
        family = binomial, exponentiate = TRUE,
        outcome_label = paste0("DC binary — ", rt)
      ),
      run_cluster_model(
        data = spatial_data_proj |> filter(age_group == "older"),
        outcome = "rudas_n", exposure = exp_var,
        covariates = c("age_n", "sex", "ses_level"),
        outcome_label = paste0("RUDAS — ", rt)
      ),
      run_cluster_model(
        data = spatial_data_proj |> filter(age_group == "younger"),
        outcome = "ace_n", exposure = exp_var,
        covariates = c("age_n", "sex", "ses_level"),
        outcome_label = paste0("ACE — ", rt)
      )
    )
  })
)

# --- View comparison table ---
road_sensitivity |>
  select(outcome, estimate, conf.low, conf.high, p.value) |>
  mutate(across(where(is.numeric), ~ round(., 3)))
```

In road-type stratified sensitivity analyses, the inverse association
between road proximity and ACE scores in younger adults was consistent
for trunk (β −0.87, p=0.048) and secondary roads (β −0.83, p=0.007),
suggesting the pooled result is not attributable to any single road
classification. The primary road estimate was in the opposite direction
but did not reach statistical significance (β +1.15, p=0.081), possibly
reflecting the distinct spatial distribution of primary roads in this
peri-urban context.

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  group_by(nearest_sensor_name) |>
  summarise(
    n                    = n(),
    mean_dist_any        = mean(dist_to_road_m,     na.rm = TRUE),
    mean_dist_trunk      = mean(dist_trunk_m,       na.rm = TRUE),
    mean_dist_primary    = mean(dist_primary_m,     na.rm = TRUE),
    mean_dist_secondary  = mean(dist_secondary_m,   na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(starts_with("mean_dist"), ~ round(., 0))) |>
  arrange(mean_dist_any)
```

## More descriptive statisitcs [yes, including in manuscript]

For continuous covariates by sensor i.e. age, BMI

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  group_by(nearest_sensor_name) |>
  summarise(
    n = n(),
    age_mean = mean(age, na.rm = TRUE),
    age_sd = sd(age, na.rm = TRUE),
    bmi_mean = mean(bmi, na.rm = TRUE),
    bmi_sd = sd(bmi, na.rm = TRUE)
  )
```

\# Categorical covariates by sensor (i.e., sex, education, insurance)

```{r}
# sex
spatial_data_proj |>
  st_drop_geometry() |>
  count(nearest_sensor_name, sex) |>
  group_by(nearest_sensor_name) |>
  mutate(percent = n / sum(n) * 100)

# education_cat
spatial_data_proj |>
  st_drop_geometry() |>
  count(nearest_sensor_name, education_cat) |>
  group_by(nearest_sensor_name) |>
  mutate(percent = n / sum(n) * 100)
```

\# Checking multiple variables \>\> stacked descriptive dataset

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  select(nearest_sensor_name,
         sex, education_cat, civil_stat_cat,
         occupation_cat, health_ins_cat,
         htn_cat, chronic_dis, native_lang,
         ses_level) |>
  pivot_longer(-nearest_sensor_name,
               names_to = "variable",
               values_to = "value") |>
  count(nearest_sensor_name, variable, value) |>
  group_by(nearest_sensor_name, variable) |>
  mutate(percent = n / sum(n) * 100)
```

\# Cleaner formatting

```{r}
library(gtsummary)
spatial_data_proj |>
  st_drop_geometry() |>
  select(
    nearest_sensor_name,
    age, sex, education_cat, civil_stat_cat,
    bmi, htn_cat, native_lang,
    health_ins_cat, occupation_cat,
    chronic_dis, ses_level
  ) |>
  tbl_summary(by = nearest_sensor_name)
```

\# Quick diagnostic visualization Super important: my exposure
assignment is sensor-based, so if covariates cluster by sensor, it
suggests potential spatial confounding (e.g., SES or age differences by
neighborhood).

However, this quick visual does not suggest spatial confounding by age.
Medians are fairly similar (\~45-55 years). Interquartile ranges overlap
a lot. Some sensors do have longer upper tails. -
GEOHealth_Carab_VictorRaulHT → slightly older median (\~52–53) -
GEOHealth_Ventanilla_NSBelen → slightly younger (\~45–47)

-   A few sensors (e.g., PuenteP_Estadio_1, MiPeru_Chavinillo) extend to
    \~90+. a few older participants near those sensors, but since the
    median and IQR are similar, this is unlikely to drive systematic
    confounding.

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  ggplot(aes(nearest_sensor_name, age)) +
  geom_boxplot() +
  coord_flip()
```

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  count(nearest_sensor_name, ses_quint) |>
  group_by(nearest_sensor_name) |>
  mutate(p = n / sum(n))

spatial_data_proj |>
  st_drop_geometry() |>
  count(nearest_sensor_name, ses_level) |>
  group_by(nearest_sensor_name) |>
  mutate(prop = n / sum(n)) |>
  ggplot(aes(x = nearest_sensor_name, y = prop, fill = factor(ses_level))) +
  geom_col() +
  coord_flip() +
  scale_x_discrete(labels = c(
    "GEOHealth_Ventanilla_NSBelen" = "Ventanilla",
    "GEOHealth_SanMa_IE005" = "SanMa",
    "GEOHealth_PuenteP_Estadio_1" = "PuenteP Estadio",
    "GEOHealth_MiPeru_Chavinillo" = "MiPeru",
    "GeoHealth_Comas_2055PAbril" = "Comas",
    "GEOHealth_Carab_VictorRaulHT" = "Carab"
  )) +
  labs(
    x = "Sensor",
    y = "Proportion",
    fill = "SES Level",
    title = "Socioeconomic Status Distribution by Nearest Sensor"
  ) +
  theme_minimal()
```
```{r}
ses_table <- spatial_data_proj |>
  st_drop_geometry() |>
  count(nearest_sensor_name, ses_level) |>
  group_by(nearest_sensor_name) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  select(-n) |>
  pivot_wider(
    names_from = ses_level,
    values_from = prop
  ) |>
  mutate(
    nearest_sensor_name = recode(nearest_sensor_name,
      "GEOHealth_Ventanilla_NSBelen" = "Ventanilla",
      "GEOHealth_SanMa_IE005" = "SanMa",
      "GEOHealth_PuenteP_Estadio_1" = "PuenteP Estadio",
      "GEOHealth_MiPeru_Chavinillo" = "MiPeru",
      "GeoHealth_Comas_2055PAbril" = "Comas",
      "GEOHealth_Carab_VictorRaulHT" = "Carab"
    )
  )

ses_table
```



## Chi-Square / Kruskal-Wallis Tests by Sector

```{r}
# 1. Categorical variables → Chi-square
cat_vars <- c("sex", "education_cat", "civil_stat_cat", "native_lang",
              "health_ins_cat", "occupation_cat", "chronic_dis", "glucose_cat",
              "htn_cat", "ses_level")

chi_results <- map_dfr(cat_vars, function(var) {
  tbl <- table(spatial_data_proj[[var]], spatial_data_proj$nearest_sensor_name, useNA = "no")
  test <- chisq.test(tbl, simulate.p.value = TRUE, B = 2000)
  tibble(
    variable  = var,
    chi_sq    = round(test$statistic, 2),
    df        = test$parameter,
    p_value   = round(test$p.value, 4),
    sig       = case_when(p_value < 0.001 ~ "***",
                          p_value < 0.01  ~ "**",
                          p_value < 0.05  ~ "*",
                          TRUE            ~ "")
  )
})

# 2. Continuous variables → Kruskal-Wallis (non-parametric; safer for skewed data)
cont_vars <- c("age", "bmi", "rudas", "ace_score", "exposure_pm25", "exposure_pm10", "exposure_temp", "exposure_humidity")

kw_results <- map_dfr(cont_vars, function(var) {
  test <- kruskal.test(spatial_data_proj[[var]] ~ spatial_data_proj$nearest_sensor_name)
  tibble(
    variable = var,
    chi_sq   = round(test$statistic, 2),
    df       = test$parameter,
    p_value  = round(test$p.value, 4),
    sig      = case_when(p_value < 0.001 ~ "***",
                         p_value < 0.01  ~ "**",
                         p_value < 0.05  ~ "*",
                         TRUE            ~ "")
  )
})

# 3. Combine and display
balance_tests <- bind_rows(
  chi_results %>% mutate(test = "Chi-square"),
  kw_results  %>% mutate(test = "Kruskal-Wallis")
) %>% select(variable, test, chi_sq, df, p_value, sig)

knitr::kable(balance_tests, caption = "Covariate balance across sensor zones")
```

No covariate differs significantly across sensor zones (all p \> 0.05).
This means sensor-zone exposure assignment is not systematically
confounded by who lives where — age, SES, education, civil status, etc.
are distributed comparably across zones, strengthening our argument.

the NA in the df column for chi-square is expected; it happens because
we used simulate.p.value = TRUE, which runs a Monte Carlo permutation
test instead of the standard asymptotic chi-square. The simulated
version doesn't return degrees of freedom. Potential footnote: p-values
estimated via Monte Carlo simulation (B = 2000) due to small expected
cell counts in some zones.

Add to manuscript: To assess potential spatial confounding, we tested
whether key sociodemographic and health covariates differed across
sensor zones using chi-square tests (categorical variables) and
Kruskal-Wallis tests (continuous variables). No significant differences
were observed (all p \> 0.05; Supplementary Table X), suggesting that
exposure assignment was not systematically associated with participant
characteristics.

New figure with error bars for: Distribution of Dementia Status by SES
Level

```{r}
prop_data <- spatial_data_proj %>%
  count(any_relevant_dem, ses_level) %>%
  group_by(any_relevant_dem) %>%
  mutate(total = sum(n), prop = n / total) %>%
  rowwise() %>%
  mutate(
    lower = binom.test(n, total)$conf.int[1],
    upper = binom.test(n, total)$conf.int[2]
  ) %>%
  ungroup() %>%
mutate(
  any_relevant_dem = dplyr::case_match(any_relevant_dem,
    "Sin DC severo" ~ "Cognitively Unimpaired",
    "DC severo con compromiso funcional" ~ "Cognitive Impairment with Functional Impairment",
    "DC severo sin compromiso funcional" ~ "Cognitive Impairment without Functional Impairment"
  ),
  ses_level = dplyr::case_match(ses_level,
    "NSE A" ~ "A (Highest)",
    "NSE B" ~ "B",
    "NSE C" ~ "C",
    "NSE D" ~ "D",
    "NSE E" ~ "E (Lowest)"
  ),
  ses_level = forcats::fct_relevel(ses_level, "A (Highest)", "B", "C", "D", "E (Lowest)")
  ) %>%
  drop_na(ses_level)

ggplot(prop_data, aes(x = prop, y = ses_level, color = ses_level)) +
  geom_pointrange(aes(xmin = lower, xmax = upper), linewidth = 0.8, size = 0.7) +
  facet_wrap(~ any_relevant_dem, ncol = 1,          # strips on TOP now
             labeller = label_wrap_gen(width = 35)) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1),
                     breaks = c(0, 0.25, 0.5, 0.75, 1)) +
  labs(
    title   = "Proportion by SES Level across Dementia Status",
    x       = "Proportion (95% CI)",
    y       = NULL,
    caption = "Error bars: 95% exact binomial confidence intervals"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position   = "none",
    strip.text        = element_text(face = "bold", size = 11, hjust = 0),  # left-aligned strip titles
    strip.background  = element_rect(fill = "grey94", color = NA),          # subtle background box
    panel.spacing     = unit(1.2, "lines"),
    panel.grid.minor  = element_blank(),
    panel.grid.major.y = element_blank(),   # remove horizontal gridlines — cleaner for dot plots
    axis.text.y       = element_text(size = 10)
  )
```

```{r}
prop_data <- spatial_data_proj %>%
  count(any_relevant_dem, ses_level) %>%
  group_by(any_relevant_dem) %>%
  mutate(total = sum(n), prop = n / total) %>%
  rowwise() %>%
  mutate(
    lower = binom.test(n, total)$conf.int[1],
    upper = binom.test(n, total)$conf.int[2]
  ) %>%
  ungroup() %>%
  mutate(
    any_relevant_dem = recode(any_relevant_dem,
      "Sin DC severo"                          = "Cognitively Unimpaired",
      "DC severo con compromiso funcional"     = "Cognitive Impairment with Functional Impairment",
      "DC severo sin compromiso funcional"     = "Cognitive Impairment without Functional Impairment"
    ),
    ses_level = recode(ses_level,
      "NSE A" = "A (Highest)",
      "NSE B" = "B",
      "NSE C" = "C",
      "NSE D" = "D",
      "NSE E" = "E (Lowest)"
    ),
    ses_level = fct_relevel(ses_level, "A (Highest)", "B", "C", "D", "E (Lowest)")
  ) %>%
  drop_na(ses_level)

ggplot(prop_data, aes(x = any_relevant_dem, y = prop, fill = ses_level)) +
  geom_col(width = 0.6) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, linewidth = 0.7, color = "gray30") +
  facet_wrap(~ ses_level, nrow = 1) +
  scale_x_discrete(labels = \(x) str_wrap(x, width = 15)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title   = "Proportion by Dementia Status, Faceted by SES Level",
    x       = NULL,
    y       = "Proportion (95% CI)",
    caption = "Error bars: 95% exact binomial confidence intervals"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position   = "none",
    strip.text        = element_text(face = "bold", size = 11),
    panel.spacing     = unit(1.5, "lines"),
    panel.grid.minor  = element_blank(),
    axis.text.x       = element_text(size = 9, lineheight = 0.9),
    plot.margin       = margin(b = 20, t = 10, l = 10, r = 10)
  )
```

```{r}
decline_by_ses <- spatial_data_proj %>%
  st_drop_geometry() %>%
  mutate(
    decline = any_relevant_dem != "Sin DC severo"  # or define your preferred cutoff
  ) %>%
  group_by(ses_level) %>%
  summarise(
    n = n(),
    n_decline = sum(decline, na.rm = TRUE),
    prev_decline = n_decline / n
  ) %>%
  ungroup()

## add confidence intervals
decline_by_ses <- decline_by_ses %>%
  mutate(
    lower = purrr::map2_dbl(n_decline, n, ~ binom.test(.x, .y)$conf.int[1]),
    upper = purrr::map2_dbl(n_decline, n, ~ binom.test(.x, .y)$conf.int[2])
  )

## decline range
decline_range <- decline_by_ses %>%
  summarise(
    min_prev = min(prev_decline, na.rm = TRUE),
    max_prev = max(prev_decline, na.rm = TRUE)
  )

## Mean ACE score by SES
ace_by_ses <- spatial_data_proj %>%
  st_drop_geometry() %>%
  group_by(ses_level) %>%
  summarise(
    mean_ace = mean(ace_score, na.rm = TRUE),
    sd_ace = sd(ace_score, na.rm = TRUE),
    n = n()
  ) %>%
  ungroup()

## Get the difference across SES levels
ace_range <- ace_by_ses %>%
  summarise(
    min_ace = min(mean_ace, na.rm = TRUE),
    max_ace = max(mean_ace, na.rm = TRUE),
    diff_ace = max_ace - min_ace
  )

summary_ace_table <- decline_by_ses %>%
  left_join(ace_by_ses, by = "ses_level")




## Mean rudas score by SES
rudas_by_ses <- spatial_data_proj %>%
  st_drop_geometry() %>%
  group_by(ses_level) %>%
  summarise(
    mean_rudas = mean(rudas_n, na.rm = TRUE),
    sd_rudas = sd(rudas_n, na.rm = TRUE),
    n = n()
  ) %>%
  ungroup()

## Get the difference across SES levels
rudas_range <- rudas_by_ses %>%
  summarise(
    min_rudas = min(mean_rudas, na.rm = TRUE),
    max_rudas = max(mean_rudas, na.rm = TRUE),
    diff_rudas = max_rudas - min_rudas
  )

summary_rudas_table <- decline_by_ses %>%
  left_join(rudas_by_ses, by = "ses_level")

```

## Map figure

```{r}
library(tmap)

# Summarise mean PM2.5 per sensor, keeping sensor geometry
sensor_summary <- spatial_data_proj |>
  group_by(nearest_sensor_id) |>
  summarise(
    mean_pm25 = mean(exposure_pm25, na.rm = TRUE),
    n = n(),
    geometry = st_union(geometry) |> st_centroid()  # just takes a representative point
  ) |>
  st_as_sf(crs = 32718)

# If you have a separate sensors sf object with one row per sensor, even better:
# sensor_summary <- sensors_sf |> left_join(...)

# Map
tm_shape(spatial_data_proj |> st_transform(4326)) +
  tm_dots(col = "grey80", size = 0.05, alpha = 0.5) +
tm_shape(sensor_summary |> st_transform(4326)) +
  tm_bubbles(
    size = "n",
    col = "mean_pm25",
    palette = "YlOrRd",
    title.col = "Mean PM2.5 (µg/m³)",
    title.size = "N participants",
    scale = 1.5
  ) +
  tm_text("nearest_sensor_id", size = 0.6, ymod = 0.8) +
  tm_layout(
    title = "Mean PM2.5 by sensor location",
    legend.outside = TRUE
  )

```

## Does SES predict road proximity?

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  group_by(ses_level) |>
  summarise(
    mean_dist_road = mean(dist_to_road_m, na.rm = TRUE),
    median_dist_road = median(dist_to_road_m, na.rm = TRUE),
    n = n()
  ) |>
  arrange(ses_level)

```

## Is the SES-road relationship statistically significant?

```{r}
summary(lm(dist_to_road_m ~ ses_level, data = spatial_data_proj |> st_drop_geometry()))
```

## Does SES vary within road distance buffers?

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  mutate(buffer_zone = case_when(
    dist_to_road_m <= 100  ~ "Within 100m",
    dist_to_road_m <= 500  ~ "100–500m",
    dist_to_road_m <= 1000 ~ "500m–1km",
    TRUE                   ~ "Beyond 1km"
  )) |>
  group_by(buffer_zone, ses_level) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(names_from = ses_level, values_from = n, values_fill = 0)
```

## Colinearity with temp, humidity, pm 2.5, pm 10

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  select(exposure_pm25, exposure_pm10, exposure_temp, exposure_humidity) |>
  cor(use = "complete.obs") |>
  round(2)

vif_model <- lm(
  ace_n ~ exposure_pm25_z + exposure_pm10_z + exposure_temp_z + exposure_humidity_z + age_n + sex,
  data = spatial_data_proj |> st_drop_geometry() |> filter(age_group == "younger")
)

vif(vif_model)
```

## Correlations?

```{r}

spatial_data_proj |>
  st_drop_geometry() |>
  mutate(
    ses_numeric        = as.numeric(ses_level),  # already ordered factor, this works
    sex_numeric        = ifelse(sex == "Femenino", 1, 0),
    chronic_dis_numeric = ifelse(chronic_dis == "Sí", 1, 0),
    education_numeric  = as.numeric(education_cat),
    health_ins_numeric = as.numeric(health_ins_cat),
    occupation_numeric = as.numeric(occupation_cat)
  ) |>
  select(dist_to_road_m, ses_numeric, age_n, sex_numeric,
         dc_binary, rudas_n, ace_n,
         education_numeric, health_ins_numeric, occupation_numeric,
         chronic_dis_numeric, bmi) |>
  cor(use = "complete.obs", method = "spearman") |>
  round(2)

```

## road proximity table
```{r}
covariate_sets <- list(
  "Unadjusted"         = NULL,
  "Age & sex adjusted" = c("age_n", "sex"),
  "Fully adjusted"     = c("age_n", "sex", "ses_level", "bmi",
                            "education_cat", "civil_stat_cat",
                            "occupation_cat", "htn_cat", "glucose_cat",
                            "chronic_dis")
)

road_full <- bind_rows(lapply(names(covariate_sets), function(model_name) {
  bind_rows(list(
    run_cluster_model(
      data           = spatial_data_proj,
      outcome        = "dc_binary",
      exposure       = "log_dist_road",
      covariates     = covariate_sets[[model_name]],
      family         = binomial,
      exponentiate   = TRUE,
      outcome_label  = "DC binary",
      keep_all_terms = TRUE
    ),
    run_cluster_model(
      data           = spatial_data_proj |> filter(age_group == "younger"),
      outcome        = "ace_n",
      exposure       = "log_dist_road",
      covariates     = covariate_sets[[model_name]],
      family         = NULL,
      exponentiate   = FALSE,
      outcome_label  = "ACE",
      keep_all_terms = TRUE
    ),
    run_cluster_model(
      data           = spatial_data_proj |> filter(age_group == "older"),
      outcome        = "rudas_n",
      exposure       = "log_dist_road",
      covariates     = covariate_sets[[model_name]],
      family         = NULL,
      exponentiate   = FALSE,
      outcome_label  = "RUDAS",
      keep_all_terms = TRUE
    )
  )) |> mutate(model = model_name)
}))

write.csv(road_full, "road_distance_full_results.csv", row.names = FALSE)
```



## Interaction Term

```{r}
library(sandwich)
library(lmtest)

# Build model data
int_data <- spatial_data_proj |>
  st_drop_geometry() |>
  filter(age_group == "younger") |>
  select(ace_n, log_dist_road, ses_level, age_n, sex, cluster_id) |>
  na.omit()

# Fit interaction model
int_model <- lm(
  ace_n ~ log_dist_road * ses_level + age_n + sex,
  data = int_data
)

# Cluster-robust SEs
vcov_cl <- sandwich::vcovCL(int_model, cluster = int_data$cluster_id)
lmtest::coeftest(int_model, vcov = vcov_cl) |> round(3)
```

## figure for interaction
```{r}

pred <- ggpredict(int_model, terms = c("log_dist_road", "ses_level"))

ggplot(pred, aes(x = x, y = predicted, color = group)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group),
              alpha = 0.2, color = NA) +
  labs(
    title = "Interaction Between Socioeconomic Status and 
    Roadway Distance in Predicting ACE Scores",
    x = "Distance to road (logged)",
    y = "ACE",
    color = "SES level",
    fill = "SES level"
  ) +
  theme_minimal()

```


## Re-structuring results to be organized

-   Basic statistics
-   Unadjusted models
-   Adjusted models

Check missingness:

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  select(dc_binary, rudas_n, ace_n,
         age_n, sex, ses_level, bmi,
         education_cat, civil_stat_cat, occupation_cat,
         htn_cat, glucose_cat, chronic_dis) |>
  summarise(across(everything(), ~ sum(is.na(.)) / n() * 100)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  mutate(pct_missing = round(pct_missing, 1)) |>
  arrange(desc(pct_missing))
```

The missingness is mostly expected and minor:

-   rudas_n (69.2%) and ace_n (29.4%) — not real missingness, just
    age-group specific instruments. Expected.
-   glucose_cat (2.6%) — fine to include
-   Everything else \<1% — no concerns

```{r}
# ============================================================
# MODELING FRAMEWORK
# ============================================================

# --- Exposure list ---
exposures_raw <- c("exposure_pm25", "exposure_pm10", 
                   "exposure_temp", "exposure_humidity")

# Adjusted covariate sets
covs_basic <- c("age_n", "sex")
covs_full  <- c("age_n", "sex", "ses_level", "bmi",
                "education_cat", "civil_stat_cat", "occupation_cat",
                "htn_cat", "glucose_cat", "chronic_dis")

# ============================================================
# STAGE 1: UNADJUSTED MODELS (crude, no covariates)
# ============================================================

run_crude_model <- function(data, outcome, exposure,
                            family = NULL,
                            exponentiate = FALSE,
                            outcome_label = "outcome") {
  model_data <- data |>
    st_drop_geometry() |>
    select(all_of(c(outcome, exposure, "cluster_id"))) |>
    na.omit()

  fml <- as.formula(paste(outcome, "~", exposure))

  if (is.null(family)) {
    model <- lm(fml, data = model_data)
  } else {
    model <- glm(fml, data = model_data, family = family)
  }

  vcov_cl <- sandwich::vcovCL(model, cluster = model_data$cluster_id)
  ct <- lmtest::coeftest(model, vcov = vcov_cl)

  ct_df <- as.data.frame(unclass(ct))
  ct_df$term <- rownames(ct_df)
  rownames(ct_df) <- NULL
  colnames(ct_df)[1:4] <- c("estimate", "std.error", "statistic", "p.value")

  ct_df <- ct_df |>
    filter(term == exposure) |>
    mutate(
      conf.low  = estimate - 1.96 * std.error,
      conf.high = estimate + 1.96 * std.error
    )

  if (!is.null(family) && exponentiate) {
    ct_df <- ct_df |>
      mutate(
        estimate  = exp(estimate),
        conf.low  = exp(conf.low),
        conf.high = exp(conf.high)
      )
  }

  ct_df |> mutate(exposure = exposure, outcome = outcome_label,
                  model = "Unadjusted")
}

# Binary — unadjusted
unadj_binary <- bind_rows(lapply(exposures_raw, function(x) {
  run_crude_model(spatial_data_proj, "dc_binary", x,
                  family = binomial, exponentiate = TRUE,
                  outcome_label = "DC binary")
}))

# RUDAS — unadjusted
unadj_rudas <- bind_rows(lapply(exposures_raw, function(x) {
  run_crude_model(
    spatial_data_proj |> filter(age_group == "older"),
    "rudas_n", x, outcome_label = "RUDAS")
}))

# ACE — unadjusted
unadj_ace <- bind_rows(lapply(exposures_raw, function(x) {
  run_crude_model(
    spatial_data_proj |> filter(age_group == "younger"),
    "ace_n", x, outcome_label = "ACE")
}))

unadjusted_results <- bind_rows(unadj_binary, unadj_rudas, unadj_ace)

# ============================================================
# STAGE 2A: AGE & SEX ADJUSTED
# ============================================================

basic_binary <- bind_rows(lapply(exposures_raw, function(x) {
  run_cluster_model(spatial_data_proj, "dc_binary", x,
                    covariates = covs_basic,
                    family = binomial, exponentiate = TRUE,
                    outcome_label = "DC binary") |>
    mutate(model = "Age & sex adjusted")
}))

basic_rudas <- bind_rows(lapply(exposures_raw, function(x) {
  run_cluster_model(
    spatial_data_proj |> filter(age_group == "older"),
    "rudas_n", x, covariates = covs_basic,
    outcome_label = "RUDAS") |>
    mutate(model = "Age & sex adjusted")
}))

basic_ace <- bind_rows(lapply(exposures_raw, function(x) {
  run_cluster_model(
    spatial_data_proj |> filter(age_group == "younger"),
    "ace_n", x, covariates = covs_basic,
    outcome_label = "ACE") |>
    mutate(model = "Age & sex adjusted")
}))

basic_adjusted_results <- bind_rows(basic_binary, basic_rudas, basic_ace)

# ============================================================
# STAGE 2B: FULLY ADJUSTED
# ============================================================

full_binary <- bind_rows(lapply(exposures_raw, function(x) {
  run_cluster_model(spatial_data_proj, "dc_binary", x,
                    covariates = covs_full,
                    family = binomial, exponentiate = TRUE,
                    outcome_label = "DC binary") |>
    mutate(model = "Fully adjusted")
}))

full_rudas <- bind_rows(lapply(exposures_raw, function(x) {
  run_cluster_model(
    spatial_data_proj |> filter(age_group == "older"),
    "rudas_n", x, covariates = covs_full,
    outcome_label = "RUDAS") |>
    mutate(model = "Fully adjusted")
}))

full_ace <- bind_rows(lapply(exposures_raw, function(x) {
  run_cluster_model(
    spatial_data_proj |> filter(age_group == "younger"),
    "ace_n", x, covariates = covs_full,
    outcome_label = "ACE") |>
    mutate(model = "Fully adjusted")
}))

full_adjusted_results <- bind_rows(full_binary, full_rudas, full_ace)

# ============================================================
# COMBINE ALL INTO MASTER TABLE
# ============================================================

master_results <- bind_rows(
  unadjusted_results,
  basic_adjusted_results,
  full_adjusted_results
) |>
  mutate(
    exposure = case_when(
      exposure == "exposure_pm25"     ~ "PM2.5",
      exposure == "exposure_pm10"     ~ "PM10",
      exposure == "exposure_temp"     ~ "Temperature",
      exposure == "exposure_humidity" ~ "Humidity",
      TRUE                            ~ exposure
    ),
    model = factor(model, levels = c("Unadjusted",
                                     "Age & sex adjusted",
                                     "Fully adjusted")),
    outcome = factor(outcome, levels = c("DC binary", "ACE", "RUDAS"))
  ) |>
  select(outcome, exposure, model, estimate, conf.low, conf.high, p.value) |>
  mutate(across(where(is.numeric), ~ round(., 3))) |>
  arrange(outcome, exposure, model)

# View
master_results

# Clean kable table
master_results |>
  kable(
    caption = "Table 3. Association between environmental exposures and cognitive outcomes",
    col.names = c("Outcome", "Exposure", "Model", 
                  "Estimate", "95% CI Low", "95% CI High", "p-value")
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) |>
  collapse_rows(columns = 1:2, valign = "top")

# ============================================================
# STAGE 3: INTERACTION (road x SES, fully adjusted)
# ============================================================

int_data <- spatial_data_proj |>
  st_drop_geometry() |>
  filter(age_group == "younger") |>
  select(ace_n, log_dist_road, ses_level, age_n, sex, bmi,
         education_cat, civil_stat_cat, occupation_cat,
         htn_cat, glucose_cat, chronic_dis, cluster_id) |>
  na.omit()

int_model <- lm(
  ace_n ~ log_dist_road * ses_level + age_n + sex + bmi +
    education_cat + civil_stat_cat + occupation_cat +
    htn_cat + glucose_cat + chronic_dis,
  data = int_data
)

vcov_cl <- sandwich::vcovCL(int_model, cluster = int_data$cluster_id)
lmtest::coeftest(int_model, vcov = vcov_cl) |>
  broom::tidy() |>
  filter(str_detect(term, "log_dist_road")) |>
  mutate(
    conf.low  = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error,
    across(where(is.numeric), ~ round(., 3)),
    term = case_when(
      term == "log_dist_road"                      ~ "Log dist road (NSE E ref)",
      term == "log_dist_road:ses_levelNSE D"       ~ "× NSE D",
      term == "log_dist_road:ses_levelNSE C"       ~ "× NSE C",
      term == "log_dist_road:ses_levelNSE B"       ~ "× NSE B",
      term == "log_dist_road:ses_levelNSE A"       ~ "× NSE A (highest)",
      TRUE                                         ~ term
    )
  ) |>
  select(term, estimate, conf.low, conf.high, p.value)

# ============================================================
# STAGE 4: SENSITIVITY — road type stratification
# ============================================================

# (road_sensitivity already computed earlier — just clean and display)
road_sensitivity |>
  select(outcome, estimate, conf.low, conf.high, p.value) |>
  mutate(across(where(is.numeric), ~ round(., 3))) |>
  kable(
    caption = "Sensitivity Table. Road type-stratified analyses (fully adjusted)",
    col.names = c("Outcome", "Estimate", "95% CI Low", "95% CI High", "p-value")
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) |>
  pack_rows("Trunk roads", 1, 3) |>
  pack_rows("Primary roads", 4, 6) |>
  pack_rows("Secondary roads", 7, 9)
```

## Check master results

```{r}
master_results
```

## Add road proximity to master results

```{r}
# Unadjusted road
unadj_road_binary <- run_crude_model(spatial_data_proj, "dc_binary", "log_dist_road", family = binomial, exponentiate = TRUE, outcome_label = "DC binary") |> mutate(model = "Unadjusted")
unadj_road_rudas  <- run_crude_model(spatial_data_proj |> filter(age_group == "older"), "rudas_n", "log_dist_road", outcome_label = "RUDAS") |> mutate(model = "Unadjusted")
unadj_road_ace    <- run_crude_model(spatial_data_proj |> filter(age_group == "younger"), "ace_n", "log_dist_road", outcome_label = "ACE") |> mutate(model = "Unadjusted")

# Basic adjusted road
basic_road_binary <- run_cluster_model(spatial_data_proj, "dc_binary", "log_dist_road", covariates = covs_basic, family = binomial, exponentiate = TRUE, outcome_label = "DC binary") |> mutate(model = "Age & sex adjusted")
basic_road_rudas  <- run_cluster_model(spatial_data_proj |> filter(age_group == "older"), "rudas_n", "log_dist_road", covariates = covs_basic, outcome_label = "RUDAS") |> mutate(model = "Age & sex adjusted")
basic_road_ace    <- run_cluster_model(spatial_data_proj |> filter(age_group == "younger"), "ace_n", "log_dist_road", covariates = covs_basic, outcome_label = "ACE") |> mutate(model = "Age & sex adjusted")

# Fully adjusted road
full_road_binary <- run_cluster_model(spatial_data_proj, "dc_binary", "log_dist_road", covariates = covs_full, family = binomial, exponentiate = TRUE, outcome_label = "DC binary") |> mutate(model = "Fully adjusted")
full_road_rudas  <- run_cluster_model(spatial_data_proj |> filter(age_group == "older"), "rudas_n", "log_dist_road", covariates = covs_full, outcome_label = "RUDAS") |> mutate(model = "Fully adjusted")
full_road_ace    <- run_cluster_model(spatial_data_proj |> filter(age_group == "younger"), "ace_n", "log_dist_road", covariates = covs_full, outcome_label = "ACE") |> mutate(model = "Fully adjusted")

# Add to master
road_results <- bind_rows(
  unadj_road_binary, unadj_road_rudas, unadj_road_ace,
  basic_road_binary, basic_road_rudas, basic_road_ace,
  full_road_binary,  full_road_rudas,  full_road_ace
) |> mutate(exposure = "Road distance (log)")

master_results <- bind_rows(master_results, road_results) |>
  mutate(
    model   = factor(model, levels = c("Unadjusted", "Age & sex adjusted", "Fully adjusted")),
    outcome = factor(outcome, levels = c("DC binary", "ACE", "RUDAS"))
  ) |>
  arrange(outcome, exposure, model)
```

## Check master results again, with road proximity added

```{r}
master_results
```

## Save results as CSV

```{r}
# Save master results and interaction table as CSVs
write.csv(master_results, "/tmp/master_results.csv", row.names = FALSE)

# Interaction results
int_table <- lmtest::coeftest(int_model, vcov = vcov_cl) |>
  broom::tidy() |>
  filter(str_detect(term, "log_dist_road")) |>
  mutate(
    conf.low  = estimate - 1.96 * std.error,
    conf.high = estimate + 1.96 * std.error,
    across(where(is.numeric), ~ round(., 3)),
    term = case_when(
      term == "log_dist_road"                      ~ "Log dist road (NSE E ref)",
      term == "log_dist_road:ses_levelNSE D"       ~ "× NSE D",
      term == "log_dist_road:ses_levelNSE C"       ~ "× NSE C",
      term == "log_dist_road:ses_levelNSE B"       ~ "× NSE B",
      term == "log_dist_road:ses_levelNSE A"       ~ "× NSE A (highest)",
      TRUE ~ term
    )
  ) |>
  select(term, estimate, conf.low, conf.high, p.value)

write.csv(int_table, "/Users/gabriellebenoit/Documents/GitHub/peru_air/int_table.csv", row.names = FALSE)
write.csv(road_sensitivity, "/Users/gabriellebenoit/Documents/GitHub/peru_air/road_sensitivity.csv", row.names = FALSE)

# Forest plot data — combined
write.csv(master_results, "/Users/gabriellebenoit/Documents/GitHub/peru_air/forest_data.csv", row.names = FALSE)
```

## Figures

```{r}

# ── Excel export ────────────────────────────────────────────────────────────
write_xlsx(
  list(
    "Table 3 - All Models"       = master_results,
    "Table 5 - Interaction"      = int_table,
    "Sensitivity - Road Type"    = road_sensitivity
  ),
  path = "results_tables.xlsx"
)

# ── Forest plot helper ───────────────────────────────────────────────────────
make_forest <- function(data, title, filename, width = 14, height = 6) {
  
  data <- data |>
    mutate(
      model = factor(model, levels = c("Unadjusted",
                                       "Age & sex adjusted",
                                       "Fully adjusted")),
      null_line = ifelse(outcome == "DC binary", 1, 0)
    )

  p <- ggplot(data, aes(x = estimate, y = exposure,
                        color = model, shape = model)) +
    geom_vline(aes(xintercept = null_line),
               linetype = "dashed", color = "grey50") +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                   height = 0.2, position = position_dodge(width = 0.6)) +
    geom_point(size = 3, position = position_dodge(width = 0.6)) +
    facet_wrap(
  ~ outcome,
  scales = "free_x",
  ncol = 3,
  labeller = labeller(outcome = c(
    "DC binary" = "Cognitive impairment (binary)"
  ))
) +
    scale_color_manual(values = c(
      "Unadjusted"         = "#95A5A6",
      "Age & sex adjusted" = "#2E75B6",
      "Fully adjusted"     = "#1F4E79"
    )) +
    scale_shape_manual(values = c(
      "Unadjusted"         = 18,
      "Age & sex adjusted" = 16,
      "Fully adjusted"     = 15
    )) +
    labs(
      title    = title,
      x        = "Effect estimate (OR for binary; β for continuous)",
      y        = NULL,
      color    = "Model",
      shape    = "Model",
      caption  = "Cluster-robust SEs | * p<0.05"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      legend.position  = "bottom",
      strip.text       = element_text(face = "bold", color = "#1F4E79"),
      panel.grid.minor = element_blank()
    )

  ggsave(filename, p, width = width, height = height, dpi = 300)
}

# ── Figure 1: Air quality only ───────────────────────────────────────────────
make_forest(
  data     = master_results |> filter(exposure %in% c("PM2.5","PM10","Temperature","Humidity")),
  title    = "Figure 1. Air quality exposures and cognitive outcomes",
  filename = "figure1_airquality.png",
  width    = 14, height = 6
)

# ── Figure 2: Road distance only ─────────────────────────────────────────────
make_forest(
  data     = master_results |> filter(exposure == "Road distance (log)"),
  title    = "Figure 2. Road proximity and cognitive outcomes",
  filename = "figure2_road.png",
  width    = 10, height = 4
)

# ── Figure 3: Combined ───────────────────────────────────────────────────────
make_forest(
  data     = master_results,
  title    = "Figure 3. All environmental exposures and cognitive outcomes",
  filename = "figure3_combined.png",
  width    = 14, height = 9
)
```



## SES by sensor-assigned PM2.5
SES was not meaningfully associated with sensor-assigned PM2.5 (mean PM2.5 was essentially stable across NSE groups: ~33.3 to 33.6 μg/m3), confirming that SES does not confound the PM2.5 null result. 
```{r}
ses_pm25_summary <- spatial_data_proj %>%
  st_drop_geometry() %>%
  group_by(ses_level) %>%
  summarise(
    n = n(),
    mean_pm25 = mean(exposure_pm25, na.rm = TRUE),
    sd_pm25   = sd(exposure_pm25, na.rm = TRUE)
  ) %>%
  arrange(ses_level)

ses_pm25_summary

anova_data <- spatial_data_proj %>%
  st_drop_geometry() %>%
  select(ses_level, exposure_pm25) %>%
  filter(!is.na(ses_level), !is.na(exposure_pm25))

pm25_aov <- aov(exposure_pm25 ~ ses_level, data = anova_data)

summary(pm25_aov)

sensor_tests <- spatial_data_proj %>%
  st_drop_geometry() %>%
  filter(!is.na(ses_level), !is.na(exposure_pm25)) %>%
  group_by(nearest_sensor_name) %>%
  do(tidy(aov(exposure_pm25 ~ ses_level, data = .))) %>%
  filter(term == "ses_level") %>%
  select(nearest_sensor_name, p.value)

sensor_tests

ses_road_summary <- spatial_data_proj %>%
  st_drop_geometry() %>%
  group_by(ses_level) %>%
  summarise(
    n = n(),
    mean_distance = mean(dist_to_road_m, na.rm = TRUE),
    sd_distance   = sd(dist_to_road_m, na.rm = TRUE)
  )

ses_road_summary


model <- lm(dist_to_road_m ~ ses_level,
            data = spatial_data_proj %>% st_drop_geometry())

summary(model)

summary(model)$r.squared

anova(model)

spatial_data_proj %>%
  st_drop_geometry() %>%
  mutate(ses_num = as.numeric(factor(ses_level, ordered = TRUE))) %>%
  lm(dist_to_road_m ~ ses_num, data = .) %>%
  summary()
```






## more coherent SES table showing the monotonic aspect
```{r}

# Data from section 2.19
ses_dc <- spatial_data_proj |>
  st_drop_geometry() |>
  filter(!is.na(ses_level)) |>
  group_by(ses_level) |>
  summarise(
    n = n(),
    n_dc = sum(dc_binary, na.rm = TRUE),
    pct_dc = n_dc / n,
    # exact binomial confidence intervals
    ci_low  = binom.test(n_dc, n)$conf.int[1],
    ci_high = binom.test(n_dc, n)$conf.int[2]
  ) |>
  mutate(ses_level = factor(ses_level, 
                             levels = c("NSE E", "NSE D", "NSE C", "NSE B", "NSE A"),
                             labels = c("E (Lowest)", "D", "C", "B", "A (Highest)")))

ggplot(ses_dc, aes(x = ses_level, y = pct_dc)) +
  geom_col(fill = "steelblue", width = 0.6) +
  geom_errorbar(aes(ymin = ci_low, ymax = ci_high), width = 0.2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 0.5)) +
  labs(
    title = "Rate of Cognitive Impairment by Socioeconomic Status",
    x = "SES Level",
    y = "% with Cognitive Impairment (95% CI)",
    caption = "Error bars: 95% exact binomial confidence intervals"
  ) +
  theme_minimal()
```




## Figure to see how dementia (old and young) varies across sensor populations
```{r}
sensor_labels <- c(
  "180119" = "180119 – PuenteP Estadio",
  "159005" = "159005 – MiPeru",
  "180043" = "180043 – SanMa",
  "159599" = "159599 – Comas",
  "159435" = "159435 – Ventanilla",
  "180021" = "180021 – Carab"
)

dementia_sensor_plot <- spatial_data_proj %>%
  count(nearest_sensor_id, dc_binary) %>%
  group_by(nearest_sensor_id) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = factor(nearest_sensor_id), y = pct, fill = factor(dc_binary))) +
  geom_col(position = "dodge") +
  geom_text(aes(label = sprintf("%.1f%%", pct)),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  scale_x_discrete(labels = sensor_labels) +
  scale_fill_discrete(labels = c("0" = "No Dementia", "1" = "Dementia")) +
  labs(x = "Sensor Zone", y = "Percentage (%)",
       fill = "Dementia",
       title = "Distribution of Cognitive Decline by Sensor Zone") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

dementia_sensor_plot
```


## Re: interaction term
```{r}

slopes(int_model, 
       variables = "log_dist_road", 
       by = "ses_level")
```


## Per Monica's question:
The districts with lower employment rates is it because they are >65 years and not working? Might check this and mention if that’s the case



```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  mutate(age_group_65 = ifelse(age_n >= 65, "65+", "<65")) |>
  group_by(nearest_sensor_name, age_group_65) |>
  summarise(
    n = n(),
    pct_not_employed = mean(occupation_cat == "not_employed", na.rm = TRUE) * 100,
    pct_employed = mean(occupation_cat == "employed", na.rm = TRUE) * 100,
    pct_homemaker = mean(occupation_cat == "homemaker", na.rm = TRUE) * 100,
    .groups = "drop"
  ) |>
  arrange(nearest_sensor_name, age_group_65)
```

```{r}
spatial_data_proj |>
  st_drop_geometry() |>
  mutate(age_65plus = age_n >= 65) |>
  group_by(nearest_sensor_name) |>
  summarise(
    n = n(),
    pct_65plus = mean(age_65plus, na.rm = TRUE) * 100,
    pct_not_employed = mean(occupation_cat == "not_employed", na.rm = TRUE) * 100,
    .groups = "drop"
  ) |>
  arrange(desc(pct_65plus))
```

